---
title: "Grain-Size-Normalised Background"
subtitle: "Backgrounds after removing the grain-size effect, three ways"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The raw [Global Background](./background.qmd) page mixes grain size in: muddy sediment carries
more metal, and the coast tends to be muddier, so part of the coastal signal is grain
size rather than pressure. Normalising to a grain-size proxy strips that out. Because the
choice of proxy matters, this page does it **three ways** side by side, per element and
fraction:

- **metal / Al** (aluminium, the textbook lithogenic grain-size normaliser),
- **metal / Fe** (iron, a secondary carrier, shown for comparison only: see the note below),
- **metal / mud** (the value per unit mud fraction, `value / (fines / 100)`), computed
  only where the sample is at least `r 10`% mud, since dividing a sandy sample by a tiny
  mud fraction is unstable.

For each we compare the offshore (> 10 km) 90th percentile to the global one, exactly as
the raw page did, so the **shift toward 1 shows how much of the coastal signal was grain
size**.

::: {.callout-important}
## Iron is a reference here, not an enrichment normaliser

Norwegian fish feed contains iron, so sediment iron near a farm is partly the pressure
being measured rather than an independent measure of grain size. Normalising to Fe would
divide the signal by something that carries the signal, biasing enrichment toward zero
exactly where enrichment matters most.

Iron stays on this page as a **comparison**, because whether Al and Fe agree is itself
informative. It is not carried into any verdict: the
[Enrichment Factor](./background-ef.qmd) and
[Pristine Classification](./background-pristine.qmd) pages use aluminium only.
:::

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
graw <- rd("refined_background_compare.csv")   # page 1 (raw)
gs   <- rd("refined_gsnorm_compare.csv")        # this page (al / fe / fines)
gsp  <- rd("refined_gsnorm_percentiles.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
basis_levels <- c("raw", "al", "fe", "fines")
theme_set(theme_minimal(base_size = 12))
```

## How much of the coastal signal is grain size?

The offshore/global 90th-percentile shift, bulk, for the raw value and the three
normalised bases. A shift near 1 means the offshore and global backgrounds agree (little
left once that basis is accounted for); a shift well below 1 means a real reduction
remains.

```{r}
#| label: shift-fig
#| fig-cap: "Offshore/global P90 shift (bulk) for the raw value and the Al, Fe and mud bases. For copper, Al-normalising lifts the shift from 0.40 to 0.87: most of copper's apparent coastal enrichment was grain size. The bases disagree, most starkly for copper (Al 0.87 vs mud 0.15). Fe is shown for comparison only and is not used downstream."
#| fig-width: 8
#| fig-height: 4
raw_b <- graw |> filter(cat == "bulk") |> transmute(symbol, basis = "raw", shift = shift_p90,
                                                     n_off = n_off10)
gs_b  <- gs |> filter(cat == "bulk") |> transmute(symbol, basis, shift = shift_p90, n_off)
d <- bind_rows(raw_b, gs_b) |>
  filter(!is.na(shift), n_off >= 30, symbol %in% c("CO","CU","MN","MO","ZN")) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         basis = factor(basis, levels = basis_levels))

ggplot(d, aes(shift, symbol, colour = basis)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey60") +
  geom_point(size = 3, alpha = 0.85) +
  scale_colour_manual(values = c(raw = "#BBBBBB", al = "#4477AA",
                                 fe = "#66CCEE", fines = "#EE6677"), name = NULL) +
  labs(x = "offshore P90 / global P90 (bulk)", y = NULL)
```

The lesson is twofold. **Aluminium-normalising explains most of copper's coastal signal**
(the shift rises from 0.40 raw to 0.87), and lifts zinc's part way (0.59 to 0.74): a good
share of what looked like coastal enrichment in the raw values was simply muddier coastal
sediment. But **the three bases disagree**, sharply for copper (Al 0.87 says little
real enrichment, mud 0.15 says a lot), because aluminium, iron and the mud percentage are
not the same measure of grain size and cover different samples. So a normalised background
is only as good as its basis, and no single one is definitive.

## The normalised background values

The 90th-percentile background per element and fraction on each basis (Al and Fe are
dimensionless ratios; mud is mg/kg per unit mud fraction). Reliable groups only.

```{r}
#| label: bg-table
gsp |>
  filter(subset == "global", reliable) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  select(symbol, cat, basis, p90) |>
  pivot_wider(names_from = basis, values_from = p90) |>
  arrange(symbol, cat) |>
  transmute(Element = symbol, Fraction = cat,
            `metal/Al` = al, `metal/Fe` = fe, `metal/mud` = fines) |>
  kable()
```

::: {.callout-note}
Normalising helps but does not settle the background: it trades the grain-size assumption
for a normaliser assumption. Aluminium is the most defensible single choice, and the only
one carried into the enrichment pages, but its bulk coverage is only about a third of the
samples; the mud basis is the most direct grain-size
measure but excludes sandy samples and is unstable at low mud. Read alongside the raw
[Global Background](./background.qmd) and the [Pressure-Based Background](./background-pressure.qmd)
pages, not on its own. Iodine and selenium remain too sparse to normalise reliably.
:::
