---
title: "Global Background"
subtitle: "Estimating a natural background concentration per element and fraction"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

A **background** concentration is what a sediment would carry without local human input,
the baseline that concentrations elsewhere are judged against. There is no single way to
define it, so this page sets out two complementary estimates for each element and
fraction and compares them:

- **(A) global distribution**: the percentiles of the value over all sites, the natural
  range. The **90th percentile** (an approach used by EFSA) is one point on it.
- **(B) offshore**: the same percentiles restricted to sites **far from the coast**
  (more than `r 10` km). Local pressure, including aquaculture, is coastal, so an
  offshore-only distribution is a cleaner background; comparing it to the global one
  shows how much the coast lifts the numbers.

Everything is on the **raw** value (mg/kg), split by fraction into **bulk**,
**sieved63** (< 63 um) and **sieved20** (< 20 um), which are never pooled. Distributional
outliers are dropped, but only extreme ones: a value is flagged only if it is both more
than 4 MADs from its group median on a log scale **and** a full order of magnitude away
from that median, which catches registration errors rather than trimming the tail. It
removes 414 of 115,811 measurements, 0.36%. Grain-size-normalised backgrounds
(metal / Al, metal / Fe) are a later page.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
pct  <- rd("refined_background_percentiles.csv")
cmp  <- rd("refined_background_compare.csv")
meta <- rd("refined_background_meta.csv")
sw   <- rd("refined_background_sitewise.csv")
rp   <- rd("refined_repeat_pressure.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
cat_levels  <- c("bulk", "sieved63", "sieved20")
pct$symbol <- factor(pct$symbol, levels = elem_levels)
pct$cat    <- factor(pct$cat, levels = cat_levels)
cmp$symbol <- factor(cmp$symbol, levels = elem_levels)
cmp$cat    <- factor(cmp$cat, levels = cat_levels)
BLUE <- "#4477AA"; RED <- "#EE6677"
theme_set(theme_minimal(base_size = 12))
```

## A. The global range

The spread of each element by fraction, drawn from the global percentiles (box = the 25th
to 75th, line = median, whiskers = 10th to 90th). Reliable groups only (at least
`r meta$min_n` values); iodine, and selenium in the sieved fractions, are too sparse to
show.

```{r}
#| label: range-fig
#| fig-cap: "Global distribution of each element by fraction (log scale): box is the interquartile range, whiskers the 10th to 90th percentiles. Sieved fractions concentrate metals above bulk."
#| fig-width: 8.5
#| fig-height: 4.6
pct |>
  filter(subset == "global", reliable) |>
  ggplot(aes(cat)) +
  geom_boxplot(aes(ymin = p10, lower = p25, middle = p50, upper = p75, ymax = p90,
                   fill = cat), stat = "identity", width = 0.6) +
  scale_y_log10() +
  scale_fill_manual(values = c(bulk = "#4477AA", sieved63 = "#66CCEE",
                               sieved20 = "#228833"), guide = "none") +
  facet_wrap(~ symbol, scales = "free_y") +
  labs(x = NULL, y = "concentration, mg/kg (log)")
```

The **90th percentile** per element and fraction, the EFSA-style background value:

```{r}
#| label: p90-table
pct |>
  filter(subset == "global", reliable) |>
  transmute(Element = symbol, Fraction = cat, n, `median` = p50, `P90 (mg/kg)` = p90) |>
  arrange(Element, Fraction) |>
  kable(format.args = list(big.mark = ","))
```

## B. Offshore versus global

Restricting to offshore sites lowers the background where the coast adds metal, and
leaves it unchanged where it does not. The ratio below is the offshore 90th percentile
over the global one: **well under 1 means a strong coastal signal removed**.

```{r}
#| label: shift-fig
#| fig-cap: "Offshore (>10 km) 90th percentile relative to the global 90th percentile. Below the dashed line, offshore is cleaner (coastal enrichment removed); at or above it, the coast is not the driver."
#| fig-width: 8
#| fig-height: 3.8
cmp |>
  filter(!is.na(shift_p90), n_off10 >= meta$min_n) |>
  mutate(lab = paste(symbol, cat)) |>
  ggplot(aes(shift_p90, reorder(lab, shift_p90))) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey60") +
  geom_col(aes(fill = shift_p90 < 1)) +
  geom_text(aes(label = sprintf("%.2f", shift_p90)), hjust = -0.15, size = 3) +
  scale_fill_manual(values = c(`TRUE` = BLUE, `FALSE` = RED), guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "offshore P90 / global P90", y = NULL)
```

Read together with the numbers, the picture is:

- **Copper and zinc in bulk** drop sharply offshore (copper to about 0.4 of the global
  value, zinc to about 0.6): a clear coastal enrichment, and the offshore figure is the
  cleaner background for these.
- **Manganese and cobalt in bulk** barely move (manganese flat, cobalt slightly lower),
  so the coast is not their main driver. **Molybdenum** behaves differently by fraction:
  in **bulk** it drops steeply offshore (P90 from about 18 to 3), reflecting enrichment
  in coastal anoxic basins (restricted fjords) rather than on the open shelf, while in
  the **sieved** fractions it rises. These redox-sensitive elements are a reminder that a
  lower offshore value is not always the more natural one.
- **The sieved fractions** hardly change offshore (ratios near or above 1): sieved
  sampling is dominated by specific monitoring programmes rather than a clean
  coast-to-offshore gradient, so the offshore filter does less for them.

### How much does the 10 km cutoff matter?

Enough to matter, and most where the signal is strongest. Between the 10 km and the 50 km
cutoff the bulk P90 falls **34% for copper** (34 to 22.4) and **40% for zinc** (130 to
77.4), falls 11% for molybdenum, and *rises* 26% for manganese (932 to 1,173); only cobalt
(+2%) and selenium (-2%) are stable. Copper and zinc are the two elements this page
identifies as having the clearest coastal signal, and they are the two that keep falling as
the cutoff moves further out.

```{r}
#| label: sensitivity-table
cmp |>
  filter(cat == "bulk", n_off10 >= meta$min_n) |>
  transmute(Element = symbol, `global P90` = p90_global,
            `>10 km` = p90_off10, `>20 km` = p90_off20, `>50 km` = p90_off50) |>
  arrange(Element) |>
  kable(caption = "Bulk 90th-percentile background (mg/kg) at three offshore cutoffs.")
```

Two things follow. Ten kilometres is evidently not far enough out for copper and zinc,
whose "> 10 km" background still carries coastal input, so the > 20 km column is the more
conservative estimate for those two. And the choice propagates: the
[Enrichment Factor](./background-ef.qmd) page builds its reference from this same > 10 km
subset, so every copper and zinc enrichment factor is anchored to the least clean of the
three cutoffs and is, if anything, too forgiving.

## Does repeat sampling tilt the percentiles?

These percentiles are per **measurement**, so a site visited in ten years counts ten times
and a site visited once counts once. That only matters if repeat-sampled sites differ
systematically from the rest, which is exactly what a monitoring programme's revisits would
do. It is testable, so here is the test.

```{r}
#| label: repeat-pressure
rp |>
  transmute(`Years sampled` = revisits, `sites` = n_sites,
            `median distance to coast, km` = median_dist_to_coast_km,
            `% more than 10 km offshore` = pct_offshore_gt10km,
            `sites with a farm distance` = n_sites_with_farm,
            `median distance to farm, km` = median_dist_to_farm_km,
            `% within 5 km of a farm` = pct_within_5km_of_farm) |>
  kable(format.args = list(big.mark = ","))
```

Revisited sites **are** the pressured ones. A site sampled in four or more years sits a
median 0.75 km from the nearest farm against 1.96 km for a site sampled once, and only 17%
of them are more than 10 km offshore against 27% of the singles. The last two columns cover
Norway only, since that is where a distance to a farm exists at all; counting the rest as
"far from a farm" would be counting "not Norwegian" as "not pressured".

So the weighting is worth a sensitivity. Below, the offshore P90 computed the usual way
against the same figure with **one value per site** (that site's median) so every place
counts once:

```{r}
#| label: sitewise-table
sw |>
  filter(cat == "bulk", subset == "offshore>10km", reliable) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol, `measurements` = n_meas, `sites` = n_sites,
            `P90 per measurement` = p90_meas, `P90 per site` = p90_site,
            `ratio` = p90_ratio) |>
  arrange(Element) |>
  kable(format.args = list(big.mark = ","))
```

The effect is real but uneven. Cobalt and zinc barely move (within 5%), copper falls 15%,
and **manganese falls 32%**: its per-measurement background is meaningfully inflated by
revisited sites. Every ratio that moves, moves **down**, which is the direction the table
above predicts.

The headline figures on this page stay per measurement, because switching would discard the
real within-site variation across years and depths, and because "site" here is only latitude
and longitude rounded to three decimals. But the site-weighted column is the better estimate
of *the seabed* as opposed to *the sampling*, and for manganese the two differ enough that
the range, not either number, is the answer. Both are in
`refined_background_sitewise.csv`.

::: {.callout-note}
This is one family of background estimates, not the final answer. "Offshore" is a coarse
proxy for pristine (a site a few km out in a pressured fjord is not clean, and a coastal
site in a remote area may be), and the redox-sensitive elements show that a lower value is
not always the more natural one. The 90th percentile is a conservative upper bound of the
background, not a typical value. Later pages will add normalised backgrounds and pressure
based definitions (distance to a fish farm, enrichment factors) to triangulate. Iodine and
selenium have too few measurements for a reliable background, especially when sieved.

These percentiles also describe **the samples we hold, not the seabed**. All five sources
are monitoring programmes whose sampling is driven by regulation and interest, so this is a
convenience sample and its percentiles are not spatially representative. No reweighting
fixes that: the section above can correct for revisited sites being counted many times, but
not for the places nobody sampled.
:::
