---
title: "Geo-Accumulation Index"
subtitle: "The index that reaches the samples the enrichment factor cannot"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The [Enrichment Factor](./background-ef.qmd) page classifies a sample by dividing by
aluminium. This one classifies it by dividing by the background concentration of the same
element:

$$I_\text{geo} = \log_2\!\left(\frac{C}{1.5 \times B}\right)$$

The difference is not cosmetic. EF needs a **normaliser**, Igeo needs a **background**. A
sample can only get an EF if it also carries aluminium, on the right measurement basis, for
an element aluminium actually predicts. It gets an Igeo if it carries a value at all.

That gap is the whole reason this page exists. EFSA asked for aquaculture monitoring data by
name, and it is the one request this pipeline answered with silence: Vannmiljo's MOMC
programme carries aluminium on **5 of 13,996** subsamples, so almost nothing near a fish
farm gets an enrichment factor. EFSA also names the way out, in the same paragraph that sets
the EF preference: the geo-accumulation index and the pollution load index "can be considered
as well in absence of EF".

::: {.callout-important}
## This page issues no verdict, on purpose

Igeo is reported **alongside** the pristine verdicts, not folded into them.

Reach is the argument for promoting it, and the argument against differs by fraction. In
**bulk**, the texture confounding measured further down is strong enough (cobalt correlates
with the mud fraction at rho 0.70) that a verdict built on Igeo would be partly a verdict
about grain size. That is why the bulk verdict stays on the enrichment factor.

In the **sieved** fractions that objection does not apply, since the sieve fixed the texture
before the chemistry. The reason Igeo is not promoted there is different: the comparison it
would make, a concentration against the offshore population of the same fraction, is already
made inside the strict verdict by the offshore-P90 criterion. Promoting Igeo as well would
add a second verdict on a different threshold rather than new information.

Reporting the index without promoting it keeps the coverage visible to EFSA, who asked for
Igeo precisely where EF is absent, without quietly changing what "pristine" means here.
:::

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
igb <- rd("refined_igeo_background.csv")
igd <- rd("refined_igeo_dist.csv")
igc <- rd("refined_igeo_classes.csv")
igv <- rd("refined_igeo_coverage.csv")
igp <- rd("refined_igeo_pressure.csv")
igm <- rd("refined_igeo_pressure_matched.csv")
igbc <- rd("refined_igeo_band_cost.csv")
igpair  <- rd("refined_igeo_paired.csv")
igreach <- rd("refined_igeo_pair_reach.csv")
igws    <- rd("refined_igeo_within_site.csv")
igwr    <- rd("refined_igeo_within_reach.csv")
igf <- rd("refined_igeo_confound.csv")
igt <- rd("refined_igeo_toc_normaliser.csv")

elem_levels  <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
bin_levels   <- c("<1km", "1-5km", "5-20km", ">20km")
class_levels <- c("0 unpolluted", "1 unpolluted-moderate", "2 moderate",
                  "3 moderate-heavy", "4 heavy", "5 heavy-extreme", "6 extreme")
class_cols <- c("0 unpolluted"          = "#228833",
                "1 unpolluted-moderate" = "#66CCEE",
                "2 moderate"            = "#CCBB44",
                "3 moderate-heavy"      = "#EE7733",
                "4 heavy"               = "#EE6677",
                "5 heavy-extreme"       = "#AA3377",
                "6 extreme"             = "#882255")
theme_set(theme_minimal(base_size = 12))
```

## What it buys

The point of the index, in one table: the share of rows that get a verdict under each
instrument, overall and split by fraction and by distance to the nearest aquaculture site.

```{r}
#| label: coverage-table
igv |>
  transmute(Axis = axis, Band = band, n,
            `n EF` = n_ef, `% EF` = pct_ef,
            `n Igeo` = n_igeo, `% Igeo` = pct_igeo) |>
  kable(format.args = list(big.mark = ","))
```

Overall coverage goes from **9.8% to 97.2%**. The row that matters for EFSA's actual request
is the near-farm one: within 1 km of a fish farm, EF classifies **0.3%** of measurements and
Igeo classifies **99.9%**. The relationship inverts with distance, which is the same fact
seen from the other end: aluminium is measured in survey work far offshore and is largely
absent from the monitoring programmes that sample beneath the cages.

The sieved fractions gain almost as much. EF classifies **none** of them, because aluminium
fails the normalisability test in every sieved group; Igeo classifies 93.8 and 97.8%.

## The reference

B is the offshore (more than 10 km from the coast) **median** of the raw concentration, per
element and fraction. It is the same population the EF reference is cut from, so the two
indices disagree about method rather than about which samples count as background.

```{r}
#| label: background-table
withheld_el <- c("MO", "SE")
igb |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol, Fraction = cat, `n offshore` = n_bg,
            `B (median, mg/kg)` = signif(bg_median, 4),
            `P90 (mg/kg)` = signif(bg_p90, 4),
            Index = case_when(
              as.character(symbol) %in% withheld_el ~ "withheld (below LOQ)",
              !reliable                             ~ "none (n < 30)",
              TRUE                                  ~ "published")) |>
  arrange(Element, Fraction) |>
  kable(format.args = list(big.mark = ","))
```

Two choices in that table are worth stating, because both could reasonably have gone the
other way.

**The median, not the P90.** The 1.5 in the denominator is Muller's allowance for
lithological variability about a *central* value; it is already doing the job a higher
percentile would do. Using the P90 as B as well would apply the tolerance twice and make
almost everything read as unpolluted. The P90 is published beside it so the spread stays
visible.

**Local, not crustal.** These are not Turekian and Wedepohl values. Their deep-sea clay
figures run 9 to 24 times above this background for the clay-borne metals, and four orders of
magnitude below it for iodine, so adopting them would have decided the answer before the data
spoke. EFSA's guidance points the same way.

The right-hand column says whether that reference is turned into an index, and there are two
reasons it is not. Iodine has no usable reference at all: 5 offshore rows in bulk and 5 in
sieved63, well under the 30-row minimum. Selenium and molybdenum have perfectly good
reference counts and are withheld anyway, for a reason that has nothing to do with the
reference and is taken up at the foot of this page.

## How much of the seabed is above background?

```{r}
#| label: class-fig
#| fig-cap: "Igeo class shares, per element and fraction. Copper in bulk is the outlier: 47% of it sits above the local offshore background, and 5% reaches class 4 or higher. Every other group is dominated by class 0."
#| fig-width: 9
#| fig-height: 5.2
igc |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         igeo_class = factor(igeo_class, levels = class_levels)) |>
  ggplot(aes(pct, symbol, fill = igeo_class)) +
  geom_col(width = 0.72) +
  facet_wrap(~ cat, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = class_cols, drop = FALSE) +
  labs(x = "% of measurements", y = NULL, fill = "Igeo class")
```

```{r}
#| label: dist-table
igd |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol, Fraction = cat, n,
            `median Igeo` = igeo_p50, `P90 Igeo` = igeo_p90,
            `% class 0` = pct_unpolluted,
            `% class 2+` = pct_moderate_up,
            `% class 4+` = pct_heavy_up) |>
  arrange(Element, Fraction) |>
  kable(format.args = list(big.mark = ","))
```

Every group has a **negative median**, which is what a median-based background has to give:
half the offshore population sits below its own median, and the 1.5 pushes the whole scale
down by a further 0.58 on the log axis. Class 0 is therefore the expected place for a typical
sample, and the interesting number in that table is the right-hand column.

Copper in bulk is the standout: **23.4% at class 2 or above and 5.1% at class 4 or above**,
against roughly 1% or less for everything else. Manganese and zinc in bulk are the next two,
at 1.0% and 1.7% at class 4 or above. In the sieved fractions the tail nearly vanishes, with no
group above 0.2% at class 4.

::: {.callout-warning}
## The class names are Muller's; the meaning is not

Muller's scale is calibrated against **average shale**. B here is a local offshore median,
which for these metals sits well below shale, so "moderate" on this page means about two and
a half times the local offshore background, not two and a half times the crust.

The boundaries are kept because they are the convention a reviewer will recognise. Do not
read the labels across to a study that used a crustal reference: the same sediment would
score lower there.
:::

## Does it track the pressure gradient?

The background uses no aquaculture information, so distance to the nearest fish farm is an
**evaluation** of the index, not an input to it.

::: {.callout-important}
## This section reported the opposite result, and the bands were at fault

An earlier version of this page concluded that distance to a fish farm is a **poor axis** for
Igeo: copper and zinc appeared to peak at 5 to 20 km rather than under the cages, and the
band beyond 20 km read dirtier than the band under them. Both statements were true of the
arithmetic and false of the seabed. The bands carried two confounders, neither of which has
anything to do with fish farms, and the note that used to sit here argued the bands needed no
cleaning because they only describe where the index lands. That is right about the numbers
and wrong about the conclusion drawn from them: a band that is two fifths urban harbour
cannot be read as "far from a farm" whatever it is being used for. With both confounders
removed the far band is the cleanest band and copper shows a gradient. The raw columns are
kept beside the corrected ones so the change is visible rather than silent.
:::

### The two confounders

**Stated pressure.** Vannmiljø files why a sample was taken, and its `pressure` class is
contaminated seabed, industry and sewage. That class makes up **55.6% of the copper and 54.3%
of the zinc** beyond 20 km from a farm, against 3.5% of both under the cages. The far band was
not remote, it was industrial, so the index was being asked to compare cages against harbours.
This is the same cleaning the [Pressure-Based Background](./background-pressure.qmd) page
applies, and it belongs here for the same reason.

**Texture.** Igeo divides by a background, and this background is cut from offshore mud: `B`
sits on sediment whose median fines content is **69.4%** for copper. The band under the cages
is **38.1%** fines, the band beyond 20 km is **74.7%**. Roughly half the mud carries roughly
half the metal, so near-cage sediment reads negative on grain size alone, before any farm
enters the arithmetic. Section [What the index does not control](#what-the-index-does-not-control)
measures that correlation directly; here it is removed by comparing like with like.

### Cleaning the bands

```{r}
#| label: pressure-table
igp |>
  filter(cat == "bulk") |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         dist_bin = factor(dist_bin, levels = bin_levels)) |>
  transmute(Element = symbol, Band = dist_bin, n,
            `median Igeo` = igeo_p50, `% class 0` = pct_unpolluted,
            `% stated pressure` = pct_stated,
            `n clean` = n_clean, `median Igeo, clean` = igeo_p50_clean,
            `% class 0, clean` = pct_unpolluted_clean) |>
  arrange(Element, Band) |>
  kable(format.args = list(big.mark = ","))
```

Cleaning alone reverses the published finding. Copper beyond 20 km moves from a median Igeo of
**+0.07 to -0.37**, and its class 0 share from **47.3% to 71.7%**, which puts it below the
54.4% under the cages: the far band becomes the cleaner band rather than the dirtier one. Zinc
does the same, 58.5% to **83.8%** class 0 against 74.7% under the cages. Cobalt and manganese
barely move, because they were never the elements carrying the contaminated rows.

Only copper and zinc have a near-cage band at all once cleaning is applied. Cobalt and
manganese have 14 and 29 clean measurements under the cages, below the threshold this site
reports, so everything that follows about the innermost band is about two elements.

### Matching texture as well

Restricting every band to sediment of comparable muddiness, and reporting the achieved median
so the match can be checked rather than trusted:

```{r}
#| label: matched-fig
#| fig-cap: "Median Igeo by distance to the nearest fish farm, bulk fraction, cleaned of stated-pressure rows and restricted to sediment that is 50 to 100% fines. Copper falls monotonically from the cages outward and is the only element above the class 0 boundary under them. Cobalt and manganese have no near-cage band that survives the match."
#| fig-width: 7.5
#| fig-height: 3.4
igm |>
  filter(cat == "bulk", window == "50-100%") |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         dist_bin = factor(dist_bin, levels = bin_levels)) |>
  ggplot(aes(dist_bin, igeo_p50, group = symbol, colour = symbol)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey50") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  labs(x = "distance to nearest fish farm", y = "median Igeo",
       colour = NULL)
```

```{r}
#| label: matched-table
igm |>
  filter(cat == "bulk") |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         dist_bin = factor(dist_bin, levels = bin_levels)) |>
  transmute(`Fines window` = window, Element = symbol, Band = dist_bin, n,
            `median fines %` = fines_p50, `median Igeo` = igeo_p50,
            `% class 0` = pct_unpolluted) |>
  arrange(`Fines window`, Element, Band) |>
  kable(format.args = list(big.mark = ","))
```

**Copper now falls from the cages outward, and it is the only element that does.** Its median
Igeo runs **+0.51** under the cages, +0.35 in both middle rings and **-0.35** beyond 20 km, a
span of 0.86 of an Igeo unit, which is a factor of **1.8** in concentration. The class 0 share
goes the same way, 30.3% under the cages against 76.4% beyond 20 km. Under the cages copper is
the one element sitting above its local offshore background rather than below it.

The tighter 60 to 90% window gives the same answer on fewer rows, +0.47 against -0.39, so the
gradient is not an artefact of where the window was cut. Zinc moves in the same direction but
much less, +0.06 against -0.46 in the wide window and -0.12 against -0.49 in the tight one:
its near-cage sediment straddles the class 0 boundary rather than crossing it.

### What the controls cost

```{r}
#| label: band-cost-table
igbc |>
  filter(cat == "bulk") |>
  mutate(dist_bin = factor(dist_bin, levels = bin_levels)) |>
  transmute(Band = dist_bin, `n raw` = n_raw, `n clean` = n_clean,
            `n matched` = n_matched, `median fines %, raw` = fines_p50_raw,
            `% kept, clean` = pct_kept_clean, `% kept, matched` = pct_kept_matched) |>
  arrange(Band) |>
  kable(format.args = list(big.mark = ","))
```

The texture match is expensive and it is expensive unevenly, which is the honest weakness of
this section. Grain size is recorded on 43.9% of the rows beyond 20 km and **2.7%** of the rows
under the cages, so the matched near-cage comparison rests on 337 copper and 236 zinc
measurements out of more than 21,000. The match is also imperfect where it survives: even
inside the 50 to 100% window the near-cage rows sit at 74.8% fines against 81.6% beyond 20 km,
a residual that pushes copper's near-cage Igeo *down*, not up. The gradient is therefore a
conservative reading of a small sample, not a precise one.

`B` itself is left uncleaned, and the columns beside it on [The reference](#the-reference)
are the reason it does not need to be. The offshore pool is 8.7% stated pressure for copper
against 55.6% in the far band, and moving to the cleaned median shifts `B` by 8.3% at worst,
0.11 of an Igeo unit. Cleaning it would also cut it from a different population than the EF
reference, which is the one thing this step is built not to do.

### Holding location fixed as well

Cleaning and texture matching both act on **what** is in a band. Neither touches **where**
the bands are. That gap matters, because the farm-size work on the [Pressure-Based
Background](./background-pressure.qmd) page found location to be the confounder that bit
hardest there: size bands that looked like a dose-response turned out to be reading out
different coastlines, and pairing farms within a municipality flattened the response
completely. The same question is owed here.

The control is the same one: take one median per municipality and band, and compare two
bands **inside the same municipality**, so the regional background cancels.

**The controls do not stack, and that is the first result.** Requiring both bands in one
municipality on top of the mud window leaves a single municipality for copper, because the
texture match has already spent 97.3% of the near-cage rows. The headline near-against-far
figure above cannot be location-checked directly, and this section does not pretend
otherwise.

```{r}
#| label: igeo-pair-reach-table
igreach |>
  filter(cat == "bulk", symbol %in% c("CU", "ZN")) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         dist_bin = factor(dist_bin, levels = bin_levels)) |>
  transmute(Element = symbol, Band = dist_bin,
            `Municipalities with a usable cell` = n_muni,
            `Rows in them` = n) |>
  arrange(Element, Band) |>
  kable(format.args = list(big.mark = ","))
```

**Paired against the > 20 km band without the texture match, the result just replays the
texture confounder.** Copper comes out at -0.41, positive in under a third of the 22
municipalities that can supply the pair. That reads like a reversal and it is not news:
inside those pairs the near side is **57 percentage points sandier** (23.3% mud against
80.8%). Pairing on location does nothing about grain size. The row is published so the trap
is visible rather than stepped into twice.

**The contrast that works is the adjacent one.** A near-cage site and a site a few
kilometres away usually share a municipality, so < 1 km against 1-5 km pairs on **79**
municipalities instead of 22, and the texture gap across it is 7.6 points rather than 57.

```{r}
#| label: igeo-paired-table
igpair |>
  filter(cat == "bulk", symbol %in% c("CU", "ZN")) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Pairs = pairs, Contrast = contrast, Element = symbol,
            Municipalities = n_muni, `Igeo diff P25` = diff_p25,
            `Igeo diff median` = diff_p50, `Igeo diff P75` = diff_p75,
            `% positive` = pct_positive,
            `Median mud gap, pp` = fines_gap_p50) |>
  arrange(Pairs, Contrast, Element) |>
  kable(format.args = list(big.mark = ","))
```

Copper is **+0.25 of an Igeo unit under the cages against the ring next door, in the same
municipality, and positive in 72.2% of them**. The comparison is conservative in two
independent ways: 1-5 km is not farm-free, so the contrast understates the true gradient,
and the near side is still the sandier of the two, which suppresses the metal. Restricting
to the pairs where the near side is *not* the sandier one is the `texture-balanced` block,
and it does what removing a suppressor should do: copper rises to **+0.46 and 82.1%
positive** on 28 municipalities.

Zinc stays weak. It is -0.01 at 49.4% across all 77 pairs, which is a coin flip, and only
reaches +0.08 at 60.7% when the texture-balanced pairs are taken. That is the fourth design
on this site to put zinc near nothing.

### The control that needs no matching: the same site, before its farm

Every control above matches one population against another and pays for it. Texture
matching spends 97.3% of the near-cage rows. Municipality pairing cannot be stacked on top
of it. The adjacent contrast that survives both is blunted, because 1-5 km is not
farm-free. All three are approximations of the one thing actually wanted: a farm-free
reading of **the same seabed**.

The database can supply that directly. `fish_farm_aqua_id` carries the farm's licence
dates, so a site sampled before its farm opened and again while it was operating is its own
control. That holds fixed what no matching reaches: the exact location, the water depth,
the regional background, the sediment's provenance, and the sampling programme. It needs no
grain-size measurement at all.

```{r}
#| label: igeo-within-reach-table
igwr |>
  filter(cat == "bulk", symbol %in% c("CU", "ZN")) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol, `Near-cage sites` = n_sites_any,
            `With a pre-farm sample` = n_sites_pre,
            `Pre-farm rows` = n_rows_pre, `Operating rows` = n_rows_op,
            `Median years before the farm opened` = lead_years_p50) |>
  arrange(Element) |>
  kable(format.args = list(big.mark = ","))
```

The pre-farm rows are almost all MOM baseline surveys, taken a median of **one year** before
the farm opens. This is a designed baseline, not an accident of the archive.

**It needs a placebo**, because a before-and-after is a time series and metals in Norwegian
sediment are not stationary. The same contrast is run in two rings further out, split on the
**same farm's** start year, so any secular trend appears there with no farm depositing under
the sampler. And the years-between-samples gap has to be matched: raw, the near-cage pairs
sit 4 years apart and the 5-20 km pairs 10, so an unmatched comparison would charge the
placebo with six extra years of drift.

```{r}
#| label: igeo-within-table
igws |>
  filter(cat == "bulk", symbol %in% c("CU", "ZN")) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Basis = basis, Element = symbol, Zone = zone, Sites = n_sites,
            `Median gap, yr` = gap_p50, `Igeo change P25` = delta_p25,
            `Igeo change median` = delta_p50, `Igeo change P75` = delta_p75,
            `% rising` = pct_rising, `Igeo per year` = rate_p50) |>
  arrange(Basis, Element, Zone) |>
  kable(format.args = list(big.mark = ","))
```

```{r}
#| label: igeo-within-fig
#| fig-cap: "Change in median Igeo at the same site between a pre-farm sample and one taken while the farm operated, by distance from the farm, gap matched to 2-7 years. Only the first ring has a farm above it; the others are placebos split on the same farm's start year. Sediment away from farms is getting cleaner, sediment under the cages is getting dirtier, and the turn happens at the cages."
#| fig-width: 7.5
#| fig-height: 3.8
igws |>
  filter(cat == "bulk", symbol %in% c("CU", "ZN"),
         basis == unique(igws$basis)[1]) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  ggplot(aes(zone, delta_p50, group = symbol, colour = symbol)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey55") +
  geom_line(linewidth = 0.7) +
  geom_point(aes(size = n_sites)) +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  scale_size_continuous(range = c(2, 5.5), name = "sites") +
  labs(x = "distance from the farm (only the first ring has one above it)",
       y = "change in median Igeo, pre-farm to operating")
```

Under the cages, copper rises **+0.22 of an Igeo unit** and is rising at **69.4%** of the 72
sites. In the 5-20 km placebo it falls **-1.05** and is rising at only 22.2%. The secular
trend in this sediment is downward, which means every cross-sectional number on this page
is measured against a background that is itself improving, and therefore understates the
farm. Read as a rate over all pairs, the gradient is monotone: **+0.050** Igeo units per
year under the cages, **-0.017** at 1-5 km, **-0.044** at 5-20 km.

::: {.callout-important title="This is the one design where zinc is not null"}
Zinc rises **+0.23** under the cages, at **70.8%** of 65 sites, against **-0.53** in the
placebo. Four cross-sectional designs on this site have put zinc at or near zero, and this
one does not.

The two are not in conflict, and the reason is worth stating. Every other design compares
**different places**, and zinc's variation between places is dominated by sediment
character, which is large enough to swamp a farm signal. This design compares **one place
with itself**, so between-place variation is removed entirely rather than matched away.
Zinc being invisible across places and visible within them is what a modest, real effect
sitting under a large nuisance variance looks like.

It is still 65 sites, and it is the minority reading. The pristine classification and the
exports are unchanged, and zinc keeps its withheld-verdict status until a design with more
baseline sites agrees.
:::

**What this cannot separate**, and it changes the reading: a farm enriches sediment with
organic matter, and organically enriched sediment is finer. Some of the rise measured here
may be the seabed becoming muddier rather than the metal load increasing. That is mediation
rather than confounding, since the farm caused both, but "the farm raised the metal" here
includes "the farm changed the sediment". No site in the design has a grain-size measurement
in both periods, so it cannot be checked on this data.

### Reading it beside the other pages

Copper is the only near-cage signal that survives every control on this site, and it now
survives two more. The [Pressure-Based Background](./background-pressure.qmd) page puts
near-cage copper at 2.6 times the cleaned far background, and
[Pressure Controls](./pressure-controls.qmd) brings that to 1.6 once the comparison is matched
within a municipality; the texture-matched Igeo here says 1.8, and the municipality-paired
Igeo above puts the near band 0.25 of a unit over the ring next door in 72% of municipalities.
Four different controls, four different row populations, one answer, and none of them needs
aluminium.

The three Igeo controls are worth keeping apart, because they fail in different directions
and none can cover for the others. Texture matching holds grain size and lets location vary,
on 337 copper measurements. Municipality pairing holds location and lets grain size vary, on
79 municipalities. The first two cannot be applied together, since their intersection is one
municipality. The before-and-after design holds *everything* fixed by using one site as its
own control, needs no grain size at all, and pays for it in a different currency: 80 sites,
and an inability to separate metal input from the farm making the seabed muddier.

Copper is above background on all three, on populations that barely overlap. That is the
strongest statement this page can make, and it is stronger than any single control in it.

Zinc is where the pages disagree, and the disagreement is informative. Texture-matched Igeo
shows zinc rising toward the cages by a factor of 1.4, while the municipality-matched
comparison on Pressure Controls puts it at 1.04 on what is effectively a coin flip. The two
control different things: one holds grain size and lets location vary, the other holds location
and lets grain size vary. That zinc needs the texture control to show anything, and loses it to
the location control, is the argument for reading its near-cage signal as sediment character
rather than as farm input. The municipality pairing says the same thing a third time: 49.4%
positive across 77 municipalities.

The before-and-after design is the one dissent, and it is a considered one rather than
noise: zinc rises at 70.8% of 65 sites where the placebo falls. The reconciliation is that
every design except that one compares different places, and zinc's between-place variation
is big enough to hide a modest effect. Zinc is therefore best described as **unresolved**
rather than absent, which is a change from what this page said before, and it is the
strongest argument on the site for collecting more pre-farm baselines.

## What the index does not control

Igeo divides by a background, not by a grain-size carrier, so nothing in it separates
"muddy" from "polluted". Measured against the mud fraction, per group, alongside EF on the
rows that carry one:

```{r}
#| label: confound-table
igf |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol, Fraction = cat,
            `n Igeo` = n, `rho Igeo vs fines` = rho_igeo_fines,
            `n EF` = n_ef, `rho EF vs fines` = rho_ef_fines) |>
  arrange(Element, Fraction) |>
  kable(format.args = list(big.mark = ","))
```

In bulk, Igeo tracks the mud fraction for every metal, worst for **cobalt at rho 0.70**. In
the sieved fractions it barely does, which is the normalisability finding from the other
side: a sieved sample is already grain-size controlled, so there is little texture left for
the index to pick up.

**EF is not uniformly the better instrument.** On the same rows it beats Igeo for cobalt
(0.24 against 0.70) and it **loses** for copper (0.55 against 0.36), tying for zinc. Dividing
by aluminium demonstrably removes the texture signal for cobalt and demonstrably does not for
copper. The comparison is between different row populations, since EF only exists where
aluminium does, so read it as a scale rather than a like-for-like test.

The case for Igeo is therefore strongest exactly where EF cannot run at all, in the sieved
fractions and the near-farm data, and a bulk Igeo should be read with the texture caveat
attached.

## The alternative that does not work

If the problem is a missing normaliser, the obvious question is whether another one is
present. Organic carbon is: 33.1% of bulk target rows carry TOC and no aluminium, 28,397
measurements, and metals do sorb to organic matter. Put through the same two-part test
aluminium had to pass (R² at least 0.30 **and** rho at least 0.50):

```{r}
#| label: toc-table
igt |>
  filter(!is.na(r2) | !is.na(rho)) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol, Fraction = cat,
            `n offshore` = n_off, `R2` = r2, rho,
            Normalisable = ifelse(normalisable_toc, "yes", "no")) |>
  arrange(Element, Fraction) |>
  kable(format.args = list(big.mark = ","))
```

It clears both limits in exactly **one group of twenty**, and that group cannot be used:
selenium sieved63, on 34 offshore rows, for an element whose verdicts are already withheld
because below-LOQ censoring deleted 68.6% of it. A strong fit to the top third of a truncated
distribution is not evidence about background.

Everywhere it could matter it fails, and it fails in the way that is hardest to argue with:
**the two measures disagree**. Copper is the best case in bulk, rho 0.60 against R² 0.21, and
copper sieved63 inverts it, R² 0.37 against rho 0.32. For aluminium the two measures agreed
on every group, which is what made the normalisability test a clean partition; for TOC they
pull apart, so there is no cut that turns it into a normaliser. Those 28,397 rows stay
unnormalised, and Igeo remains the only route to a verdict for them.

## What Igeo does not fix

**Selenium and molybdenum stay withheld.** They carry no index, no class and no verdict, so
they are absent from every table on this page that reports one.

Their offshore reference is still published in the reference table, and they still appear in
the organic-carbon test, on the same principle the enrichment pages follow: what is withheld
is the verdict, not the measurement or the background behind it.

Their exclusion was never about normalisers. The clean stage removes below-LOQ rows, which
deletes **68.6% of selenium and 52.2% of molybdenum**, so what survives is an upper tail
wearing the name "background". Both ends of the ratio are truncated: the offshore median in
the denominator is inflated, and the numerator holds only the measurements that cleared the
limit. A different index over the same truncated distribution inherits the same problem.

Normalisability, by contrast, does **not** apply here. That test asks whether aluminium
predicts the metal, and nothing on this page divides by aluminium. Igeo escapes the
normaliser constraint and not the censoring one.
