---
title: "Pressure-Based Background"
subtitle: "Background from distance to a fish farm, and what the size of that farm adds"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Instead of a spatial proxy for pristine, this page uses a direct **pressure gradient**:
distance to the nearest **fish farm**. Sites far from any farm are the low-pressure
**background**; sites near the cages show the **enrichment**. This is the aquaculture
question asked directly, and it is **Norway only** (the aquaculture reference is
Norwegian).

For each element and fraction, the value is summarised in distance bands (< 1 km, 1-5 km,
5-20 km, > 20 km). The **> 20 km** band is read as the background; the ratio of the
nearest band to it (the 90th percentiles) is the near-cage enrichment.

Because it is built from the farm positions, this pressure gradient is held out of the
background and pristine estimates themselves. Its role is to **evaluate** them: the
[Enrichment Factor](./background-ef.qmd) and [Pristine
Classification](./background-pristine.qmd) pages are estimated without aquaculture
information, and this page is the yardstick that the enrichment they report matches the
real pressure.

::: {.callout-important
title="This page was rebuilt in August 2026 and its numbers moved"}
It used to measure distance to the nearest **aquaculture site of any kind**, and a
land-based smolt plant and a mussel raft are both aquaculture. Two corrections follow,
and they push in opposite directions: the near band was too generous (it counted samples
next to facilities that put nothing on the seabed) and the far band was not a background
at all (over half of it was contaminated-seabed monitoring). The old copper figure of
about 3.2x and molybdenum figure of about 13x came from that pairing. Both sections
below show the change rather than quietly replacing it.
:::

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
pp   <- rd("refined_pressure_percentiles.csv")
pc   <- rd("refined_pressure_compare.csv")
ax   <- rd("refined_pressure_axis.csv")
axd  <- rd("refined_pressure_axis_dropped.csv")
sz   <- rd("refined_pressure_size.csv")
szc  <- rd("refined_pressure_size_covariates.csv")
st   <- rd("refined_pressure_stated.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
bin_levels  <- c("<1km", "1-5km", "5-20km", ">20km")
band_levels <- c("small", "medium", "large")
pp$symbol   <- factor(pp$symbol, levels = elem_levels)
pp$dist_bin <- factor(pp$dist_bin, levels = bin_levels)
theme_set(theme_minimal(base_size = 12))
```

## The axis: a fish farm, not aquaculture

A fish farm is the pressure this work is about, because the mechanisms are feed
(selenium, zinc, and the organic enrichment that drives the redox chemistry) and the
copper in net antifouling. A mussel raft has none of them, and a land-based smolt plant
puts its effluent through a pipe rather than onto the seabed under the licence
coordinate. Counting either as "near a farm" fills the near band with samples that carry
none of the pressure the band exists to isolate.

The refined `site` table carries both distances, so the change can be shown rather than
asserted. What the old axis added to the near band, by the kind of facility the sample
was actually next to:

```{r}
#| label: dropped-table
axd |>
  transmute(Placement = placement, `Licensed in` = capacity_unit,
            Measurements = n, `% of dropped` = pct) |>
  kable(format.args = list(big.mark = ","))
```

```{r}
#| label: dropped-note
#| echo: false
#| results: asis
land <- round(sum(axd$pct[axd$placement == "land"]), 1)
sea_other <- round(sum(axd$pct[axd$placement != "land"]), 1)
cat(sprintf(
  "So **%s%%** of what the old near band gained sat next to a *land-based* facility, and %s%% next to something at sea that is not a fish farm (mostly blue mussel, licensed by area in decare). None of them is a sea cage.\n",
  land, sea_other))
```

The effect on the two elements with enough data to see it:

```{r}
#| label: axis-table
ax |>
  filter(cat == "bulk", !is.na(enrich_near_farm), n_near_farm >= 100) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol,
            `n near (farm)` = n_near_farm,
            `n added by old axis` = n_near_gain,
            `old ratio` = enrich_near_aqua,
            `new ratio, raw far band` = enrich_near_farm) |>
  arrange(Element) |>
  kable(format.args = list(big.mark = ","))
```

The new raw ratios are *below* 1, which is the second problem rather than a finding about
farms.

## The background had to be cleaned too

The far band is only a background if what sits in it is unpressured, and it is not.
Vannmiljø files why each sample was taken, and the > 20 km pool turns out to be more than
half **stated pressure**: the contaminated-seabed, industry and sewage programmes, which
is urban fjord sediment carrying the very metals being measured.

```{r}
#| label: farband-table
pc |>
  filter(cat == "bulk", !is.na(enrich_near)) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol,
            `n far` = `n_>20km`, `P90 far, raw` = `>20km`,
            `% stated pressure` = pct_far_stated,
            `n far, cleaned` = n_far_clean,
            `P90 far, cleaned` = p90_far_clean) |>
  arrange(Element) |>
  kable(format.args = list(big.mark = ","))
```

Removing the stated-pressure rows takes the copper background from 102 to 25.7 mg/kg and
the zinc background from 273 to 126.5. Only rows a provider has **stated** are pressure
monitoring are removed; the four sources that record no programme keep every row, so this
removes stated pressure and never inferred pressure.

**Every ratio below is taken against the cleaned far band.**

## Concentration versus distance to a fish farm

The 90th percentile by distance band, bulk (reliable groups only). A value that falls as
distance grows is a pressure signal; the far band is the background.

```{r}
#| label: gradient-fig
#| fig-cap: "Bulk 90th-percentile concentration by distance to the nearest fish farm (log scale). Only cobalt falls monotonically with distance. Copper, zinc and molybdenum peak in the 1-20 km bands rather than under the cages, and manganese is diagenetic rather than a pressure signal."
#| fig-width: 8.5
#| fig-height: 4.4
pp |>
  filter(cat == "bulk", reliable, symbol %in% c("CO","CU","MN","MO","ZN")) |>
  ggplot(aes(dist_bin, p90, group = symbol, colour = symbol)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  labs(x = "distance to nearest fish farm", y = "P90 concentration, mg/kg (log)")
```

## Near-cage enrichment

The nearest-band (< 1 km) 90th percentile over the cleaned far background (> 20 km).
Groups below 30 measurements in the near band are marked rather than dropped, because
which elements *lack* near-cage data is itself the finding:

```{r}
#| label: enrich-table
pc |>
  filter(cat == "bulk", !is.na(enrich_near_clean)) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol,
            `n < 1 km` = `n_<1km`,
            `P90 < 1 km` = `<1km`,
            `P90 far (cleaned)` = p90_far_clean,
            `enrichment` = enrich_near_clean,
            Reading = if_else(`n_<1km` >= 30, "", "too few to read")) |>
  arrange(desc(enrichment)) |>
  kable()
```

What survives:

- **Copper, about 2.6x**, on 11,475 measurements. Antifouling treatment of nets and
  cages. This is the one strong, well-supported near-cage signal on the page.
- **Cobalt, about 1.5x**, but on 33 measurements, which is barely over the threshold.
- **Zinc, about 1.1x**, on 10,277 measurements. Feed and galvanised structures, and on
  this axis the effect is small.
- **Manganese, 0.35**, near well below far. It is diagenetic, not a pressure signal, the
  same behaviour it showed offshore.
- **Molybdenum and selenium have almost no near-cage data at all**: 21 and 1 bulk
  measurements within 1 km of a sea cage. The old page reported them as the strongest
  signals on the page (about 13x and about 5x). Those figures came from the near band
  when it still included land-based facilities. They are withdrawn, not restated. Both
  elements are also under the [LOQ censoring
  decision](./background.qmd), so they carry no verdict anywhere in this project.

The gradient is not a dose-response curve. Copper, zinc and molybdenum all peak in the
1-20 km bands rather than under the cages, which is what the next section is for.

## Does the size of the farm change the signal

Norwegian finfish licences are issued in MTB (maximum permitted standing biomass), and
the standard concession is 780 t, so the refined `site` table carries the nearest farm's
licensed capacity and a size band: **small** (up to 2 concessions), **medium** (up to 4),
**large** (above 4). A 780 t farm and a 19,000 t farm at the same distance are not the
same pressure.

This is a better test than near against far, because both sides sit under a cage in a
fjord: the setting is roughly held constant and only the licensed capacity moves.

```{r}
#| label: size-table
sz |>
  filter(cat == "bulk", reliable) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         band = factor(band, levels = band_levels)) |>
  transmute(Element = symbol, `Size band` = band,
            `Median MTB, t` = mtb_median, n = n,
            `P90 < 1 km` = p90, `P90 far (cleaned)` = p90_far_clean,
            enrichment = enrich_clean) |>
  arrange(Element, `Size band`) |>
  kable(format.args = list(big.mark = ","))
```

```{r}
#| label: size-fig
#| fig-cap: "Near-cage enrichment by the licensed size of the nearest farm, bulk, cleaned far background. Both elements peak at medium and fall at large, which is the opposite of a dose-response."
#| fig-width: 7
#| fig-height: 3.6
sz |>
  filter(cat == "bulk", reliable) |>
  mutate(band = factor(band, levels = band_levels)) |>
  ggplot(aes(band, enrich_clean, group = symbol, colour = symbol)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey55") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.6) +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  labs(x = "licensed size of the nearest fish farm",
       y = "near/far enrichment (P90)")
```

Only copper and zinc have enough near-cage data for the split; cobalt, manganese and
molybdenum fall under 30 measurements in every band. Both behave the same way: small and
medium are close, and **large is the lowest of the three**. Copper goes 2.91, 3.15, 2.13
and zinc 1.11, 1.26, 0.95.

That is the opposite of a dose-response, and the reason is almost certainly siting rather
than chemistry:

```{r}
#| label: size-cov-table
szc |>
  mutate(band = factor(band, levels = band_levels)) |>
  transmute(`Size band` = band, Sites = n_sites, `Median MTB, t` = mtb_median,
            `Median depth, m` = median_depth_m,
            `Median distance to coast, km` = median_dist_coast_km,
            `Median mud, %` = median_fines_pct,
            `% with a mud measurement` = pct_with_fines) |>
  arrange(`Size band`) |>
  kable(format.args = list(big.mark = ","))
```

Large and medium farms sit in roughly **twice the water depth** of small ones (107 m and
121 m against 66 m) and further from shore. Norway licenses the big farms into deep,
well-flushed water precisely because it disperses the load, so a bigger farm depositing
more organic matter over a larger and better-flushed footprint can leave *less* under the
licence coordinate. A flat or inverted size response is therefore not evidence that the
pressure is absent, and this table is what the one above has to be read against.

The honest reading: **size does not give a clean dose-response on this data, and the
confounder is large enough to explain why.** Separating the two needs depth-matched
comparison within a size band, which this page does not do.

## The provider's own label agrees with the geometry

Vannmiljø records why a sample was taken, independently of any coordinate arithmetic. If
the distance axis is measuring what it claims to, the rows the provider filed as
aquaculture monitoring should sit near a farm, and the ones filed as reference conditions
should not:

```{r}
#| label: stated-table
st |>
  filter(pressure_class %in% c("aquaculture", "reference", "pressure")) |>
  mutate(dist_bin = factor(dist_bin, levels = bin_levels)) |>
  transmute(`Stated purpose` = pressure_class, `Distance band` = dist_bin,
            Measurements = n, `% of that purpose` = pct_of_class) |>
  arrange(`Stated purpose`, `Distance band`) |>
  kable(format.args = list(big.mark = ","))
```

**80% of the aquaculture-monitoring measurements fall within 1 km of a fish farm and
95.7% within 5 km, with 0.3% beyond 20 km. Every one of the 64 reference-condition
measurements is beyond 20 km.** Two independent things, a Norwegian programme code and a
great-circle distance to a licence coordinate, agree about which samples are near a farm.
That is the strongest evidence on this page that the axis is sound.

The same table shows the far-band problem from the other side: 40.9% of the stated
**pressure** measurements also sit beyond 20 km, which is what the cleaned background
removes.

## What this page still does not fix

Three things, in descending order of how much they matter.

1. **Time.** Every sample is keyed to its nearest farm whether or not that farm existed
   when the sediment was sampled. The aquaculture reference carries licence start and end
   years, so this is measurable, and [Pressure Controls](./pressure-controls.qmd) measures
   it on the same axis and the same cleaned background as this page. It barely bites:
   87.8% of near-cage copper and 86.9% of near-cage zinc are contemporary with an
   operating farm, and restricting to those moves copper from 2.62 to 2.66.
2. **Geography.** The near band is fjord and the far band is largely open shelf (a median
   72 km from the coast against 400 m), so the comparison is two settings as much as two
   pressures. This is the control that bites: matching within municipality takes copper
   from 2.62 to 1.60 and zinc from 1.09 to 1.04. Copper keeps a clear excess and zinc does
   not. See [Pressure Controls](./pressure-controls.qmd).
3. **Texture.** The near band is a median 32-41% mud depending on size band, and only
   12-16% of near-cage subsamples have a mud measurement at all. The
   [Grain-Size-Normalised](./background-gsnorm.qmd) page is the texture cross-check.

::: {.callout-note}
This is Norway only, and "distance to a farm" remains a coarse pressure measure (a distant
but industrial site is not pristine, and a near site in a well-flushed location may be
clean). The > 20 km cleaned band is a practical low-pressure background, not a guaranteed
pristine one. Read with the raw [Global Background](./background.qmd), the
[Grain-Size-Normalised](./background-gsnorm.qmd) page and
[Geo-Accumulation Index](./background-igeo.qmd) to triangulate.
:::
