---
title: "Pristine Classification"
subtitle: "Combining the background estimates into a per-sample verdict, and where it cannot be made"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

This page draws the background estimates together into a per-measurement **pristine**
(at or below background) verdict, per element and fraction (a sample can be pristine for
one element and not another). It draws on three "at/below background" criteria from the earlier
pages: **EF < 1** (the [Enrichment Factor](./background-ef.qmd)), **below the
[distribution-mixture threshold](./background-mixture.qmd)**, and **below the offshore P90**
([Global Background](./background.qmd)). Which of the three apply depends on how the
fraction's grain size is controlled, and that is the subject of the next two sections.

Every one of these criteria is estimated with **no aquaculture information**, so the
pristine verdict is independent of the pressure it is meant to reveal. The aquaculture
sites enter only afterwards, as an **evaluation**: the closing section checks whether the
pristine fraction falls as sediments approach the cages. That check turns out to be a sanity
test rather than a validation, for reasons set out there.

Only **grain-size-controlled** verdicts are trusted. In bulk, a raw-concentration version was
tried and rejected: sandy near-cage sediment slips under a concentration threshold regardless
of pressure, and (see below) aluminium is missing exactly where pressure is highest, so a
concentration flag validated **backwards**. Where a bulk sample has no aluminium, and hence
no EF, it is therefore left **unclassified** rather than guessed.

That rejection is about **uncontrolled** concentrations, and it is worth being precise about
why it does not extend to the sieved fractions. The objection to a concentration threshold is
that grain size confounds it. In a sieved sample grain size has already been fixed, by the
sieve, so the same threshold is no longer measuring texture. The two concentration criteria
are untrustworthy in bulk and sound on a sieved fraction for one and the same reason: what
makes them unsafe is missing there.

What counts as grain-size control, though, depends on the fraction, and this page used to
get that wrong. In **bulk** the control is statistical, and it has to be earned: dividing by
a variable uncorrelated with the metal is not a control however it is dressed up. Measured by
regression, aluminium predicts cobalt, copper and zinc in bulk and nothing else, so manganese
in bulk is unclassified for the same reason selenium and molybdenum are
([Regression Normalisation](./background-regression.qmd)).

In a **sieved** sample the control is physical and was applied before the chemistry started.
Cut to below 63 or 20 um, the texture variation is gone, which is precisely why aluminium
has nothing left to track there. Reading that as a failed normaliser and withholding the
verdict, as this page did until August 2026, treated "no correction needed" as "the
correction failed" and penalised the fraction with the better control. It is the better
control: taking each fraction's offshore background and asking how far it moves between
seas, sieved63 copper spans 2.5x and zinc 1.7x untouched, against 4.4x and 4.9x for bulk
**after** aluminium normalisation. So the sieved fractions are classified on the sieve, and
aluminium is not one of their criteria. See [Method Revisions](./method-revisions.qmd).

Two flags are reported, and which criteria they contain depends on the fraction:

- **EF rule** (permissive, EFSA-preferred): pristine where EF < 1. Bulk only, since no
  enrichment factor is defined where aluminium is not the normaliser.
- **Strict rule** (conservative): every criterion that applies to the fraction agrees. In
  bulk that is EF < 1 **plus two raw-concentration screens**, the mixture threshold and the
  offshore P90. In the sieved fractions it is those two screens, resting on the sieve for
  their grain-size control. The sieved fractions carry the strict rule only, which is the
  right way round while their background cannot be checked in the north (see below).

The three criteria are not peers, and the strict rule is not three independent votes. Only
EF < 1 is grain-size controlled; the other two are screens on the raw concentration, the
same kind of test this page has just rejected on its own. ANDing them onto EF < 1 cannot
make a sample falsely pristine, only falsely non-pristine, which is the safe direction for a
conservative rule. It does mean a **muddy but clean** sample can fail the strict rule on
grain size alone. Cobalt used to be the clearest case, 56% pristine on the EF rule against
20% strict, which reflected its unreliable mixture threshold rather than any enrichment.
That threshold is now recognised as unusable and dropped from the rule, so the two rules
have converged for cobalt; the mechanism still applies wherever a usable threshold sits low
against muddy sediment.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
summ <- rd("refined_pristine_summary.csv")
cov  <- rd("refined_pristine_coverage.csv")
covs <- rd("refined_pristine_coverage_source.csv")
val  <- rd("refined_pristine_validation.csv")
ref  <- rd("refined_pristine_reference.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
BLUE <- "#4477AA"; RED <- "#EE6677"
theme_set(theme_minimal(base_size = 12))
```

## How much is pristine, and how much is classifiable

Among the samples that can be classified (those with aluminium), the share flagged pristine
under each rule, bulk. Both rules rest on EF < 1, whose reference is now computed within a
single aluminium measurement basis (see the [Enrichment Factor](./background-ef.qmd) page).
That is why the **classifiable** column is low: a sample off its fraction's basis is
withdrawn rather than misjudged. Only **cobalt, copper and zinc** appear in bulk. Selenium
and molybdenum are withheld because most of their measurements fell below the limit of
quantification and were removed upstream; manganese in bulk is withheld because aluminium
does not predict it ([Method Revisions](./method-revisions.qmd)). For copper and zinc the
classifiable column is the further catch: only about one sample in nine carries usable
aluminium at all. The sieved fractions, which need none of it, are in their own section
below.

```{r}
#| label: summary-table
summ |>
  filter(cat == "bulk", reliable == TRUE) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol, `samples` = n, `% classifiable` = pct_classifiable,
            `pristine (EF rule)` = pct_ef, `pristine (strict)` = pct_strict) |>
  arrange(Element) |>
  kable(format.args = list(big.mark = ","))
```

```{r}
#| label: pristine-fig
#| fig-cap: "Share of classifiable bulk samples flagged pristine, EF rule versus strict, for the three groups that carry a verdict. The two rules now differ by at most one point: since the mixture criterion is dropped where the fit does not separate, the raw screens rarely bite."
#| fig-width: 7.5
#| fig-height: 3.4
summ |>
  filter(cat == "bulk", reliable == TRUE) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  select(symbol, EF = pct_ef, strict = pct_strict) |>
  pivot_longer(-symbol, names_to = "rule", values_to = "pct") |>
  ggplot(aes(pct, symbol, fill = rule)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(EF = BLUE, strict = RED), name = NULL) +
  labs(x = "% of classifiable samples pristine", y = NULL)
```

## The sieved fractions, judged on the sieve

These carry the strict rule only. Every criterion in it is a concentration screen, and both
screens stand on the sieve for their grain-size control rather than on aluminium.

```{r}
#| label: sieved-table
summ |>
  filter(cat != "bulk", reliable == TRUE) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol, Fraction = cat, `samples` = n,
            `% classifiable` = pct_classifiable, `pristine (strict)` = pct_strict) |>
  arrange(Element, Fraction) |>
  kable(format.args = list(big.mark = ","))
```

Classifiable is at or near 100% throughout, which is the whole point: no aluminium is
required, so nothing is withdrawn for lacking it. Contrast the bulk column above, where
copper and zinc lose eight samples in nine.

Manganese appears here and not in bulk. That is consistent rather than odd: its bulk verdict
is withheld because aluminium never carried it (manganese is redox-mobile rather than
clay-borne, so it is not a normalisation that failed but a normaliser that was never
appropriate), and in a sieved sample the sieve supplies the grain-size control that
aluminium was being asked for. Its redox mobility is still a caveat on reading the number,
but it is not a grain-size question. Selenium and molybdenum stay withheld in every
fraction, on censoring rather than on normalisation.

### The standing caveat: it is not the same sea

Sieving is a national convention, not a property of a sample. Norway reports bulk and
Germany, Belgium and the Netherlands sieve, so "offshore, more than 10 km out" resolves to a
different ocean in each fraction:

```{r}
#| label: reference-table
ref |>
  transmute(Fraction = cat, `grain-size control` = gs_control,
            `reference samples` = n, `median latitude` = paste0(lat_p50, "\u00b0N"),
            `median depth (m)` = depth_p50, `mainly` = top_seas,
            `north of 60N` = n_north_60,
            `within 5 km of a farm` = n_farm_lt5km) |>
  kable(format.args = list(big.mark = ","))
```

::: {.callout-warning}
## Read the sieved verdicts against their own sea, not against Norway

The bulk reference is the deep Norwegian and Barents Sea. The sieved reference is the
southern North Sea and the Baltic, on a shallow shelf off an industrialised continental
coast. The `north of 60N` column is why this cannot be repaired by re-cutting the
reference: there is almost no sieved data up there to build a northern one from, so the
sieved background **cannot be checked in Norwegian waters**.

The last column is the matching limit on the evaluation below. The sieved fractions hold no
samples within 5 km of a Norwegian fish farm at all, so the distance check that follows
cannot be run on them and stays on bulk.
:::

## The data gap: aluminium is missing where the farms are

This is the key result for the bulk fraction, and a limitation for the aquaculture question.
Aluminium coverage collapses toward the cages: the bulk sediments most in need of a
pristine-or-not verdict are the ones least able to receive one. It is a gap in the
aluminium, so it does not touch the sieved fractions, which are classified without it.

```{r}
#| label: coverage-fig
#| fig-cap: "Percentage of samples that can be classified, by distance to the nearest fish farm. The figure covers cobalt, copper and zinc in bulk, so it measures the aluminium gap alone. The sieved fractions are excluded because they need no aluminium and hold no samples within 5 km of a farm, so they would move every denominator without adding a case."
#| fig-width: 7.5
#| fig-height: 2.8
cov |>
  filter(axis == "distance to fish farm") |>
  mutate(band = factor(band, levels = c("<1km", "1-5km", "5-20km", ">20km"))) |>
  ggplot(aes(band, pct_classifiable)) +
  geom_col(fill = "#AA3377", width = 0.65) +
  geom_text(aes(label = paste0(pct_classifiable, "%")), vjust = -0.4, size = 3.2) +
  scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.1))) +
  labs(x = "distance to nearest fish farm", y = "% classifiable")
```

That gradient is real, but it is **not a property of near-cage sampling**. Splitting the
same measurements by data source shows coverage that does not depend on distance at all:

```{r}
#| label: coverage-source-table
covs |>
  filter(source %in% c("Vannmiljø", "Mareano")) |>
  mutate(band = factor(band, levels = c("<1km", "1-5km", "5-20km", ">20km"))) |>
  tidyr::pivot_wider(id_cols = band, names_from = source,
                     values_from = c(pct_classifiable, pct_of_band)) |>
  transmute(`Distance band` = band,
            `Vannmiljø, % classifiable` = `pct_classifiable_Vannmiljø`,
            `Mareano, % classifiable` = pct_classifiable_Mareano,
            `Vannmiljø share of band` = `pct_of_band_Vannmiljø`,
            `Mareano share of band` = pct_of_band_Mareano) |>
  arrange(`Distance band`) |>
  kable()
```

Within Vannmiljø, aluminium is reported for under 4% of measurements at **every** distance,
including beyond 20 km. Within Mareano it is reported for essentially all of them, including
under 1 km. What changes across the bands is not sampling practice near cages, it is
**which programme sampled there**: the < 1 km band is
`r covs$pct_of_band[covs$band == "<1km" & covs$source == "Vannmiljø"]`% Vannmiljø, while
beyond 20 km Mareano supplies
`r covs$pct_of_band[covs$band == ">20km" & covs$source == "Mareano"]`% of the measurements.

So the finding is sharper than "near-cage sediment cannot be assessed". It is that
**Vannmiljø does not report aluminium, and Vannmiljø is the programme that samples around
the farms.** That makes the recommendation concrete rather than general: aluminium (or
another grain-size normaliser) reported alongside the trace elements in Norwegian coastal
monitoring would close most of this gap on its own.

## Validation: what the distance trend can and cannot show

On the samples that *can* be classified, the pristine share drops as sites approach the
coast and the farms, which is the direction a verdict tracking real pressure should move.

The bands are not sampled by the same programmes, so this could have been a difference
between monitoring programmes wearing a distance label. Restricting the reference to one
aluminium basis made them homogeneous enough to test that directly, by repeating the
comparison **within a single source**, and it holds in both sources that span more than one
band:

```{r}
#| label: within-source-table
vsrc <- rd("refined_pristine_validation_source.csv")
vv <- val |> filter(axis == "distance to fish farm", rule == "ef")
vsrc |>
  mutate(Band = factor(band, levels = c("<1km", "1-5km", "5-20km", ">20km"))) |>
  transmute(Source = source, Band, samples = n,
            `% pristine (EF rule)` = pct_ef, `% pristine (strict)` = pct_strict) |>
  arrange(Source, Band) |>
  kable(format.args = list(big.mark = ","))
```

Mareano rises monotonically across the three bands it spans, from
`r vsrc$pct_ef[vsrc$source == "Mareano" & vsrc$band == "1-5km"]`% pristine at 1 to 5 km to
`r vsrc$pct_ef[vsrc$source == "Mareano" & vsrc$band == ">20km"]`% beyond 20 km. **Vannmiljø
does not**: it climbs from `r vsrc$pct_ef[vsrc$source == "Vannmiljø" & vsrc$band == "<1km"]`%
under 1 km to `r vsrc$pct_ef[vsrc$source == "Vannmiljø" & vsrc$band == "5-20km"]`% at 5 to
20 km and then falls to `r vsrc$pct_ef[vsrc$source == "Vannmiljø" & vsrc$band == ">20km"]`%
beyond 20 km, on `r vsrc$n[vsrc$source == "Vannmiljø" & vsrc$band == ">20km"]` classifiable
samples.

That reversal is not noise about sampling practice, it is the far band itself. Vannmiljø's
beyond-20 km measurements are largely its contaminated-seabed, industry and sewage
programmes, which is why the [Pressure-Based Background](./background-pressure.qmd) page
removes them before using that band as a background. So the honest statement is narrower than
it was: **within the survey programme that samples across the whole gradient the trend is
clean and monotonic**, and the one source that reverses it does so because its own far band is
a pressure programme rather than a remote one. Near-cage points still rest on few classifiable
samples, so read the trend, not the wiggle.

The gradient is **steeper** than it was before the verdicts were restricted to groups whose
normaliser works:
`r paste(vv$pct_pristine[order(factor(vv$band, levels = c("<1km","1-5km","5-20km",">20km")))], collapse = ", ")`%
across the four distance bands, against 15, 15, 28 and 49% when manganese and the sieved
fractions were included. Removing groups whose EF was noise-division sharpened the held-out
check rather than blunting it, which is the outcome a sound restriction should produce.

```{r}
#| label: validation-fig
#| fig-cap: "Percentage pristine among classifiable samples, by distance to the nearest fish farm and to coast (EF rule). Pristine rises with distance from pressure; the table above shows the same rise holds within each source separately."
#| fig-width: 8.5
#| fig-height: 3.4
val |>
  filter(rule == "ef") |>
  mutate(band = factor(band, levels = c("<1km","1-5km","5-20km",">20km","1-10km","10-50km",">50km"))) |>
  ggplot(aes(band, pct_pristine, group = axis, colour = axis)) +
  geom_line(linewidth = 0.7) + geom_point(size = 2) +
  scale_colour_manual(values = c("distance to fish farm" = "#AA3377",
                                 "distance to coast" = BLUE), name = NULL) +
  facet_wrap(~ axis, scales = "free_x") +
  labs(x = NULL, y = "% pristine (classifiable)") +
  theme(legend.position = "none")
```

::: {.callout-note}
"Pristine" here means at or below the local background by a grain-size-controlled measure,
per element, not an absolute or a site-level judgement. The strict rule no longer inherits
an unreliable mixture threshold: where the distribution does not support one, that criterion
is dropped rather than applied (see [Method Revisions](./method-revisions.qmd)). The honest headline is the coverage gap: the classification is trustworthy where
aluminium exists, which is most of the open-shelf data but almost none of the near-cage data,
so the aquaculture sites remain largely unassessable from these measurements alone.

Three further limitations belong on the record. The distance gradient this page validates
against is weaker than it looks: on raw concentrations it does not survive matching within a
Norwegian municipality, and a sixth of the near-cage measurements were taken outside the
farm's operating window ([Pressure Controls](./pressure-controls.qmd)). The verdicts here are
aluminium-normalised, which is a stronger control than the raw gradient, but they can only be
checked against it. What no distance design on this data can separate is a fjord-wide effect
from no effect, because sediment 5 km from a farm in a farming municipality is not unfarmed
sediment; that needs farmed and unfarmed fjords matched on basin, which this database cannot
identify. And the [Enrichment Map](./enrichment-map.qmd) applies
these thresholds to per-site averages, which makes it a **different classifier** from the
per-measurement one tabulated here, not just a coarser drawing of it. This is a
synthesis of the [Global Background](./background.qmd),
[Enrichment Factor](./background-ef.qmd) and
[Distribution-Mixture](./background-mixture.qmd) pages; iodine is too sparse throughout.
:::
