---
title: "Enrichment Factor"
subtitle: "Classifying samples against a local background, and testing the aquaculture signal"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The background pages estimate a baseline; this one turns it into a per-sample verdict. The
**enrichment factor** is

$$\text{EF} = \frac{(\text{metal} / \text{Al})_\text{sample}}{(\text{metal} / \text{Al})_\text{background}}$$

against a **local, data-driven background**: the offshore (> 10 km from the coast)
**median** of metal / Al, per element and fraction. This follows EFSA's guidance to use a
local background and to **avoid** literature crustal values (Turekian and Wedepohl), which
distort marine backgrounds. Because aluminium is the normaliser, EF already controls for
grain size, so it separates real enrichment from merely muddier sediment.

**EF < 1** is read as adequate (at or below the local background); the classes 1-2, 2-5
and > 5 mark rising enrichment **relative to the local offshore background** (so these are
not the usual crustal-referenced numbers).

::: {.callout-important}
## The reference is restricted to one aluminium basis

The sources do not all measure aluminium the same way: some report near-total aluminium,
some report what an acid extraction leaches, which is two to three times less. Aluminium is
the denominator of every number on this page, so a reference pooled over both meant nothing
and pushed the under-reported samples toward falsely pristine.

Each fraction's reference is therefore computed within a **single basis**, inferred per
sample from Fe/Al, and a sample on the other basis, or with no iron to place it, is left
**unclassified** rather than judged against a reference it does not belong to. Bulk uses the
extraction basis; the sieved fractions use the total basis. It costs roughly half the
classifiable rows.

Two consequences worth carrying. Bulk aluminium here is acid-leachable, so **bulk EF is
internally comparable but not comparable with literature EF values**. And the classifiable
share on the [Pristine Classification](./background-pristine.qmd) page is much lower than it
was, because samples are now withdrawn rather than misjudged.

[Method Revisions](./method-revisions.qmd) sets out what this changed and by how much.
:::

::: {.callout-important}
## And only where aluminium predicts the metal

Dividing by aluminium is a grain-size control only if aluminium carries the metal. Tested by
regression on the offshore reference, it does for **cobalt, copper and zinc in bulk**
(R² 0.46 to 0.58) and for nothing else: manganese, molybdenum and selenium in bulk, and every
sieved fraction of every element, sit at 0.10 or below, two of them at or below zero.

Those groups therefore carry **no enrichment factor and no pristine verdict**. Their
concentrations, their offshore references and their distributions are still published; what
is withheld is the verdict, on the same principle that already withholds one where the
aluminium basis is wrong. [Regression Normalisation](./background-regression.qmd) has the
measurements behind it.
:::

The background here uses **no aquaculture information**: it is the offshore distribution
alone. That keeps the estimate independent of the pressure it is meant to detect. The
distance-to-aquaculture panels below are therefore an **evaluation**, not an input, testing
whether the enrichment the factor reports lines up with the known pressure gradient (the
[Pressure-Based Background](./background-pressure.qmd) page).

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
efd <- rd("refined_ef_dist.csv")
efp <- rd("refined_ef_pressure.csv")
efb <- rd("refined_ef_background.csv")
efs <- rd("refined_ef_source.csv")
efb2 <- rd("refined_ef_basis.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
bin_levels  <- c("<1km", "1-5km", "5-20km", ">20km")
blank <- function(x) ifelse(is.na(x), "", as.character(x))
theme_set(theme_minimal(base_size = 12))
```

## How enriched is the data?

The share of samples in each EF class, for the bulk groups that carry an enrichment factor.
A large green block means much of the data sits at or below the local background, a
pristine-candidate pool.

```{r}
#| label: class-fig
#| fig-cap: "EF class shares for the three groups that carry an enrichment factor. Between 43 and 48% of each sits at or below the local background, and the remainder is mostly in the 1-2 band. The other elements and every sieved fraction are absent because aluminium does not predict them."
#| fig-width: 8
#| fig-height: 3.6
efd |>
  filter(cat == "bulk", reliable == TRUE, !is.na(pct_lt1)) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  select(symbol, `<1` = pct_lt1, `1-2` = pct_1_2, `2-5` = pct_2_5, `>5` = pct_gt5) |>
  pivot_longer(-symbol, names_to = "EF", values_to = "pct") |>
  mutate(EF = factor(EF, levels = c("<1", "1-2", "2-5", ">5"))) |>
  ggplot(aes(pct, symbol, fill = EF)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("<1" = "#228833", "1-2" = "#CCBB44",
                               "2-5" = "#EE6677", ">5" = "#AA3377")) +
  labs(x = "% of samples", y = NULL, fill = "EF")
```

The background reference and the EF summary per element and fraction. Groups with no verdict
keep their row and their n, so it is visible that they exist and why they are blank.

```{r}
#| label: summary-table
efd |>
  filter(reliable == TRUE) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  left_join(efb |> select(symbol, cat, bg_ratio_al), by = c("symbol", "cat")) |>
  transmute(Element = symbol, Fraction = cat, n,
            `background metal/Al` = bg_ratio_al,
            `median EF` = blank(ef_p50), `P90 EF` = blank(ef_p90),
            `% adequate (EF<1)` = blank(pct_lt1),
            `% adequate (P90 ref)` = blank(pct_lt1_p90ref),
            verdict = case_when(withheld ~ "withheld (below LOQ)",
                                unnormalisable ~ "withheld (Al does not predict it)",
                                TRUE ~ "")) |>
  arrange(Element, Fraction) |>
  kable(format.args = list(big.mark = ","))
```

One reading caution on the sieved rows. A sieved fraction is already a grain-size control:
the sample has been cut to particles below 63 or 20 um, so most of the texture variation is
gone before the chemistry starts. Dividing by aluminium on top of that controls grain size
twice, and the second control acts on whatever variation in aluminium is left, which is
mineralogy rather than texture. That is the most likely reason the sieved metal / Al
backgrounds sit so far below the bulk ones for molybdenum (8.2e-5 sieved63 against 4.4e-4
bulk). Bulk EF is the grain-size-controlled number; sieved EF is closer to a
mineralogy-controlled one, and the two are not the same quantity even though they share a
name. Tested directly, aluminium explains under 9% of the metal's variation in every sieved
group and under 0.1% in some, against about half in bulk for cobalt, copper and zinc: see
[Regression Normalisation](./background-regression.qmd), which also shows that among the bulk
elements only those three have a normaliser that predicts them at all. Read the manganese,
molybdenum and selenium EF numbers, and every sieved EF number, with that in mind.

## Is the reference comparable across sources?

Per source, bulk: how much of the pooled reference that source supplies, how its own
offshore metal / Al reference compares with the pooled one (**rel.**, above 1 meaning the
source runs richer than the pool), and the share it scores adequate under each. Reliable
groups only.

```{r}
#| label: source-table
efs |>
  filter(cat == "bulk", reliable == TRUE) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         set = factor(set, levels = c("pooled", "adopted"))) |>
  transmute(Reference = set, Element = symbol, Source = source, n,
            `% of reference` = bg_share, `rel.` = bg_rel,
            `% adequate (this ref)` = pct_lt1,
            `% adequate (own ref)` = pct_lt1_own) |>
  arrange(Reference, Element, Source) |>
  kable(format.args = list(big.mark = ","))
```

Read the `pooled` block first: that is how the reference was built before. The same seabed
is 97% adequate for cobalt if ICES-DOME measured it and 41% if Mareano did, while under each
source's own reference every group lands between 35 and 48%, which is what a median
reference has to give. The spread between the sources was the artefact and the convergence
was the arithmetic.

In the `adopted` block Mareano's **rel.** sits at 1.00 to 1.02, where it ran to 1.21 before.
A residual gap remains on ICES-DOME's extraction-basis rows, a few hundred pan-European
samples against Mareano's Norwegian shelf, so part of that is the regional difference no
source comparison can rule out.

The strata themselves, and what the restriction set aside:

```{r}
#| label: basis-table
efb2 |>
  mutate(Fraction = factor(cat, levels = c("bulk", "sieved63", "sieved20"))) |>
  transmute(Fraction, Basis = al_basis, Used = ifelse(adopted, "yes", "no"),
            n, `% of fraction` = pct_of_frac, `offshore rows` = n_offshore) |>
  arrange(Fraction, desc(Used), desc(n)) |>
  kable(format.args = list(big.mark = ","))
```

## Does the aquaculture signal survive normalisation?

The [Pressure-Based](./background-pressure.qmd) page found a near-cage rise in the **raw**
values, part of which could be grain size. EF removes grain size, so if the rise persists
it is real. The median EF by distance to the nearest farm (bulk, Norway):

```{r}
#| label: pressure-fig
#| fig-cap: "Median EF by distance to aquaculture (bulk, log scale, dashed line = background), for the three groups whose normaliser predicts the metal. Copper and zinc stay enriched near the cages after grain-size normalisation; cobalt does not. Manganese, molybdenum and selenium are absent: aluminium does not predict them, so an EF gradient for them would not be a grain-size-controlled result."
#| fig-width: 8.5
#| fig-height: 4.4
efp |>
  filter(cat == "bulk", symbol %in% c("CO","CU","MN","MO","SE","ZN")) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         aq_bin = factor(aq_bin, levels = bin_levels)) |>
  ggplot(aes(aq_bin, ef_p50, group = symbol, colour = symbol)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  labs(x = "distance to nearest aquaculture farm", y = "median EF (log)")
```

**Copper** climbs to a median EF of about 2.4 in the 1 to 5 km band and falls to background
beyond 20 km; **zinc** climbs to about 2. Both are grain-size controlled, because aluminium
genuinely predicts both, so this is the strongest evidence on the site for a near-cage effect.
**Cobalt** stays at background throughout. Normalising also **sharpens the picture** for
copper: in the raw values it was actually *lower* right at the cages (those sites are
sandier), but per aluminium it is clearly enriched there, a gradient the raw numbers hid.

**Molybdenum and selenium are no longer on this figure, and that is a correction.** The site
used to report that their near-cage enrichment survived grain-size normalisation. It did, but
the normalisation was not doing anything: aluminium is uncorrelated with molybdenum
(Spearman's rho 0.00) and nearly so with selenium (0.15), so dividing by it leaves their
gradient essentially untouched. A signal cannot survive a test that was never applied. Their
raw near-cage ratios are real observations and live on the
[Pressure-Based Background](./background-pressure.qmd) page; what cannot be claimed for them
is grain-size control. Manganese leaves the figure for the same reason.

Two limits on what remains. The copper and zinc gradient on **raw** values does not survive a
geographic control: matched within a municipality rather than against the open shelf, copper
falls from 3.2 times to about 1.1 and zinc from 1.8 to about 0.9. This figure is the better
controlled comparison, but it rests on the roughly 1% of near-cage samples that carry
aluminium, far too few to match within a municipality in turn. So the aluminium-normalised
gradient is at once the strongest evidence for a near-cage effect and the least representative
subset of the data. See [Pressure Controls](./pressure-controls.qmd).

::: {.callout-note}
EF depends on its background: here the local offshore median of metal / Al, a defensible
data-driven choice, but a different reference (a lower percentile, or the far-from-cage
median) would shift every EF up or down. So "EF < 1 = adequate" is relative to this
background, not an absolute verdict. The offshore P90 column beside it reads the same data
against the other conventional reference, and the gap between the two columns is the honest
width of the answer. Aluminium is measured for only a minority of bulk samples and only some
of those sit on the adopted basis, so EF covers that subset; iodine is too sparse for any of
this, and **selenium and molybdenum carry no verdict at all**, because most of their
measurements fell below the limit of quantification and were removed upstream (see
[Method Revisions](./method-revisions.qmd)). Nor do manganese or any sieved fraction, because
aluminium does not predict them. What is left is cobalt, copper and zinc in bulk: three
groups and about 11,300 measurements, from which every enrichment number on this site now
comes. Read with the three background pages, which set the reference EF builds on.
:::
