---
title: "Global Background"
subtitle: "Estimating a natural background concentration per element and fraction"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

A **background** concentration is what a sediment would carry without local human input,
the baseline that concentrations elsewhere are judged against. There is no single way to
define it, so this page sets out two complementary estimates for each element and
fraction and compares them:

- **(A) global distribution**: the percentiles of the value over all sites, the natural
  range. The **90th percentile** (an approach used by EFSA) is one point on it.
- **(B) offshore**: the same percentiles restricted to sites **far from the coast**
  (more than `r 10` km). Local pressure, including aquaculture, is coastal, so an
  offshore-only distribution is a cleaner background; comparing it to the global one
  shows how much the coast lifts the numbers.

Everything is on the **raw** value (mg/kg), split by fraction into **bulk**,
**sieved63** (< 63 um) and **sieved20** (< 20 um), which are never pooled. Distributional
outliers are dropped, but only extreme ones: a value is flagged only if it is both more
than 4 MADs from its group median on a log scale **and** a full order of magnitude away
from that median, which catches registration errors rather than trimming the tail. It
removes 414 of 115,820 measurements, 0.36%. Grain-size-normalised backgrounds
(metal / Al, metal / Fe) are a later page.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
pct  <- rd("refined_background_percentiles.csv")
cmp  <- rd("refined_background_compare.csv")
meta <- rd("refined_background_meta.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
cat_levels  <- c("bulk", "sieved63", "sieved20")
pct$symbol <- factor(pct$symbol, levels = elem_levels)
pct$cat    <- factor(pct$cat, levels = cat_levels)
cmp$symbol <- factor(cmp$symbol, levels = elem_levels)
cmp$cat    <- factor(cmp$cat, levels = cat_levels)
BLUE <- "#4477AA"; RED <- "#EE6677"
theme_set(theme_minimal(base_size = 12))
```

## A. The global range

The spread of each element by fraction, drawn from the global percentiles (box = the 25th
to 75th, line = median, whiskers = 10th to 90th). Reliable groups only (at least
`r meta$min_n` values); iodine, and selenium in the sieved fractions, are too sparse to
show.

```{r}
#| label: range-fig
#| fig-cap: "Global distribution of each element by fraction (log scale): box is the interquartile range, whiskers the 10th to 90th percentiles. Sieved fractions concentrate metals above bulk."
#| fig-width: 8.5
#| fig-height: 4.6
pct |>
  filter(subset == "global", reliable) |>
  ggplot(aes(cat)) +
  geom_boxplot(aes(ymin = p10, lower = p25, middle = p50, upper = p75, ymax = p90,
                   fill = cat), stat = "identity", width = 0.6) +
  scale_y_log10() +
  scale_fill_manual(values = c(bulk = "#4477AA", sieved63 = "#66CCEE",
                               sieved20 = "#228833"), guide = "none") +
  facet_wrap(~ symbol, scales = "free_y") +
  labs(x = NULL, y = "concentration, mg/kg (log)")
```

The **90th percentile** per element and fraction, the EFSA-style background value:

```{r}
#| label: p90-table
pct |>
  filter(subset == "global", reliable) |>
  transmute(Element = symbol, Fraction = cat, n, `median` = p50, `P90 (mg/kg)` = p90) |>
  arrange(Element, Fraction) |>
  kable(format.args = list(big.mark = ","))
```

## B. Offshore versus global

Restricting to offshore sites lowers the background where the coast adds metal, and
leaves it unchanged where it does not. The ratio below is the offshore 90th percentile
over the global one: **well under 1 means a strong coastal signal removed**.

```{r}
#| label: shift-fig
#| fig-cap: "Offshore (>10 km) 90th percentile relative to the global 90th percentile. Below the dashed line, offshore is cleaner (coastal enrichment removed); at or above it, the coast is not the driver."
#| fig-width: 8
#| fig-height: 3.8
cmp |>
  filter(!is.na(shift_p90), n_off10 >= meta$min_n) |>
  mutate(lab = paste(symbol, cat)) |>
  ggplot(aes(shift_p90, reorder(lab, shift_p90))) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey60") +
  geom_col(aes(fill = shift_p90 < 1)) +
  geom_text(aes(label = sprintf("%.2f", shift_p90)), hjust = -0.15, size = 3) +
  scale_fill_manual(values = c(`TRUE` = BLUE, `FALSE` = RED), guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "offshore P90 / global P90", y = NULL)
```

Read together with the numbers, the picture is:

- **Copper and zinc in bulk** drop sharply offshore (copper to about 0.4 of the global
  value, zinc to about 0.6): a clear coastal enrichment, and the offshore figure is the
  cleaner background for these.
- **Manganese and cobalt in bulk** barely move (manganese flat, cobalt slightly lower),
  so the coast is not their main driver. **Molybdenum** behaves differently by fraction:
  in **bulk** it drops steeply offshore (P90 from about 18 to 3), reflecting enrichment
  in coastal anoxic basins (restricted fjords) rather than on the open shelf, while in
  the **sieved** fractions it rises. These redox-sensitive elements are a reminder that a
  lower offshore value is not always the more natural one.
- **The sieved fractions** hardly change offshore (ratios near or above 1): sieved
  sampling is dominated by specific monitoring programmes rather than a clean
  coast-to-offshore gradient, so the offshore filter does less for them.

### How much does the 10 km cutoff matter?

Enough to matter, and most where the signal is strongest. Between the 10 km and the 50 km
cutoff the bulk P90 falls **34% for copper** (34 to 22.4) and **40% for zinc** (130 to
77.4), falls 11% for molybdenum, and *rises* 26% for manganese (932 to 1,173); only cobalt
(+2%) and selenium (-2%) are stable. Copper and zinc are the two elements this page
identifies as having the clearest coastal signal, and they are the two that keep falling as
the cutoff moves further out.

```{r}
#| label: sensitivity-table
cmp |>
  filter(cat == "bulk", n_off10 >= meta$min_n) |>
  transmute(Element = symbol, `global P90` = p90_global,
            `>10 km` = p90_off10, `>20 km` = p90_off20, `>50 km` = p90_off50) |>
  arrange(Element) |>
  kable(caption = "Bulk 90th-percentile background (mg/kg) at three offshore cutoffs.")
```

Two things follow. Ten kilometres is evidently not far enough out for copper and zinc,
whose "> 10 km" background still carries coastal input, so the > 20 km column is the more
conservative estimate for those two. And the choice propagates: the
[Enrichment Factor](./background-ef.qmd) page builds its reference from this same > 10 km
subset, so every copper and zinc enrichment factor is anchored to the least clean of the
three cutoffs and is, if anything, too forgiving.

::: {.callout-note}
This is one family of background estimates, not the final answer. "Offshore" is a coarse
proxy for pristine (a site a few km out in a pressured fjord is not clean, and a coastal
site in a remote area may be), and the redox-sensitive elements show that a lower value is
not always the more natural one. The 90th percentile is a conservative upper bound of the
background, not a typical value. Later pages will add normalised backgrounds and pressure
based definitions (distance to aquaculture, enrichment factors) to triangulate. Iodine and
selenium have too few measurements for a reliable background, especially when sieved.

These percentiles also describe **the samples we hold, not the seabed**. All five sources
are monitoring programmes whose sampling is driven by regulation and interest, so this is a
convenience sample and its percentiles are not spatially representative. Re-sampled sites
contribute one row per measurement, so heavily monitored locations, typically the pressured
ones, carry more weight in these percentiles than lightly sampled ones do.
:::
