---
title: "Background Summary"
subtitle: "The four background estimates side by side, per element and fraction"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The four Background pages each define a background differently, and none of them is
privileged. The [global and offshore percentiles](./background.qmd) are spatial, the
[distribution mixture](./background-mixture.qmd) is distributional, the
[grain-size-normalised](./background-gsnorm.qmd) values control for texture, and the
[pressure bands](./background-pressure.qmd) are defined by distance to a known source.
They rest on different assumptions and partly on different samples, so **where they agree,
the agreement is the evidence**, and where they disagree the disagreement is the finding.

This page recomputes nothing. Every number below is read from the summary files the four
analysis pages write, so it cannot say anything they do not.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
pct  <- rd("refined_background_percentiles.csv")
cmp  <- rd("refined_background_compare.csv")
bmet <- rd("refined_background_meta.csv")
mix  <- rd("refined_mixture_components.csv")
gsp  <- rd("refined_gsnorm_percentiles.csv")
prs  <- rd("refined_pressure_compare.csv")
sw   <- rd("refined_background_sitewise.csv")
ps   <- rd("refined_pristine_summary.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
cat_levels  <- c("bulk", "sieved63", "sieved20")
theme_set(theme_minimal(base_size = 12))

# reliable groups, exactly as the Global Background page defines them
ok <- pct |> filter(subset == "global", reliable) |> select(symbol, cat)

# the Al-normalised background, reliable groups only
al <- gsp |>
  filter(basis == "al", subset == "global", reliable) |>
  select(symbol, cat, p90_al = p90)

# the Norwegian far-from-farm band, where it has enough measurements. The CLEANED
# far band: the raw >20 km pool is over half stated-pressure monitoring (contaminated
# seabed, industry, sewage), which is not a background. See the Pressure-Based
# Background page.
nor <- prs |>
  transmute(symbol, cat, p90_nor = p90_far_clean, n_nor = n_far_clean) |>
  mutate(p90_nor = if_else(!is.na(n_nor) & n_nor >= 30, p90_nor, NA_real_)) |>
  select(symbol, cat, p90_nor)

held <- ps |> distinct(symbol, withheld) |> filter(withheld) |> pull(symbol)

master <- ok |>
  left_join(cmp |> select(symbol, cat, n = n_global, p90_global, p90_off10, p90_off20),
            by = c("symbol", "cat")) |>
  left_join(mix |> select(symbol, cat, threshold, usable), by = c("symbol", "cat")) |>
  left_join(al, by = c("symbol", "cat")) |>
  left_join(nor, by = c("symbol", "cat")) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         cat = factor(cat, levels = cat_levels),
         verdict = case_when(is.na(usable) ~ NA_character_,
                             usable ~ "two populations",
                             TRUE ~ "continuum"),
         withheld = as.character(symbol) %in% held) |>
  arrange(symbol, cat)

# the spread of the estimates that share units (mg/kg). A threshold from a fit that does
# not separate is not a background, so it does not enter the spread.
spread_of <- function(m) apply(m, 1, function(v) {
  v <- v[!is.na(v) & v > 0]
  if (length(v) < 2) NA_real_ else round(max(v) / min(v), 2)
})
master <- master |>
  mutate(thr_ok = if_else(usable %in% TRUE, threshold, NA_real_)) |>
  mutate(spread_all = spread_of(cbind(p90_global, p90_off10, p90_off20, thr_ok, p90_nor)),
         spread_off = spread_of(cbind(p90_off10, p90_off20, thr_ok, p90_nor)))

sig <- function(x, d = 4) ifelse(is.na(x), "", formatC(signif(x, d), format = "g",
                                                      big.mark = ",", digits = d))
blank <- function(x) ifelse(is.na(x), "", as.character(x))
```

## The master table

One row per element and fraction, reliable groups only (at least `r bmet$min_n`
measurements, the rule the analysis pages use). The four estimates that share units are in
mg/kg; the aluminium-normalised background is a dimensionless metal / Al ratio and is not
comparable with them.

```{r}
#| label: master-table
master |>
  transmute(Element = symbol, Fraction = cat, n,
            `global P90` = sig(p90_global),
            `offshore >10 km` = sig(p90_off10),
            `offshore >20 km` = sig(p90_off20),
            `mixture threshold` = sig(threshold),
            `mixture fit` = blank(verdict),
            `metal/Al P90` = sig(p90_al, 3),
            `Norway >20 km (cleaned)` = sig(p90_nor)) |>
  kable(format.args = list(big.mark = ","), align = "llrrrrrlrr")
```

Empty cells are absences, not zeroes, and there are three reasons for one:

- **metal/Al** is blank where too few samples in the group carry aluminium. Bulk aluminium
  coverage is about a third of samples overall, and much lower in some programmes.
- **Norway >20 km** is blank wherever fewer than `r bmet$min_n` Norwegian measurements sit
  beyond 20 km of a farm, which is most of the sieved fractions. The band is Norway-only by
  construction.
- **Iodine** has no row at all (10 bulk and 5 sieved measurements), and neither does
  selenium sieved20 (one measurement).

Two rows carry numbers that are **not** background values. Molybdenum and selenium have
most of their measurements below the limit of quantification, removed during cleaning, so
what remains is the detected upper tail and every percentile of it sits far above the real
one. Their concentrations are published; **no background verdict is issued for either**,
here or anywhere on this site. See [Method Revisions](./method-revisions.qmd).

## How well do the estimates agree?

The convergence indicator is the ratio of the largest to the smallest of the available
mg/kg estimates for a group: the global P90, both offshore P90s, the Norwegian far-band
P90, and the mixture threshold where the fit actually separates two populations. A ratio
near 1 means the definitions land in the same place and the background is well constrained.
The second column drops the global P90, which is not a background wherever the coast lifts
it, and so isolates how much of the disagreement the global figure alone causes.

```{r}
#| label: spread-table
master |>
  filter(!is.na(spread_all)) |>
  transmute(Element = symbol, Fraction = cat,
            `spread, all estimates` = spread_all,
            `spread, excluding global P90` = spread_off,
            `background verdict` = if_else(withheld, "withheld (below LOQ)", "")) |>
  kable(align = "llrrl")
```

```{r}
#| label: spread-fig
#| fig-cap: "Convergence of the mg/kg background estimates per element and fraction: the ratio of the largest to the smallest, with and without the global P90. Points at 1 mean the definitions agree. The gap between the two points is the damage the global P90 does on its own."
#| fig-width: 8
#| fig-height: 4.2
master |>
  filter(!is.na(spread_all)) |>
  transmute(lab = paste(symbol, cat), all = spread_all, off = spread_off) |>
  pivot_longer(c(all, off), names_to = "set", values_to = "spread") |>
  mutate(set = factor(set, levels = c("all", "off"),
                      labels = c("all estimates", "excluding global P90"))) |>
  ggplot(aes(spread, reorder(lab, spread), colour = set)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey60") +
  geom_line(aes(group = lab), colour = "grey80", linewidth = 0.6) +
  geom_point(size = 2.6, alpha = 0.9) +
  scale_colour_manual(values = c("all estimates" = "#EE6677",
                                 "excluding global P90" = "#4477AA"), name = NULL) +
  labs(x = "max / min of the available mg/kg estimates", y = NULL)
```

The result is cleaner than expected. **In bulk, once the global P90 is set aside, every
element but manganese agrees within about 1.5 times**: cobalt to 1.01, selenium to 1.35,
copper to 1.48, zinc to 1.50, molybdenum to 1.51. Three definitions built on different
assumptions, one of them distributional rather than spatial and one of them Norwegian, put
the background in the same place. That is the strongest thing this page has to say.

Where the indicator is large, it is large for three distinct reasons.

**The global P90 is the outlier, not the background.** Molybdenum in bulk spans 7.19 times
across all estimates and 1.51 without the global figure: its global P90 of 17.6 mg/kg sits
against 2.4 to 3.7 everywhere else, because coastal anoxic basins in restricted fjords
enrich molybdenum for redox reasons. Copper (3.74 to 1.48) and zinc (2.53 to 1.50) are the
same story in milder form, and are the elements the
[10 km cutoff sensitivity](./background.qmd#how-much-does-the-10-km-cutoff-matter) already
flagged: copper's P90 falls 34 to 23 mg/kg between the 10 and 20 km cutoffs, zinc's 130 to
87. For these three the global P90 is a percentile of the coast, and should not be used as
a background value.

**The mixture threshold sits far below the spatial estimates for manganese.** Manganese in
bulk spans `r round(master$spread_all[master$symbol=="MN" & master$cat=="bulk"], 2)` times and
dropping the global P90 does not help at all, because the two extremes are its mixture
threshold (`r sig(master$threshold[master$symbol=="MN" & master$cat=="bulk"])` mg/kg) and the
Norwegian far band (`r sig(master$p90_nor[master$symbol=="MN" & master$cat=="bulk"])`). Both are
defensible and they mean different things: the mixture finds the lower mode of a
distribution, the far band is a place with no farm nearby. For a redox-mobile element that
precipitates offshore, those need not coincide, and here they do not.

**The estimates genuinely disagree in the sieved20 fraction.** Manganese sieved20 spans
4.45 times and zinc sieved20 3.39, again with the mixture threshold at one end. The offshore
filter does almost nothing for the sieved fractions (their offshore/global ratios are near or
above 1), so nothing pulls the spatial estimates together either, and no single number should
be quoted for them.

### Sensitivity to sample weighting

The percentiles above count every measurement once, and revisited sites are counted many
times. Weighting one value per site instead moves the offshore bulk P90 down for all four elements
that carry a background verdict:

```{r}
#| label: sitewise-table
sw |>
  filter(subset == "offshore>10km", reliable, cat == "bulk",
         !symbol %in% held) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  arrange(symbol) |>
  transmute(Element = symbol, sites = n_sites,
            `P90 per measurement` = sig(p90_meas),
            `P90 per site` = sig(p90_site),
            ratio = p90_ratio) |>
  kable(format.args = list(big.mark = ","))
```

Cobalt and zinc barely move, copper falls about 15% and manganese about 32%. The headline
figures stay per measurement, but for manganese the two weightings differ enough that the
range, not either number, is the answer. The [Global Background](./background.qmd#does-repeat-sampling-tilt-the-percentiles)
page shows why: revisited sites are the pressured ones.

## What to use, element by element

**Cobalt.** Use the offshore P90, about `r sig(master$p90_off20[master$symbol=="CO" & master$cat=="bulk"])`
mg/kg in bulk. It is the best-constrained element on the page: the 10 km, 20 km and Norwegian
far-band figures sit within 1% of one another, the global P90 is only 14% above them, and the
coast barely lifts it (offshore/global ratio 0.88). The mixture fit does not separate two populations for cobalt,
so its threshold of 6.45 mg/kg is not a background and is not used downstream.

**Copper.** Use the **>20 km** offshore P90, about
`r sig(master$p90_off20[master$symbol=="CU" & master$cat=="bulk"])` mg/kg in bulk,
corroborated by the Norwegian far band at `r sig(master$p90_nor[master$symbol=="CU" & master$cat=="bulk"])`
and the mixture crossover at `r sig(master$threshold[master$symbol=="CU" & master$cat=="bulk"])`.
Do not use the global P90 of `r sig(master$p90_global[master$symbol=="CU" & master$cat=="bulk"])`,
and do not use the 10 km figure either: ten kilometres is not far enough out for copper.
Note also that most of copper's apparent coastal enrichment is
[grain size](./background-gsnorm.qmd), not pressure: Al-normalising lifts its offshore/global
shift from 0.40 to 0.87.

**Manganese.** No single value. The offshore filter does not clean manganese, it raises it
(the >20 km P90 of `r sig(master$p90_off20[master$symbol=="MN" & master$cat=="bulk"])` mg/kg
exceeds the global `r sig(master$p90_global[master$symbol=="MN" & master$cat=="bulk"])`), and
the Norwegian far band is higher still. Manganese is redox-mobile and precipitates in
oxygenated offshore sediment, so a higher offshore value is the natural one, not a
contaminated one. The mixture threshold of
`r sig(master$threshold[master$symbol=="MN" & master$cat=="bulk"])` mg/kg is the one estimate
that separates cleanly, and it is roughly half the spatial figures. Quote a range of about
500 to 1,200 mg/kg in bulk and say which definition you used.

**Molybdenum.** **No background value is issued.** 52% of molybdenum measurements were
below the limit of quantification and were removed before these percentiles were computed,
86% of them in Mareano, which supplies almost all of the far-from-coast reference. What
survives is the detected tail. Had a value been issued it would have been the >20 km figure
of about `r sig(master$p90_off20[master$symbol=="MO" & master$cat=="bulk"])` mg/kg, with the
global P90 of `r sig(master$p90_global[master$symbol=="MO" & master$cat=="bulk"])` discarded
as a percentile of coastal anoxic basins. Molybdenum's near-cage enrichment is a separate
question and it survives, reduced: 13.5 times against the open shelf, 8.0 times once only
samples contemporary with an operating farm are counted, on 28 measurements. See
[Pressure Controls](./pressure-controls.qmd) and the
[Enrichment Summary](./enrichment-summary.qmd).

**Selenium.** **No background value is issued**, for the same reason and more of it: 69% of
selenium measurements were below the limit of quantification. The bulk estimates are tightly
clustered (a spread of 1.35), but that agreement is between four statistics of the same
truncated distribution and is not evidence.

**Zinc.** Use the **>20 km** offshore P90, about
`r sig(master$p90_off20[master$symbol=="ZN" & master$cat=="bulk"])` mg/kg in bulk,
corroborated by the Norwegian far band at `r sig(master$p90_nor[master$symbol=="ZN" & master$cat=="bulk"])`
and the mixture crossover at `r sig(master$threshold[master$symbol=="ZN" & master$cat=="bulk"])`.
The global P90 of `r sig(master$p90_global[master$symbol=="ZN" & master$cat=="bulk"])` is a
coastal figure. As with copper, part of the coastal signal is grain size, though less of it
(the Al-normalised shift rises from 0.59 to 0.74).

**Iodine.** Too sparse for any of this: 10 bulk and 5 sieved measurements in the whole
database, against a minimum of `r bmet$min_n`. No estimate of any kind is offered.

::: {.callout-note}
Three cautions apply to every number above.

**"Offshore" is a proxy for pristine, and a coarse one.** A site a few kilometres out in a
pressured fjord is not clean, and a coastal site in a remote area may be. The Norwegian
distance-to-farm band is the only estimate here defined by an actual pressure, and it covers
one country.

**The 90th percentile is a conservative upper bound, not a typical value.** It is the EFSA
style convention these pages follow, and for the skewed elements it is several times the
median. A background used for screening and a background used for describing normal
sediment are not the same statistic.

**These are percentiles of the samples we hold, not of the seabed.** All five sources are
monitoring programmes whose sampling is driven by regulation and interest, so harbours,
fjords and impact studies are heavily over-represented and open shelf is not. The
site-weighting table above corrects for revisited sites being counted many times; nothing
corrects for the places nobody sampled. Adding a source will move these percentiles.

Where aluminium is missing is a property of the reporting programme, not of the place. The
metal/Al column covers whichever samples carry aluminium, and coverage varies from near
complete in Mareano to 1 to 4% in Vannmiljø at every distance from the coast. A reader who
takes a coverage gradient at face value will conclude that near-cage sediment is
undermeasured; what is actually undermeasured is everything one programme touches. The
[Enrichment Summary](./enrichment-summary.qmd) sets this out in full.

The three grain-size normalisers also disagree with each other and cover different samples,
so the metal/Al column is one choice among three and is aluminium's answer alone; the
[Grain-Size-Normalised Background](./background-gsnorm.qmd) page sets out the disagreement, and
[Regression Normalisation](./background-regression.qmd) shows that aluminium's grip varies by
element: about half the variation of cobalt, copper and zinc in bulk, under a tenth of
manganese, molybdenum and selenium, and essentially none in any sieved fraction. Only those
three bulk groups now receive an enrichment or pristine verdict anywhere on this site. The
metal/Al column above is still published for every group, as a description rather than a
verdict.
The [Pressure-Based Background](./background-pressure.qmd) page is Norway-only and is
deliberately held out of the estimation, so its far band is corroboration here and evidence
on the [Enrichment Summary](./enrichment-summary.qmd), never an input. Its *near*-cage ratios
are upper bounds, for reasons set out on [Pressure Controls](./pressure-controls.qmd); the far
band used here is unaffected, being defined by the absence of a farm rather than by proximity
to one.
:::
