---
title: "Distribution-Mixture Background"
subtitle: "Separating a background population from an enriched one, from the distribution alone"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The other pages lean on where or how far from pressure a sample sits. This one uses only
the **shape of the distribution**. A concentration distribution is often a mixture of two
populations: a natural **background** and an **enriched** one. Fitting a two-component
Gaussian mixture to the log concentration separates them, and the **crossover** between the
components is a background threshold (the upper bound of the natural population), with no
external reference, no offshore or aquaculture filter.

It is fitted on the **raw** value (mg/kg) per element and fraction. Raw values carry grain
size, so a muddy-versus-sandy split can imitate a background-versus-enriched one; this is
the distribution-only view, to be read against the offshore, grain-size-normalised and
enrichment-factor pages.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
comp <- rd("refined_mixture_components.csv")
hist <- rd("refined_mixture_hist.csv")
offc <- rd("refined_background_compare.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
theme_set(theme_minimal(base_size = 12))
```

## The two populations

Each panel is the log concentration (bulk): the grey histogram, the fitted **background**
(blue) and **enriched** (red) components, their sum (black), and the crossover threshold
(dashed). Where the two humps are distinct the split is clean; where they overlap the
distribution is really a continuum.

```{r}
#| label: mixture-fig
#| fig-cap: "Two-component Gaussian mixture on log10(concentration), bulk. Manganese, molybdenum and selenium show a distinct enriched population; copper, zinc and cobalt are closer to a single continuum, so their threshold is a soft split."
#| fig-width: 9
#| fig-height: 5.4
cb <- comp |> filter(cat == "bulk", symbol %in% c("CO","CU","MN","MO","SE","ZN")) |>
  mutate(mu_bg = log10(gm_bg), mu_en = log10(gm_en), thr = log10(threshold))
hb <- hist |> mutate(symbol = factor(symbol, levels = elem_levels))

# fitted mixture curves, scaled from density to counts (n * bin width)
curve <- cb |> rowwise() |> do({
  r <- .
  h <- hb[hb$symbol == r$symbol, ]
  bw <- mean(diff(sort(h$bin_mid)))
  x  <- seq(min(h$bin_mid), max(h$bin_mid), length.out = 200)
  scale <- r$n * bw
  data.frame(symbol = r$symbol, x = x,
             bg  = r$lambda_bg * dnorm(x, r$mu_bg, r$sd_bg_log) * scale,
             en  = (1 - r$lambda_bg) * dnorm(x, r$mu_en, r$sd_en_log) * scale) |>
    mutate(tot = bg + en)
}) |> ungroup() |> mutate(symbol = factor(symbol, levels = elem_levels))

ggplot() +
  geom_col(data = hb, aes(bin_mid, count), fill = "grey85", width = 0.06) +
  geom_line(data = curve, aes(x, bg), colour = "#4477AA", linewidth = 0.6) +
  geom_line(data = curve, aes(x, en), colour = "#EE6677", linewidth = 0.6) +
  geom_line(data = curve, aes(x, tot), colour = "black", linewidth = 0.5) +
  geom_vline(data = cb, aes(xintercept = thr), linetype = "dashed", colour = "grey40") +
  facet_wrap(~ symbol, scales = "free") +
  labs(x = "log10 concentration (mg/kg)", y = "count")
```

## The threshold, and does it agree with the offshore background?

The mixture threshold is a background upper bound derived from the distribution alone. The
offshore 90th percentile (from the [Global Background](./background.qmd) page) is one derived from
location. The two use different reasoning, but not different data: the offshore samples are a
large share of the global set the mixture is fitted on, so the two statistics overlap.
Agreement between them is **corroboration**, not independent confirmation.

```{r}
#| label: compare-table
off <- offc |> filter(cat == "bulk") |> select(symbol, offshore_p90 = p90_off10)
comp |> filter(cat == "bulk") |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  left_join(off, by = "symbol") |>
  transmute(Element = symbol,
            `background geom-mean` = gm_bg,
            `mixture threshold` = threshold,
            `offshore P90` = offshore_p90,
            `% background` = pct_bg,
            `two populations?` = ifelse(separated == TRUE, "yes", "continuum")) |>
  arrange(Element) |>
  kable(caption = "Bulk background from the mixture, and the offshore P90 for comparison (mg/kg).")
```

The two methods **line up** where the data is richest: copper's mixture threshold (about 28)
sits near its offshore P90 (about 34), and molybdenum's (about 2.4) near its offshore value
(about 2.9), despite one being a distribution split and the other a spatial filter. That
convergence is worth having, with the shared data above in mind. Manganese and zinc come out
lower on the mixture than offshore.

The **two populations? / continuum** column matters: manganese, molybdenum and selenium
have a genuinely distinct enriched hump, while copper, zinc and cobalt do not, their
distributions are closer to a single skewed population, so their threshold is a soft cut,
not a natural break. A distinct hump is not always contamination either: for manganese it
may be the muddy-versus-sandy grain-size split showing through in the raw value.

::: {.callout-note}
The mixture is unsupervised and knows nothing about pressure, so its "background"
population is only the lower mode of the distribution, which is a good proxy for natural
sediment only when enrichment is limited to a minority of sites. It is fitted on raw
values, so grain size can create or hide a second population; the
[Grain-Size-Normalised](./background-gsnorm.qmd) and [Enrichment Factor](./background-ef.qmd)
pages are the controls. Read as one line of evidence that agrees with the spatial background
for the well-covered metals, remembering that the two share most of their data. Iodine is too
sparse to fit.
:::
