---
title: "Pristine Classification"
subtitle: "Combining the background estimates into a per-sample verdict, and where it cannot be made"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

This page draws the background estimates together into a per-measurement **pristine**
(at or below background) verdict, per element and fraction (a sample can be pristine for
one element and not another). It combines three "at/below background" criteria from the
earlier pages: **EF < 1** (the [Enrichment Factor](./background-ef.qmd), grain-size
controlled), **below the [distribution-mixture threshold](./background-mixture.qmd)**, and
**below the offshore P90** ([Global Background](./background.qmd)).

Every one of these criteria is estimated with **no aquaculture information**, so the
pristine verdict is independent of the pressure it is meant to reveal. The aquaculture
sites enter only afterwards, as an **evaluation**: the closing section checks whether the
pristine fraction falls as sediments approach the cages. That check turns out to be a sanity
test rather than a validation, for reasons set out there.

Only **grain-size-controlled** verdicts are trusted. A raw-concentration version was tried
and rejected: sandy near-cage sediment slips under a concentration threshold regardless of
pressure, and (see below) aluminium is missing exactly where pressure is highest, so a
concentration flag validated **backwards**. Where aluminium (hence EF) is absent, a sample
is therefore left **unclassified** rather than guessed. Two flags are reported:

- **EF rule** (permissive, EFSA-preferred): pristine where EF < 1.
- **Strict rule** (conservative): EF < 1 **plus two raw-concentration screens**, the
  mixture threshold and the offshore P90.

The three criteria are not peers, and the strict rule is not three independent votes. Only
EF < 1 is grain-size controlled; the other two are screens on the raw concentration, the
same kind of test this page has just rejected on its own. ANDing them onto EF < 1 cannot
make a sample falsely pristine, only falsely non-pristine, which is the safe direction for a
conservative rule. It does mean a **muddy but clean** sample can fail the strict rule on
grain size alone. Cobalt below is the clearest case, 56% pristine on the EF rule against 20%
strict: that gap reflects cobalt's unreliable mixture threshold and its raw values, not
evidence that cobalt is enriched.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
summ <- rd("refined_pristine_summary.csv")
cov  <- rd("refined_pristine_coverage.csv")
val  <- rd("refined_pristine_validation.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
BLUE <- "#4477AA"; RED <- "#EE6677"
theme_set(theme_minimal(base_size = 12))
```

## How much is pristine, and how much is classifiable

Among the samples that can be classified (those with aluminium), the share flagged pristine
under each rule, bulk. The **classifiable** column is the catch: for copper and zinc, only
about a quarter of samples carry the aluminium needed to judge them at all.

```{r}
#| label: summary-table
summ |>
  filter(cat == "bulk", reliable == TRUE) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol, `samples` = n, `% classifiable` = pct_classifiable,
            `pristine (EF rule)` = pct_ef, `pristine (strict)` = pct_strict) |>
  arrange(Element) |>
  kable(format.args = list(big.mark = ","))
```

```{r}
#| label: pristine-fig
#| fig-cap: "Share of classifiable bulk samples flagged pristine, EF rule versus strict. Molybdenum is least pristine; the strict rule is much lower for cobalt because its mixture threshold is unreliable (no clear two-population split)."
#| fig-width: 7.5
#| fig-height: 3.4
summ |>
  filter(cat == "bulk", reliable == TRUE) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  select(symbol, EF = pct_ef, strict = pct_strict) |>
  pivot_longer(-symbol, names_to = "rule", values_to = "pct") |>
  ggplot(aes(pct, symbol, fill = rule)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(EF = BLUE, strict = RED), name = NULL) +
  labs(x = "% of classifiable samples pristine", y = NULL)
```

## The data gap: aluminium is missing where the farms are

This is the key result, and a limitation for the aquaculture question. Aluminium coverage
collapses toward the cages: the sediments most in need of a pristine-or-not verdict are the
ones least able to receive one.

```{r}
#| label: coverage-fig
#| fig-cap: "Percentage of samples that are classifiable (carry aluminium) by distance to the nearest farm. Only about 2% of the samples within 1 km of a farm can be judged, against about 90% beyond 20 km."
#| fig-width: 7.5
#| fig-height: 2.8
cov |>
  filter(axis == "distance to aquaculture") |>
  mutate(band = factor(band, levels = c("<1km", "1-5km", "5-20km", ">20km"))) |>
  ggplot(aes(band, pct_classifiable)) +
  geom_col(fill = "#AA3377", width = 0.65) +
  geom_text(aes(label = paste0(pct_classifiable, "%")), vjust = -0.4, size = 3.2) +
  scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.1))) +
  labs(x = "distance to nearest aquaculture farm", y = "% classifiable")
```

That gradient is real, but it is **not a property of near-cage sampling**. Splitting the
same measurements by data source shows coverage that does not depend on distance at all:

| Distance band | Vannmiljø, % with Al | Mareano, % with Al | Vannmiljø share of band |
|---------------|---------------------:|-------------------:|------------------------:|
| < 1 km        | 1.4%                 | 100%               | 99.8%                   |
| 1-5 km        | 3.7%                 | 99.9%              | 93.4%                   |
| 5-20 km       | 1.4%                 | 100%               | 92.8%                   |
| > 20 km       | 3.1%                 | 99.6%              | 8.3%                    |

Within Vannmiljø, aluminium is reported for 1 to 4% of measurements at **every** distance,
including beyond 20 km. Within Mareano it is reported for essentially all of them, including
under 1 km. What changes across the bands is not sampling practice near cages, it is
**which programme sampled there**: the < 1 km band is 99.8% Vannmiljø, the > 20 km band is
85% Mareano.

So the finding is sharper than "near-cage sediment cannot be assessed". It is that
**Vannmiljø does not report aluminium, and Vannmiljø is the programme that samples around
the farms.** That makes the recommendation concrete rather than general: aluminium (or
another grain-size normaliser) reported alongside the trace elements in Norwegian coastal
monitoring would close most of this gap on its own.

## Validation: what the distance trend can and cannot show

On the samples that *can* be classified, the pristine share drops as sites approach the
coast and the farms, which is the direction a verdict tracking real pressure should move.

The section above makes this a weak test rather than a confirmation, though. The classifiable
samples are not the same population across the bands: within 1 km of a farm they are roughly
350 Vannmiljø and 46 Mareano measurements, while beyond 20 km they are roughly 12,600 Mareano
and 38 Vannmiljø. The distance axis is very nearly a source axis, so part of any trend along
it is a difference between monitoring programmes and the sea areas they cover. Read the
figure as consistency, not as validation: a real test has to be run **within a single
source**, and has not yet been done. Near-cage points also rest on few classifiable samples,
so read the trend, not the wiggle.

```{r}
#| label: validation-fig
#| fig-cap: "Percentage pristine among classifiable samples, by distance to aquaculture and to coast (EF rule). Pristine rises with distance from pressure, but the classifiable samples in the near and far bands come from different monitoring programmes, so the comparison is confounded with source."
#| fig-width: 8.5
#| fig-height: 3.4
val |>
  filter(rule == "ef") |>
  mutate(band = factor(band, levels = c("<1km","1-5km","5-20km",">20km","1-10km","10-50km",">50km"))) |>
  ggplot(aes(band, pct_pristine, group = axis, colour = axis)) +
  geom_line(linewidth = 0.7) + geom_point(size = 2) +
  scale_colour_manual(values = c("distance to aquaculture" = "#AA3377",
                                 "distance to coast" = BLUE), name = NULL) +
  facet_wrap(~ axis, scales = "free_x") +
  labs(x = NULL, y = "% pristine (classifiable)") +
  theme(legend.position = "none")
```

::: {.callout-note}
"Pristine" here means at or below the local background by a grain-size-controlled measure,
per element, not an absolute or a site-level judgement. The strict rule inherits the
distribution-mixture caveat (unreliable where there is no clear two-population split, as for
cobalt). The honest headline is the coverage gap: the classification is trustworthy where
aluminium exists, which is most of the open-shelf data but almost none of the near-cage data,
so the aquaculture sites remain largely unassessable from these measurements alone.

Three further limitations belong on the record. Norwegian farms sit in sheltered,
organic-rich, periodically anoxic fjords, so part of any near-cage signal for the
redox-sensitive elements may be fjord geochemistry that happens to be co-located with farms
rather than farm input; separating the two needs a matched comparison of farmed and unfarmed
fjords. Distance to the nearest farm is also not the same thing as the pressure a sediment
experienced, because the aquaculture reference includes closed sites and the sample date is
not aligned to any operating period. And the [Enrichment Map](./enrichment-map.qmd) applies
these thresholds to per-site averages, which makes it a **different classifier** from the
per-measurement one tabulated here, not just a coarser drawing of it. This is a
synthesis of the [Global Background](./background.qmd),
[Enrichment Factor](./background-ef.qmd) and
[Distribution-Mixture](./background-mixture.qmd) pages; iodine is too sparse throughout.
:::
