---
title: "Pristine Classification"
subtitle: "Combining the background estimates into a per-sample verdict, and where it cannot be made"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

This page draws the background estimates together into a per-measurement **pristine**
(at or below background) verdict, per element and fraction (a sample can be pristine for
one element and not another). It combines three "at/below background" criteria from the
earlier pages: **EF < 1** (the [Enrichment Factor](./background-ef.qmd), grain-size
controlled), **below the [distribution-mixture threshold](./background-mixture.qmd)**, and
**below the offshore P90** ([Global Background](./background.qmd)).

Every one of these criteria is estimated with **no aquaculture information**, so the
pristine verdict is independent of the pressure it is meant to reveal. The aquaculture
sites enter only afterwards, as an **evaluation**: the closing section checks whether the
pristine fraction falls as sediments approach the cages. That check turns out to be a sanity
test rather than a validation, for reasons set out there.

Only **grain-size-controlled** verdicts are trusted. A raw-concentration version was tried
and rejected: sandy near-cage sediment slips under a concentration threshold regardless of
pressure, and (see below) aluminium is missing exactly where pressure is highest, so a
concentration flag validated **backwards**. Where aluminium (hence EF) is absent, a sample
is therefore left **unclassified** rather than guessed. Two flags are reported:

- **EF rule** (permissive, EFSA-preferred): pristine where EF < 1.
- **Strict rule** (conservative): EF < 1 **plus two raw-concentration screens**, the
  mixture threshold and the offshore P90.

The three criteria are not peers, and the strict rule is not three independent votes. Only
EF < 1 is grain-size controlled; the other two are screens on the raw concentration, the
same kind of test this page has just rejected on its own. ANDing them onto EF < 1 cannot
make a sample falsely pristine, only falsely non-pristine, which is the safe direction for a
conservative rule. It does mean a **muddy but clean** sample can fail the strict rule on
grain size alone. Cobalt used to be the clearest case, 56% pristine on the EF rule against
20% strict, which reflected its unreliable mixture threshold rather than any enrichment.
That threshold is now recognised as unusable and dropped from the rule, so the two rules
have converged for cobalt; the mechanism still applies wherever a usable threshold sits low
against muddy sediment.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
summ <- rd("refined_pristine_summary.csv")
cov  <- rd("refined_pristine_coverage.csv")
val  <- rd("refined_pristine_validation.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
BLUE <- "#4477AA"; RED <- "#EE6677"
theme_set(theme_minimal(base_size = 12))
```

## How much is pristine, and how much is classifiable

Among the samples that can be classified (those with aluminium), the share flagged pristine
under each rule, bulk. Both rules rest on EF < 1, whose reference is now computed within a
single aluminium measurement basis (see the [Enrichment Factor](./background-ef.qmd) page).
That is why the **classifiable** column is low: a sample off its fraction's basis is
withdrawn rather than misjudged. **Selenium and molybdenum are absent** because their
verdicts are withheld entirely, most of their measurements having fallen below the limit of
quantification and been removed upstream ([Method Revisions](./method-revisions.qmd)). The **classifiable** column is the catch: for copper and zinc, only
about a quarter of samples carry the aluminium needed to judge them at all.

```{r}
#| label: summary-table
summ |>
  filter(cat == "bulk", reliable == TRUE) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol, `samples` = n, `% classifiable` = pct_classifiable,
            `pristine (EF rule)` = pct_ef, `pristine (strict)` = pct_strict) |>
  arrange(Element) |>
  kable(format.args = list(big.mark = ","))
```

```{r}
#| label: pristine-fig
#| fig-cap: "Share of classifiable bulk samples flagged pristine, EF rule versus strict. Molybdenum is least pristine; the strict rule is much lower for cobalt because its mixture threshold is unreliable (no clear two-population split)."
#| fig-width: 7.5
#| fig-height: 3.4
summ |>
  filter(cat == "bulk", reliable == TRUE) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  select(symbol, EF = pct_ef, strict = pct_strict) |>
  pivot_longer(-symbol, names_to = "rule", values_to = "pct") |>
  ggplot(aes(pct, symbol, fill = rule)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(EF = BLUE, strict = RED), name = NULL) +
  labs(x = "% of classifiable samples pristine", y = NULL)
```

## The data gap: aluminium is missing where the farms are

This is the key result, and a limitation for the aquaculture question. Aluminium coverage
collapses toward the cages: the sediments most in need of a pristine-or-not verdict are the
ones least able to receive one.

```{r}
#| label: coverage-fig
#| fig-cap: "Percentage of samples that are classifiable (carry aluminium) by distance to the nearest farm. Only about 2% of the samples within 1 km of a farm can be judged, against about 90% beyond 20 km."
#| fig-width: 7.5
#| fig-height: 2.8
cov |>
  filter(axis == "distance to aquaculture") |>
  mutate(band = factor(band, levels = c("<1km", "1-5km", "5-20km", ">20km"))) |>
  ggplot(aes(band, pct_classifiable)) +
  geom_col(fill = "#AA3377", width = 0.65) +
  geom_text(aes(label = paste0(pct_classifiable, "%")), vjust = -0.4, size = 3.2) +
  scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.1))) +
  labs(x = "distance to nearest aquaculture farm", y = "% classifiable")
```

That gradient is real, but it is **not a property of near-cage sampling**. Splitting the
same measurements by data source shows coverage that does not depend on distance at all:

| Distance band | Vannmiljø, % with Al | Mareano, % with Al | Vannmiljø share of band |
|---------------|---------------------:|-------------------:|------------------------:|
| < 1 km        | 1.4%                 | 100%               | 99.8%                   |
| 1-5 km        | 3.7%                 | 99.9%              | 93.4%                   |
| 5-20 km       | 1.4%                 | 100%               | 92.8%                   |
| > 20 km       | 3.1%                 | 99.6%              | 8.3%                    |

Within Vannmiljø, aluminium is reported for 1 to 4% of measurements at **every** distance,
including beyond 20 km. Within Mareano it is reported for essentially all of them, including
under 1 km. What changes across the bands is not sampling practice near cages, it is
**which programme sampled there**: the < 1 km band is 99.8% Vannmiljø, the > 20 km band is
85% Mareano.

So the finding is sharper than "near-cage sediment cannot be assessed". It is that
**Vannmiljø does not report aluminium, and Vannmiljø is the programme that samples around
the farms.** That makes the recommendation concrete rather than general: aluminium (or
another grain-size normaliser) reported alongside the trace elements in Norwegian coastal
monitoring would close most of this gap on its own.

## Validation: what the distance trend can and cannot show

On the samples that *can* be classified, the pristine share drops as sites approach the
coast and the farms, which is the direction a verdict tracking real pressure should move.

The bands are not sampled by the same programmes, so this could have been a difference
between monitoring programmes wearing a distance label. Restricting the reference to one
aluminium basis made them homogeneous enough to test that directly, by repeating the
comparison **within a single source**, and it holds in both sources that span more than one
band:

```{r}
#| label: within-source-table
vsrc <- rd("refined_pristine_validation_source.csv")
vsrc |>
  mutate(Band = factor(band, levels = c("<1km", "1-5km", "5-20km", ">20km"))) |>
  transmute(Source = source, Band, samples = n,
            `% pristine (EF rule)` = pct_ef, `% pristine (strict)` = pct_strict) |>
  arrange(Source, Band) |>
  kable(format.args = list(big.mark = ","))
```

Mareano rises from 12% pristine under 1 km to 49% beyond 20 km, and Vannmiljø from 16% to
29% across the three bands it spans. Both rise monotonically, so **the gradient is not an
artefact of which programme sampled where**. Near-cage points still rest on few classifiable
samples, so read the trend, not the wiggle.

```{r}
#| label: validation-fig
#| fig-cap: "Percentage pristine among classifiable samples, by distance to aquaculture and to coast (EF rule). Pristine rises with distance from pressure; the table above shows the same rise holds within each source separately."
#| fig-width: 8.5
#| fig-height: 3.4
val |>
  filter(rule == "ef") |>
  mutate(band = factor(band, levels = c("<1km","1-5km","5-20km",">20km","1-10km","10-50km",">50km"))) |>
  ggplot(aes(band, pct_pristine, group = axis, colour = axis)) +
  geom_line(linewidth = 0.7) + geom_point(size = 2) +
  scale_colour_manual(values = c("distance to aquaculture" = "#AA3377",
                                 "distance to coast" = BLUE), name = NULL) +
  facet_wrap(~ axis, scales = "free_x") +
  labs(x = NULL, y = "% pristine (classifiable)") +
  theme(legend.position = "none")
```

::: {.callout-note}
"Pristine" here means at or below the local background by a grain-size-controlled measure,
per element, not an absolute or a site-level judgement. The strict rule no longer inherits
an unreliable mixture threshold: where the distribution does not support one, that criterion
is dropped rather than applied (see [Method Revisions](./method-revisions.qmd)). The honest headline is the coverage gap: the classification is trustworthy where
aluminium exists, which is most of the open-shelf data but almost none of the near-cage data,
so the aquaculture sites remain largely unassessable from these measurements alone.

Three further limitations belong on the record. Norwegian farms sit in sheltered,
organic-rich, periodically anoxic fjords, so part of any near-cage signal for the
redox-sensitive elements may be fjord geochemistry that happens to be co-located with farms
rather than farm input; separating the two needs a matched comparison of farmed and unfarmed
fjords. Distance to the nearest farm is also not the same thing as the pressure a sediment
experienced, because the aquaculture reference includes closed sites and the sample date is
not aligned to any operating period. And the [Enrichment Map](./enrichment-map.qmd) applies
these thresholds to per-site averages, which makes it a **different classifier** from the
per-measurement one tabulated here, not just a coarser drawing of it. This is a
synthesis of the [Global Background](./background.qmd),
[Enrichment Factor](./background-ef.qmd) and
[Distribution-Mixture](./background-mixture.qmd) pages; iodine is too sparse throughout.
:::
