---
title: "Pressure-Based Background"
subtitle: "Background from distance to a fish farm, and what the size of that farm adds"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Instead of a spatial proxy for pristine, this page uses a direct **pressure gradient**:
distance to the nearest **fish farm**. Sites far from any farm are the low-pressure
**background**; sites near the cages show the **enrichment**. This is the aquaculture
question asked directly, and it is **Norway only** (the aquaculture reference is
Norwegian).

For each element and fraction, the value is summarised in distance bands (< 1 km, 1-5 km,
5-20 km, > 20 km). The **> 20 km** band is read as the background; the ratio of the
nearest band to it (the 90th percentiles) is the near-cage enrichment.

Because it is built from the farm positions, this pressure gradient is held out of the
background and pristine estimates themselves. Its role is to **evaluate** them: the
[Enrichment Factor](./background-ef.qmd) and [Pristine
Classification](./background-pristine.qmd) pages are estimated without aquaculture
information, and this page is the yardstick that the enrichment they report matches the
real pressure.

::: {.callout-important
title="This page was rebuilt in August 2026 and its numbers moved"}
It used to measure distance to the nearest **aquaculture site of any kind**, and a
land-based smolt plant and a mussel raft are both aquaculture. Two corrections follow,
and they push in opposite directions: the near band was too generous (it counted samples
next to facilities that put nothing on the seabed) and the far band was not a background
at all (over half of it was contaminated-seabed monitoring). The old copper figure of
about 3.2x and molybdenum figure of about 13x came from that pairing. Both sections
below show the change rather than quietly replacing it.
:::

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
pp   <- rd("refined_pressure_percentiles.csv")
pc   <- rd("refined_pressure_compare.csv")
ax   <- rd("refined_pressure_axis.csv")
axd  <- rd("refined_pressure_axis_dropped.csv")
sz   <- rd("refined_pressure_size.csv")
szc  <- rd("refined_pressure_size_covariates.csv")
szd  <- rd("refined_pressure_size_depth.csv")
szdc <- rd("refined_pressure_size_depth_cost.csv")
szs  <- rd("refined_pressure_size_strata.csv")
szp  <- rd("refined_pressure_size_paired.csv")
szg  <- rd("refined_pressure_size_geography.csv")
st   <- rd("refined_pressure_stated.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
bin_levels  <- c("<1km", "1-5km", "5-20km", ">20km")
band_levels <- c("small", "medium", "large")
pp$symbol   <- factor(pp$symbol, levels = elem_levels)
pp$dist_bin <- factor(pp$dist_bin, levels = bin_levels)
theme_set(theme_minimal(base_size = 12))
```

## The axis: a fish farm, not aquaculture

A fish farm is the pressure this work is about, because the mechanisms are feed
(selenium, zinc, and the organic enrichment that drives the redox chemistry) and the
copper in net antifouling. A mussel raft has none of them, and a land-based smolt plant
puts its effluent through a pipe rather than onto the seabed under the licence
coordinate. Counting either as "near a farm" fills the near band with samples that carry
none of the pressure the band exists to isolate.

The refined `site` table carries both distances, so the change can be shown rather than
asserted. What the old axis added to the near band, by the kind of facility the sample
was actually next to:

```{r}
#| label: dropped-table
axd |>
  transmute(Placement = placement, `Licensed in` = capacity_unit,
            Measurements = n, `% of dropped` = pct) |>
  kable(format.args = list(big.mark = ","))
```

```{r}
#| label: dropped-note
#| echo: false
#| results: asis
land <- round(sum(axd$pct[axd$placement == "land"]), 1)
sea_other <- round(sum(axd$pct[axd$placement != "land"]), 1)
cat(sprintf(
  "So **%s%%** of what the old near band gained sat next to a *land-based* facility, and %s%% next to something at sea that is not a fish farm (mostly blue mussel, licensed by area in decare). None of them is a sea cage.\n",
  land, sea_other))
```

The effect on the two elements with enough data to see it:

```{r}
#| label: axis-table
ax |>
  filter(cat == "bulk", !is.na(enrich_near_farm), n_near_farm >= 100) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol,
            `n near (farm)` = n_near_farm,
            `n added by old axis` = n_near_gain,
            `old ratio` = enrich_near_aqua,
            `new ratio, raw far band` = enrich_near_farm) |>
  arrange(Element) |>
  kable(format.args = list(big.mark = ","))
```

The new raw ratios are *below* 1, which is the second problem rather than a finding about
farms.

## The background had to be cleaned too

The far band is only a background if what sits in it is unpressured, and it is not.
Vannmiljø files why each sample was taken, and the > 20 km pool turns out to be more than
half **stated pressure**: the contaminated-seabed, industry and sewage programmes, which
is urban fjord sediment carrying the very metals being measured.

```{r}
#| label: farband-table
pc |>
  filter(cat == "bulk", !is.na(enrich_near)) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol,
            `n far` = `n_>20km`, `P90 far, raw` = `>20km`,
            `% stated pressure` = pct_far_stated,
            `n far, cleaned` = n_far_clean,
            `P90 far, cleaned` = p90_far_clean) |>
  arrange(Element) |>
  kable(format.args = list(big.mark = ","))
```

Removing the stated-pressure rows takes the copper background from 102 to 25.7 mg/kg and
the zinc background from 273 to 126.5. Only rows a provider has **stated** are pressure
monitoring are removed; the four sources that record no programme keep every row, so this
removes stated pressure and never inferred pressure.

**Every ratio below is taken against the cleaned far band.**

## Concentration versus distance to a fish farm

The 90th percentile by distance band, bulk (reliable groups only). A value that falls as
distance grows is a pressure signal; the far band is the background.

```{r}
#| label: gradient-fig
#| fig-cap: "Bulk 90th-percentile concentration by distance to the nearest fish farm (log scale). Only cobalt falls monotonically with distance. Copper, zinc and molybdenum peak in the 1-20 km bands rather than under the cages, and manganese is diagenetic rather than a pressure signal."
#| fig-width: 8.5
#| fig-height: 4.4
pp |>
  filter(cat == "bulk", reliable, symbol %in% c("CO","CU","MN","MO","ZN")) |>
  ggplot(aes(dist_bin, p90, group = symbol, colour = symbol)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  labs(x = "distance to nearest fish farm", y = "P90 concentration, mg/kg (log)")
```

## Near-cage enrichment

The nearest-band (< 1 km) 90th percentile over the cleaned far background (> 20 km).
Groups below 30 measurements in the near band are marked rather than dropped, because
which elements *lack* near-cage data is itself the finding:

```{r}
#| label: enrich-table
pc |>
  filter(cat == "bulk", !is.na(enrich_near_clean)) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol,
            `n < 1 km` = `n_<1km`,
            `P90 < 1 km` = `<1km`,
            `P90 far (cleaned)` = p90_far_clean,
            `enrichment` = enrich_near_clean,
            Reading = if_else(`n_<1km` >= 30, "", "too few to read")) |>
  arrange(desc(enrichment)) |>
  kable()
```

What survives:

- **Copper, about 2.6x**, on 11,475 measurements. Antifouling treatment of nets and
  cages. This is the one strong, well-supported near-cage signal on the page.
- **Cobalt, about 1.5x**, but on 33 measurements, which is barely over the threshold.
- **Zinc, about 1.1x**, on 10,277 measurements. Feed and galvanised structures, and on
  this axis the effect is small.
- **Manganese, 0.35**, near well below far. It is diagenetic, not a pressure signal, the
  same behaviour it showed offshore.
- **Molybdenum and selenium have almost no near-cage data at all**: 21 and 1 bulk
  measurements within 1 km of a sea cage. The old page reported them as the strongest
  signals on the page (about 13x and about 5x). Those figures came from the near band
  when it still included land-based facilities. They are withdrawn, not restated. Both
  elements are also under the [LOQ censoring
  decision](./background.qmd), so they carry no verdict anywhere in this project.

The gradient is not a dose-response curve. Copper, zinc and molybdenum all peak in the
1-20 km bands rather than under the cages, which is what the next section is for.

## Does the size of the farm change the signal

Norwegian finfish licences are issued in MTB (maximum permitted standing biomass), and
the standard concession is 780 t, so the refined `site` table carries the nearest farm's
licensed capacity and a size band: **small** (up to 2 concessions), **medium** (up to 4),
**large** (above 4). A 780 t farm and a 19,000 t farm at the same distance are not the
same pressure.

This is a better test than near against far, because both sides sit under a cage in a
fjord: the setting is roughly held constant and only the licensed capacity moves.

```{r}
#| label: size-table
sz |>
  filter(cat == "bulk", reliable) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         band = factor(band, levels = band_levels)) |>
  transmute(Element = symbol, `Size band` = band,
            `Median MTB, t` = mtb_median, n = n,
            `P90 < 1 km` = p90, `P90 far (cleaned)` = p90_far_clean,
            enrichment = enrich_clean) |>
  arrange(Element, `Size band`) |>
  kable(format.args = list(big.mark = ","))
```

```{r}
#| label: size-fig
#| fig-cap: "Near-cage enrichment by the licensed size of the nearest farm, bulk, cleaned far background. Both elements peak at medium and fall at large, which is the opposite of a dose-response. The section below shows that this ordering is geography rather than size."
#| fig-width: 7
#| fig-height: 3.6
sz |>
  filter(cat == "bulk", reliable) |>
  mutate(band = factor(band, levels = band_levels)) |>
  ggplot(aes(band, enrich_clean, group = symbol, colour = symbol)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey55") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.6) +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  labs(x = "licensed size of the nearest fish farm",
       y = "near/far enrichment (P90)")
```

Only copper and zinc have enough near-cage data for the split; cobalt, manganese and
molybdenum fall under 30 measurements in every band. Both behave the same way: small and
medium are close, and **large is the lowest of the three**. Copper goes 2.91, 3.15, 2.13
and zinc 1.11, 1.26, 0.95.

That is the opposite of a dose-response. The obvious suspect is siting rather than
chemistry, because the size bands differ in more than size:

```{r}
#| label: size-cov-table
szc |>
  mutate(band = factor(band, levels = band_levels)) |>
  transmute(`Size band` = band, Sites = n_sites, `Median MTB, t` = mtb_median,
            `Median depth, m` = median_depth_m,
            `Median distance to coast, km` = median_dist_coast_km,
            `Median mud, %` = median_fines_pct,
            `% with a mud measurement` = pct_with_fines) |>
  arrange(`Size band`) |>
  kable(format.args = list(big.mark = ","))
```

Large and medium farms sit in roughly **twice the water depth** of small ones (107 m and
121 m against 66 m) and further from shore. That gives a ready explanation: Norway
licenses the big farms into deep, well-flushed water precisely because it disperses the
load, so a bigger farm depositing more organic matter over a larger and better-flushed
footprint could leave *less* under the licence coordinate. A flat or inverted size
response is therefore not by itself evidence that the pressure is absent, and this table
is what the one above has to be read against.

It is an explanation, though, not a finding, and it is the one tested below. It fails.

The honest reading of the table above on its own: **size does not give a clean
dose-response, and there is a confounder big enough to explain why.** The section that
follows is the test of that explanation, and it does not survive it.

## Separating size from where the farm was sited

The claim to check is that large farms read cleanest because Norway licenses them into
deep, well-flushed water. It is a testable claim: hold depth constant and the inversion
should go away. It does not. Something else turns out to be doing the work.

The comparison is built as a ladder, because each rung removes a different thing and the
answer changes twice on the way up.

| Rung | What it removes |
|---|---|
| `raw` | nothing: the near band exactly as the table above reports it |
| `clean` | rows Vannmiljø filed as `pressure` monitoring |
| `depth` | every band cut to a common 50-150 m window |
| `depth+texture` | a 20-80% mud window on top |

### Cleaning first, because the bands are not charged equally

Stated-pressure rows are not spread evenly across the size bands. The small-farm near
band is **13.8%** contaminated-seabed, industry and sewage monitoring; the large-farm
near band is **1.3%**. Raw, the small band is being charged for harbour sediment that
happens to sit within a kilometre of a small farm.

```{r}
#| label: size-cost-table
szdc |>
  mutate(band = factor(band, levels = band_levels)) |>
  transmute(`Size band` = band, `Rows, raw` = n_raw,
            `% stated pressure` = pct_stated,
            `Median depth, m` = depth_p50_raw,
            `Rows, cleaned` = n_clean, `Rows, depth-matched` = n_depth,
            `% kept` = pct_kept_depth,
            `Rows, + texture` = n_texture) |>
  arrange(`Size band`) |>
  kable(format.args = list(big.mark = ","))
```

### The ladder

```{r}
#| label: size-ladder-table
szd |>
  filter(cat == "bulk", symbol %in% c("CU", "ZN"), reliable) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         band = factor(band, levels = band_levels),
         tier = factor(tier, levels = c("raw", "clean", "depth",
                                        "depth+texture"))) |>
  transmute(Control = tier, Element = symbol, `Size band` = band, n = n,
            `Median depth, m` = depth_p50, `Median mud, %` = fines_p50,
            `P90 < 1 km` = p90, enrichment = enrich_clean) |>
  arrange(Control, Element, `Size band`) |>
  kable(format.args = list(big.mark = ","))
```

Matching depth works, in the sense that the three bands come out at a median of 84, 97
and 95 m instead of 66, 123 and 110. It changes nothing about the ordering. Copper's
near-cage P90 reads 65.1, 96.4 and 50.7 for small, medium and large, so **large is still
the lowest of the three at the same depth**, and zinc behaves the same way. Adding the
texture window pushes the same direction again, on numbers too small to lean on: the
small band falls under 30 measurements and drops out of the table entirely.

### It is not the window either

A single window can always be accused of flattering one band. The same contrast read
inside coarse depth strata gives the same answer at every depth:

```{r}
#| label: size-strata-fig
#| fig-cap: "Bulk median concentration under the cages by depth stratum and licensed farm size, stated-pressure rows removed. Concentration climbs steeply with depth in every band (the muddier deep seabed holds more metal), and the large band sits at or below the others in every stratum."
#| fig-width: 8.5
#| fig-height: 4
szs |>
  filter(cat == "bulk", symbol %in% c("CU", "ZN"), reliable) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         band = factor(band, levels = band_levels),
         depth_band = factor(depth_band, levels = c("<50m", "50-100m",
                                                    "100-150m", "150-250m",
                                                    ">250m"))) |>
  ggplot(aes(depth_band, p50, group = band, colour = band)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.4) +
  facet_wrap(~ symbol, scales = "free_y") +
  scale_colour_brewer(palette = "Set2", name = "farm size") +
  labs(x = "water depth at the sample", y = "median concentration, mg/kg")
```

Two things are visible. Depth itself is a strong driver: copper roughly doubles and zinc
roughly triples from the shallowest stratum to the deepest, which is the ordinary
grain-size effect (deeper seabed here is muddier, and mud holds metal). And within every
stratum the large band sits at or below the other two. **Depth is a real covariate and
matching it was worth doing, but it is not the explanation for the size inversion.**

### What is doing the work: the coast the farm is on

The size bands are not samples of the same coastline.

```{r}
#| label: size-geo-table
szg |>
  filter(pct_of_band >= 1) |>
  mutate(band = factor(band, levels = band_levels)) |>
  transmute(`Size band` = band, Sea = sea_name, Sites = n,
            `% of band` = pct_of_band) |>
  arrange(`Size band`, desc(Sites)) |>
  kable(format.args = list(big.mark = ","))
```

Large farms are overwhelmingly Norwegian Sea, and they are the only band with Barents Sea
sites at all; small farms carry more than twice the North Sea share. Those coasts do not
have the same sediment or the same industrial history, so a pooled comparison of size
bands is partly a comparison of regions.

### Pairing farms inside one municipality

The fix is to never compare across regions: take one median per municipality and size
band, form the ratio between two size bands **in the same municipality**, and summarise
those ratios. The regional background then cancels by construction. `% above 1` is the
sign test that goes with the ratio: near half is a null however the ratio itself reads.

```{r}
#| label: size-paired-table
szp |>
  filter(cat == "bulk", symbol %in% c("CU", "ZN")) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         config = factor(config, levels = c("50-150m", "30-200m", "any depth"))) |>
  transmute(`Depth window` = config, Element = symbol, Contrast = contrast,
            Municipalities = n_muni, `Ratio P25` = ratio_p25,
            `Ratio median` = ratio_p50, `Ratio P75` = ratio_p75,
            `% above 1` = pct_above_1) |>
  arrange(`Depth window`, Element, Contrast) |>
  kable(format.args = list(big.mark = ","))
```

The inversion is gone. Every contrast lands near 1 (the eighteen ratios span 0.83 to
1.28), the direction flips between depth windows, and the share of municipalities falling
each way hovers around half in almost every row. Copper says large farms are *higher*
than small ones and *lower* than medium ones in the same breath, which is what a null
looks like when it is forced to pick a sign.

### What this does and does not say

**It says the proxy fails.** Licensed MTB is not a usable stand-in for seabed metal load
at the licence coordinate. The pooled size response reported above is mostly geography,
some of it is stated-pressure monitoring landing unevenly on the small band, and none of
it survives comparing farms to their own neighbours. Modelling pressure as
size-weighted distance, which was the reason for carrying capacity onto `site` in the
first place, is not supported by this data.

**It does not say anything about the near-cage signal itself.** The near/far enrichment
earlier on this page is untouched: it compares distance bands, not size bands. Copper is
still about 2.5x under the cages against the cleaned far background, and that number
never depended on farm size.

Three reasons the proxy might fail even if capacity does drive deposition, none of which
this data can separate: MTB is a licensed **ceiling** rather than the stock in the water
when the sediment was sampled; a bigger farm spreads a bigger footprint, so the licence
coordinate is not the same place relative to the cages for a 780 t farm and a 19,000 t
one; and siting, feed conversion and fallowing practice all improve with operator scale.
The honest summary is that **capacity does not predict what is under the cages**, not
that capacity does not matter.

::: {.callout-note title="The depth match was the test that was asked for, and it came back negative"}
This page previously closed by saying that separating size from siting needed a
depth-matched comparison, which was not built. It is built now, and depth turned out not
to be the confounder that mattered: the inversion survives depth matching and survives
depth stratification. Location was the confounder, and pairing farms within a
municipality removes the size response entirely.
:::

## The provider's own label agrees with the geometry

Vannmiljø records why a sample was taken, independently of any coordinate arithmetic. If
the distance axis is measuring what it claims to, the rows the provider filed as
aquaculture monitoring should sit near a farm, and the ones filed as reference conditions
should not:

```{r}
#| label: stated-table
st |>
  filter(pressure_class %in% c("aquaculture", "reference", "pressure")) |>
  mutate(dist_bin = factor(dist_bin, levels = bin_levels)) |>
  transmute(`Stated purpose` = pressure_class, `Distance band` = dist_bin,
            Measurements = n, `% of that purpose` = pct_of_class) |>
  arrange(`Stated purpose`, `Distance band`) |>
  kable(format.args = list(big.mark = ","))
```

**80% of the aquaculture-monitoring measurements fall within 1 km of a fish farm and
95.7% within 5 km, with 0.3% beyond 20 km. Every one of the 64 reference-condition
measurements is beyond 20 km.** Two independent things, a Norwegian programme code and a
great-circle distance to a licence coordinate, agree about which samples are near a farm.
That is the strongest evidence on this page that the axis is sound.

The same table shows the far-band problem from the other side: 40.9% of the stated
**pressure** measurements also sit beyond 20 km, which is what the cleaned background
removes.

## What this page still does not fix

Three things, in descending order of how much they matter.

1. **Time.** Every sample is keyed to its nearest farm whether or not that farm existed
   when the sediment was sampled. The aquaculture reference carries licence start and end
   years, so this is measurable, and [Pressure Controls](./pressure-controls.qmd) measures
   it on the same axis and the same cleaned background as this page. It barely bites:
   87.8% of near-cage copper and 86.9% of near-cage zinc are contemporary with an
   operating farm, and restricting to those moves copper from 2.62 to 2.66.
2. **Geography.** The near band is fjord and the far band is largely open shelf (a median
   72 km from the coast against 400 m), so the comparison is two settings as much as two
   pressures. This is the control that bites: matching within municipality takes copper
   from 2.62 to 1.60 and zinc from 1.09 to 1.04. Copper keeps a clear excess and zinc does
   not. See [Pressure Controls](./pressure-controls.qmd).
3. **Texture.** The near band is a median 32-41% mud depending on size band, and only
   12-16% of near-cage subsamples have a mud measurement at all. The
   [Grain-Size-Normalised](./background-gsnorm.qmd) page is the texture cross-check.

::: {.callout-note}
This is Norway only, and "distance to a farm" remains a coarse pressure measure (a distant
but industrial site is not pristine, and a near site in a well-flushed location may be
clean). The > 20 km cleaned band is a practical low-pressure background, not a guaranteed
pristine one. Read with the raw [Global Background](./background.qmd), the
[Grain-Size-Normalised](./background-gsnorm.qmd) page and
[Geo-Accumulation Index](./background-igeo.qmd) to triangulate.
:::
