---
title: "Geo-Accumulation Index"
subtitle: "The index that reaches the samples the enrichment factor cannot"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The [Enrichment Factor](./background-ef.qmd) page classifies a sample by dividing by
aluminium. This one classifies it by dividing by the background concentration of the same
element:

$$I_\text{geo} = \log_2\!\left(\frac{C}{1.5 \times B}\right)$$

The difference is not cosmetic. EF needs a **normaliser**, Igeo needs a **background**. A
sample can only get an EF if it also carries aluminium, on the right measurement basis, for
an element aluminium actually predicts. It gets an Igeo if it carries a value at all.

That gap is the whole reason this page exists. EFSA asked for aquaculture monitoring data by
name, and it is the one request this pipeline answered with silence: Vannmiljo's MOMC
programme carries aluminium on **5 of 13,996** subsamples, so almost nothing near a fish
farm gets an enrichment factor. EFSA also names the way out, in the same paragraph that sets
the EF preference: the geo-accumulation index and the pollution load index "can be considered
as well in absence of EF".

::: {.callout-important}
## This page issues no verdict, on purpose

Igeo is reported **alongside** the pristine verdicts, not folded into them. The verdicts stay
on EF.

Reach is the argument for promoting it. The texture confounding measured further down is the
argument against, and in bulk it is strong enough (cobalt correlates with the mud fraction at
rho 0.70) that a verdict built on Igeo would be partly a verdict about grain size.

Reporting the index without promoting it keeps the coverage visible to EFSA, who asked for
Igeo precisely where EF is absent, without quietly changing what "pristine" means here.
:::

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
igb <- rd("refined_igeo_background.csv")
igd <- rd("refined_igeo_dist.csv")
igc <- rd("refined_igeo_classes.csv")
igv <- rd("refined_igeo_coverage.csv")
igp <- rd("refined_igeo_pressure.csv")
igf <- rd("refined_igeo_confound.csv")
igt <- rd("refined_igeo_toc_normaliser.csv")

elem_levels  <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
bin_levels   <- c("<1km", "1-5km", "5-20km", ">20km")
class_levels <- c("0 unpolluted", "1 unpolluted-moderate", "2 moderate",
                  "3 moderate-heavy", "4 heavy", "5 heavy-extreme", "6 extreme")
class_cols <- c("0 unpolluted"          = "#228833",
                "1 unpolluted-moderate" = "#66CCEE",
                "2 moderate"            = "#CCBB44",
                "3 moderate-heavy"      = "#EE7733",
                "4 heavy"               = "#EE6677",
                "5 heavy-extreme"       = "#AA3377",
                "6 extreme"             = "#882255")
theme_set(theme_minimal(base_size = 12))
```

## What it buys

The point of the index, in one table: the share of rows that get a verdict under each
instrument, overall and split by fraction and by distance to the nearest aquaculture site.

```{r}
#| label: coverage-table
igv |>
  transmute(Axis = axis, Band = band, n,
            `n EF` = n_ef, `% EF` = pct_ef,
            `n Igeo` = n_igeo, `% Igeo` = pct_igeo) |>
  kable(format.args = list(big.mark = ","))
```

Overall coverage goes from **9.8% to 97.2%**. The row that matters for EFSA's actual request
is the near-farm one: within 1 km of a fish farm, EF classifies **0.4%** of measurements and
Igeo classifies **99.4%**. The relationship inverts with distance, which is the same fact
seen from the other end: aluminium is measured in survey work far offshore and is largely
absent from the monitoring programmes that sample beneath the cages.

The sieved fractions gain almost as much. EF classifies **none** of them, because aluminium
fails the normalisability test in every sieved group; Igeo classifies 93.8 and 97.8%.

## The reference

B is the offshore (more than 10 km from the coast) **median** of the raw concentration, per
element and fraction. It is the same population the EF reference is cut from, so the two
indices disagree about method rather than about which samples count as background.

```{r}
#| label: background-table
withheld_el <- c("MO", "SE")
igb |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol, Fraction = cat, `n offshore` = n_bg,
            `B (median, mg/kg)` = signif(bg_median, 4),
            `P90 (mg/kg)` = signif(bg_p90, 4),
            Index = case_when(
              as.character(symbol) %in% withheld_el ~ "withheld (below LOQ)",
              !reliable                             ~ "none (n < 30)",
              TRUE                                  ~ "published")) |>
  arrange(Element, Fraction) |>
  kable(format.args = list(big.mark = ","))
```

Two choices in that table are worth stating, because both could reasonably have gone the
other way.

**The median, not the P90.** The 1.5 in the denominator is Muller's allowance for
lithological variability about a *central* value; it is already doing the job a higher
percentile would do. Using the P90 as B as well would apply the tolerance twice and make
almost everything read as unpolluted. The P90 is published beside it so the spread stays
visible.

**Local, not crustal.** These are not Turekian and Wedepohl values. Their deep-sea clay
figures run 9 to 24 times above this background for the clay-borne metals, and four orders of
magnitude below it for iodine, so adopting them would have decided the answer before the data
spoke. EFSA's guidance points the same way.

The right-hand column says whether that reference is turned into an index, and there are two
reasons it is not. Iodine has no usable reference at all: 5 offshore rows in bulk and 5 in
sieved63, well under the 30-row minimum. Selenium and molybdenum have perfectly good
reference counts and are withheld anyway, for a reason that has nothing to do with the
reference and is taken up at the foot of this page.

## How much of the seabed is above background?

```{r}
#| label: class-fig
#| fig-cap: "Igeo class shares, per element and fraction. Copper in bulk is the outlier: 47% of it sits above the local offshore background, and 5% reaches class 4 or higher. Every other group is dominated by class 0."
#| fig-width: 9
#| fig-height: 5.2
igc |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         igeo_class = factor(igeo_class, levels = class_levels)) |>
  ggplot(aes(pct, symbol, fill = igeo_class)) +
  geom_col(width = 0.72) +
  facet_wrap(~ cat, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = class_cols, drop = FALSE) +
  labs(x = "% of measurements", y = NULL, fill = "Igeo class")
```

```{r}
#| label: dist-table
igd |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol, Fraction = cat, n,
            `median Igeo` = igeo_p50, `P90 Igeo` = igeo_p90,
            `% class 0` = pct_unpolluted,
            `% class 2+` = pct_moderate_up,
            `% class 4+` = pct_heavy_up) |>
  arrange(Element, Fraction) |>
  kable(format.args = list(big.mark = ","))
```

Every group has a **negative median**, which is what a median-based background has to give:
half the offshore population sits below its own median, and the 1.5 pushes the whole scale
down by a further 0.58 on the log axis. Class 0 is therefore the expected place for a typical
sample, and the interesting number in that table is the right-hand column.

Copper in bulk is the standout: **23.4% at class 2 or above and 5.1% at class 4 or above**,
against roughly 1% or less for everything else. Manganese and zinc in bulk are the next two,
at 1.0% and 1.7% at class 4 or above. In the sieved fractions the tail nearly vanishes, with no
group above 0.2% at class 4.

::: {.callout-warning}
## The class names are Muller's; the meaning is not

Muller's scale is calibrated against **average shale**. B here is a local offshore median,
which for these metals sits well below shale, so "moderate" on this page means about two and
a half times the local offshore background, not two and a half times the crust.

The boundaries are kept because they are the convention a reviewer will recognise. Do not
read the labels across to a study that used a crustal reference: the same sediment would
score lower there.
:::

## Does it track the pressure gradient?

The background uses no aquaculture information, so distance to the nearest farm is an
**evaluation** of the index, not an input to it.

```{r}
#| label: pressure-fig
#| fig-cap: "Median Igeo by distance to the nearest aquaculture site, bulk fraction. The four elements split: copper and zinc peak at 5 to 20 km, while cobalt and manganese are highest under the cages. Only copper and zinc fall away sharply beyond 20 km."
#| fig-width: 7.5
#| fig-height: 3.4
igp |>
  filter(cat == "bulk") |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         aq_band = factor(aq_band, levels = bin_levels)) |>
  ggplot(aes(aq_band, igeo_p50, group = symbol, colour = symbol)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey50") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  labs(x = "distance to nearest aquaculture site", y = "median Igeo",
       colour = NULL)
```

```{r}
#| label: pressure-table
igp |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         aq_band = factor(aq_band, levels = bin_levels)) |>
  transmute(Element = symbol, Fraction = cat, Band = aq_band, n,
            `median Igeo` = igeo_p50, `% class 0` = pct_unpolluted) |>
  arrange(Element, Fraction, Band) |>
  kable(format.args = list(big.mark = ","))
```

One thing is consistent and one is not.

**Consistent:** by class share, every element is cleanest in the far band. The share scoring
class 0 beyond 20 km is the highest of the four bands for all of them, from 92.9% for cobalt
down to 65.9% for manganese. That is the check that the index works at all, since the far
band is where the background was cut from.

**Not consistent:** the medians do not agree about where the peak is. Copper and zinc peak in
the middle, rising from -0.13 and -0.80 under the cages to +0.66 and +0.14 at 5 to 20 km
before falling away beyond 20 km. Cobalt and manganese peak **under the cages** and decline
outward, manganese most sharply, from +0.24 to -0.50.

So the gradient is real but it is **not monotonic**, and the two pairs of elements disagree
about its shape. That is the honest finding rather than an inconvenience, and this page cannot
resolve it. Two readings are available: the 1 to 20 km ring contains more of the fjord-head
and harbour sediment that carries the region's other pressures, or the immediately-under-cage
sediment is coarser and more organic, which moves concentration without moving contamination.
Note also that the near bands are dominated by copper and zinc (13,066 and 11,854
measurements under the cages, against 144 and 138 for cobalt and manganese), so the two elements that peak nearest
the farms are also the two with the thinnest data there. The [Pressure-Based
Background](./background-pressure.qmd) and [Pressure Controls](./pressure-controls.qmd) pages
attack this directly.

## What the index does not control

Igeo divides by a background, not by a grain-size carrier, so nothing in it separates
"muddy" from "polluted". Measured against the mud fraction, per group, alongside EF on the
rows that carry one:

```{r}
#| label: confound-table
igf |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol, Fraction = cat,
            `n Igeo` = n, `rho Igeo vs fines` = rho_igeo_fines,
            `n EF` = n_ef, `rho EF vs fines` = rho_ef_fines) |>
  arrange(Element, Fraction) |>
  kable(format.args = list(big.mark = ","))
```

In bulk, Igeo tracks the mud fraction for every metal, worst for **cobalt at rho 0.70**. In
the sieved fractions it barely does, which is the normalisability finding from the other
side: a sieved sample is already grain-size controlled, so there is little texture left for
the index to pick up.

**EF is not uniformly the better instrument.** On the same rows it beats Igeo for cobalt
(0.24 against 0.70) and it **loses** for copper (0.55 against 0.36), tying for zinc. Dividing
by aluminium demonstrably removes the texture signal for cobalt and demonstrably does not for
copper. The comparison is between different row populations, since EF only exists where
aluminium does, so read it as a scale rather than a like-for-like test.

The case for Igeo is therefore strongest exactly where EF cannot run at all, in the sieved
fractions and the near-farm data, and a bulk Igeo should be read with the texture caveat
attached.

## The alternative that does not work

If the problem is a missing normaliser, the obvious question is whether another one is
present. Organic carbon is: 33.1% of bulk target rows carry TOC and no aluminium, 28,397
measurements, and metals do sorb to organic matter. Put through the same two-part test
aluminium had to pass (R² at least 0.30 **and** rho at least 0.50):

```{r}
#| label: toc-table
igt |>
  filter(!is.na(r2) | !is.na(rho)) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol, Fraction = cat,
            `n offshore` = n_off, `R2` = r2, rho,
            Normalisable = ifelse(normalisable_toc, "yes", "no")) |>
  arrange(Element, Fraction) |>
  kable(format.args = list(big.mark = ","))
```

It clears both limits in exactly **one group of twenty**, and that group cannot be used:
selenium sieved63, on 34 offshore rows, for an element whose verdicts are already withheld
because below-LOQ censoring deleted 68.6% of it. A strong fit to the top third of a truncated
distribution is not evidence about background.

Everywhere it could matter it fails, and it fails in the way that is hardest to argue with:
**the two measures disagree**. Copper is the best case in bulk, rho 0.60 against R² 0.21, and
copper sieved63 inverts it, R² 0.37 against rho 0.32. For aluminium the two measures agreed
on every group, which is what made the normalisability test a clean partition; for TOC they
pull apart, so there is no cut that turns it into a normaliser. Those 28,397 rows stay
unnormalised, and Igeo remains the only route to a verdict for them.

## What Igeo does not fix

**Selenium and molybdenum stay withheld.** They carry no index, no class and no verdict, so
they are absent from every table on this page that reports one.

Their offshore reference is still published in the reference table, and they still appear in
the organic-carbon test, on the same principle the enrichment pages follow: what is withheld
is the verdict, not the measurement or the background behind it.

Their exclusion was never about normalisers. The clean stage removes below-LOQ rows, which
deletes **68.6% of selenium and 52.2% of molybdenum**, so what survives is an upper tail
wearing the name "background". Both ends of the ratio are truncated: the offshore median in
the denominator is inflated, and the numerator holds only the measurements that cleared the
limit. A different index over the same truncated distribution inherits the same problem.

Normalisability, by contrast, does **not** apply here. That test asks whether aluminium
predicts the metal, and nothing on this page divides by aluminium. Igeo escapes the
normaliser constraint and not the censoring one.
