---
title: "DB Schema"
subtitle: "The refined marine-sediment database the analyses on this site read"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
.mermaid { max-width: 680px; margin-left: auto; margin-right: auto; }
.mermaid svg { max-width: 100% !important; width: 100%; height: auto; }
```

The **refined** database (`multised_refined.sqlite`) is a restructured, slimmed version
of the merged database, holding the reusable results of the earlier analyses so the
pristine / background-sediment analyses on this site do not recompute them. This page is
the schema reference. How the database is **built** from the merged one (the restructure,
the normaliser, the baked ratios, the aquaculture and repeat links) is documented on the
[merged site](https://seafood-hazards.github.io/multised-merged/) under **Refined DB**.

## The rule behind it

One rule decides what the refined database holds:

> **Bake in facts and already-computed reusable values; defer every classification or
> threshold.**

So it carries the **normalisation ratios** (metal / Fe, metal / Al, metal / organic
carbon), the grain-size fines, the aquaculture link and the repeat-sampling grouping, all
as ready-to-use columns. It carries **none** of the decisions the pristine work will make:
no depth bands, no offshore or pressure flags, no enrichment-factor thresholds, no pristine
label. The raw material for those (ratios, distances, raw depth) is present; the calls are
left to the analyses.

## Schema diagram

```{mermaid}
erDiagram
  dataset     ||--o{ event              : "dataset_id"
  site        ||--o{ event              : "site_id"
  aquaculture ||--o{ site               : "aqua_id"
  event       ||--o{ subsample          : "event_id"
  subsample   ||--o{ measurement        : "subsample_id"
  subsample   ||--o{ normaliser         : "subsample_id"
  element     ||--o{ measurement        : "symbol"
  element     ||--o{ method             : "symbol"
  method      ||--o{ measurement        : "method_id"

  aquaculture {
    INTEGER aqua_id PK
  }
  site {
    INTEGER site_id PK
    INTEGER aqua_id FK
    TEXT repeat_group
  }
  event {
    INTEGER event_id PK
    INTEGER dataset_id FK
    INTEGER site_id FK
  }
  subsample {
    INTEGER subsample_id PK
    INTEGER event_id FK
    REAL fines_lt63
  }
  normaliser {
    INTEGER subsample_id FK
    TEXT frac_class
    REAL fe
    REAL al
    REAL corg
  }
  measurement {
    INTEGER measurement_id PK
    INTEGER subsample_id FK
    TEXT symbol FK
    TEXT frac_class
    REAL ratio_fe
  }
  element {
    TEXT symbol PK
  }
  method {
    INTEGER method_id PK
    TEXT symbol FK
  }
```

```{r}
library(tibble)
library(knitr)
```

The nine tables below are in the same order as the
[merged schema](https://seafood-hazards.github.io/multised-merged/db-schema-merged.html),
so the two pages can be read side by side: the seven shared with merged first, then
`normaliser` and `aquaculture`, which exist only here.

## dataset

One row per data-collection campaign. Carried over from the merged database unchanged.

```{r}
tribble(
  ~Column,          ~Type,     ~Meaning,
  "dataset_id",     "INTEGER", "Primary key.",
  "dataset_name",   "TEXT",    "Name of the data-collection campaign / dataset.",
  "dataset_code",   "TEXT",    "Source-native dataset code, where recorded.",
  "dataset_group",  "TEXT",    "Dataset grouping, where recorded.",
  "source",         "TEXT",    "Source database name.",
  "source_type",    "TEXT",    "Kind of source (e.g. national, international).",
  "url",            "TEXT",    "Source URL, where recorded.",
  "doi",            "TEXT",    "Dataset DOI, where recorded.",
  "country",        "TEXT",    "Country of origin.",
  "region",         "TEXT",    "Marine region, where recorded.",
  "institute",      "TEXT",    "Institute responsible for the data.",
  "institute_code", "TEXT",    "Institute code, where recorded.",
  "accessed",       "TEXT",    "Date the source was accessed.",
  "src_dataset_id", "TEXT",    "Original source-prefixed key (provenance)."
) |> kable()
```

## site

One row per geographic location. The geo-enrichment is carried over from merged; the
last three columns are **added here** for the pressure and repeat-sampling analyses.

```{r}
tribble(
  ~Column,               ~Type,     ~Meaning,
  "site_id",             "INTEGER", "Primary key.",
  "latitude",            "REAL",    "Latitude in decimal degrees.",
  "longitude",           "REAL",    "Longitude in decimal degrees.",
  "depth",               "REAL",    "Water depth (m), where recorded.",
  "country",             "TEXT",    "Nearest country.",
  "country_code",        "TEXT",    "ISO country code.",
  "dist_to_coast",       "REAL",    "Distance to nearest coastline (km).",
  "municipality",        "TEXT",    "Nearest municipality.",
  "sea_name",            "TEXT",    "Name of the sea or ocean body.",
  "dist_to_aquaculture", "REAL",    "Distance to nearest Norwegian aquaculture site (km); Norway sites only, else NULL.",
  "aqua_id",             "INTEGER", "Added here: -> aquaculture, the nearest aquaculture site (Norway only, else NULL).",
  "source",              "TEXT",    "Source database name.",
  "src_site_id",         "TEXT",    "Original source-prefixed key (provenance).",
  "repeat_group",        "TEXT",    "Added here: location cell (lat/lon to 2 dp, ~1.1 km) for repeat-sampling grouping.",
  "n_years",             "INTEGER", "Added here: distinct sampling years at that cell (over all dated events)."
) |> kable()
```

## event

One row per sampling event. Carried over from the merged database unchanged.

```{r}
tribble(
  ~Column,         ~Type,     ~Meaning,
  "event_id",      "INTEGER", "Primary key.",
  "dataset_id",    "INTEGER", "-> dataset.",
  "site_id",       "INTEGER", "-> site.",
  "year",          "INTEGER", "Year of the sampling event.",
  "date",          "TEXT",    "Date of the sampling event, where recorded.",
  "sampling_tool", "TEXT",    "Sampling gear, descriptive name.",
  "n_layers",      "INTEGER", "Number of subsamples (depth layers) under the event.",
  "source",        "TEXT",    "Source database name.",
  "src_event_id",  "TEXT",    "Original source-prefixed key (provenance)."
) |> kable()
```

## subsample

Physical properties of a depth slice, including the derived grain-size fines. Depths are
kept **raw**: no banding is applied, so the analyses that need bands choose their own.

```{r}
tribble(
  ~Column,             ~Type,     ~Meaning,
  "subsample_id",      "INTEGER", "Primary key.",
  "event_id",          "INTEGER", "-> event.",
  "depth_from",        "REAL",    "Top of the slice, cm (raw, not banded), else NULL.",
  "depth_to",          "REAL",    "Bottom of the slice, cm, else NULL.",
  "depth_flag",        "TEXT",    "'implausible' where the reported depth was removed (Vannmiljø only), else NULL.",
  "fe_exist",          "INTEGER", "1/0: an Fe value exists for this subsample.",
  "al_exist",          "INTEGER", "1/0: an Al value exists.",
  "org_exist",         "INTEGER", "1/0: an organic carbon value exists.",
  "comp_exist",        "INTEGER", "1/0: grain-size composition exists.",
  "target_frac_class", "TEXT",    "Target-fraction summary: 'bulk', 'sieved', or 'mixed', else NULL.",
  "target_sieve_um",   "REAL",    "Target sieve cutoff where unambiguous, else NULL.",
  "fines_lt63",        "REAL",    "% finer than 63 um (mud content) of the PARENT sediment as collected, else NULL. It describes the sample the subsample was taken from, never a sieved aliquot: a cut below 63 um would be 100% by definition, yet the median here for subsamples whose targets are sieved to <63 um is about 15%, lower than the 50% for bulk, because sandy sediment is what gets sieved. So it must not be used to normalise a sieved measurement.",
  "fines_basis",       "TEXT",    "How fines_lt63 was derived.",
  "source",            "TEXT",    "Source database name.",
  "src_subsample_id",  "TEXT",    "Original source-prefixed key (provenance)."
) |> kable()
```

## measurement

One row per target-element reading, bulk and sieved told apart by `frac_class` (never
pooled). Holds only the 7 target elements (Fe, Al and organic carbon moved to
`normaliser`), with the normalisation ratios baked in. Sieve columns are NULL for bulk.

```{r}
tribble(
  ~Column,             ~Type,     ~Meaning,
  "measurement_id",    "INTEGER", "Primary key.",
  "subsample_id",      "INTEGER", "-> subsample.",
  "symbol",            "TEXT",    "Target element (-> element); one of CO CU I MN MO SE ZN.",
  "value",             "REAL",    "Value in the original unit.",
  "unit",              "TEXT",    "Original reporting unit.",
  "value_std",         "REAL",    "Standardised value, mg/kg.",
  "value_sd",          "REAL",    "SD across replicates, else NULL.",
  "n_rep",             "INTEGER", "Replicates averaged (1 = single).",
  "value_uncrt",       "REAL",    "Analytical uncertainty, mg/kg, else NULL.",
  "frac_class",        "TEXT",    "Sediment fraction: 'bulk' or 'sieved'.",
  "sieve_um_std",      "REAL",    "Harmonised sieve cutoff (um); NULL for bulk.",
  "sieve_class",       "TEXT",    "Sieve label, e.g. '<63um'; NULL for bulk.",
  "method_id",         "INTEGER", "-> method.",
  "source",            "TEXT",    "Source database name.",
  "src_measurement_id","TEXT",    "Original key (provenance).",
  "outlier_flag",      "TEXT",    "Distributional outlier marker: 'high' / 'low', else NULL.",
  "ratio_fe",          "REAL",    "Added here: value_std / Fe (same subsample, fraction), else NULL.",
  "ratio_al",          "REAL",    "Added here: value_std / Al, else NULL.",
  "ratio_corg",        "REAL",    "Added here: value_std / organic carbon, else NULL."
) |> kable()
```

## element

The shared element lookup, cut down to the 7 targets. A single vocabulary across all
sources, so it carries **no** `source` column.

```{r}
tribble(
  ~Column,    ~Type,  ~Meaning,
  "symbol",   "TEXT", "Primary key: element in the ICES name (CO, CU, I, MN, MO, SE, ZN).",
  "name",     "TEXT", "Full element name.",
  "category", "TEXT", "Measurand class: 'target' for every row here.",
  "cas",      "TEXT", "CAS registry number, where known."
) |> kable()
```

## method

One row per distinct analytical method, keeping the detection limits. Carried over from
the merged database, restricted to the methods the target measurements use.

```{r}
tribble(
  ~Column,              ~Type,     ~Meaning,
  "method_id",          "INTEGER", "Primary key.",
  "symbol",             "TEXT",    "-> element.",
  "method",             "TEXT",    "Analytical technique (e.g. ICP-MS, ICP-AES).",
  "method_description", "TEXT",    "Free-text method description, where recorded.",
  "lab",                "TEXT",    "Laboratory that performed the analysis, where recorded.",
  "lab_name",           "TEXT",    "Full laboratory name, where recorded.",
  "lod",                "REAL",    "Limit of detection, where recorded.",
  "loq",                "REAL",    "Limit of quantification, where recorded.",
  "limit_unit",         "TEXT",    "Reporting unit of lod / loq.",
  "source",             "TEXT",    "Source database name.",
  "src_method_id",      "TEXT",    "Original source-prefixed key (provenance)."
) |> kable()
```

## normaliser

**This database only.** The Fe / Al / organic-carbon values, one wide row per
**subsample and fraction** (the grain a ratio needs, since a subsample can carry both
fractions). In merged these are measurement rows like any other; here they are pulled
out so a target row can reach its normaliser without a self-join.

```{r}
tribble(
  ~Column,        ~Type,     ~Meaning,
  "subsample_id", "INTEGER", "-> subsample.",
  "frac_class",   "TEXT",    "'bulk' or 'sieved'.",
  "fe",           "REAL",    "Iron, mg/kg, else NULL.",
  "al",           "REAL",    "Aluminium, mg/kg, else NULL.",
  "corg",         "REAL",    "Organic carbon, mg/kg, else NULL."
) |> kable()
```

## aquaculture

**This database only.** The Norwegian aquaculture reference, folded in as a table rather
than referenced, so the mart is self-contained. Linked from `site.aqua_id`. Norwegian
marine aquaculture sites (active and closed), from the Norwegian Directorate of Fisheries
register; **Norway only**, so `aqua_id` and `dist_to_aquaculture` are populated only for
Norwegian sediment sites. How the table is built is on the
[clean-analyses site](https://seafood-hazards.github.io/multised-clean/).

```{r}
tribble(
  ~Column,           ~Type,     ~Meaning,
  "aqua_id",         "INTEGER", "Primary key (referenced by site.aqua_id).",
  "loknr",           "INTEGER", "Site number in the Norwegian register.",
  "name",            "TEXT",    "Site name (kept in Norwegian).",
  "latitude",        "REAL",    "Latitude (WGS84), from the national grid (ETRS89 / UTM 33N).",
  "longitude",       "REAL",    "Longitude (WGS84), from the national grid.",
  "start_year",      "INTEGER", "Year the site was cleared.",
  "end_year",        "INTEGER", "Year the site was withdrawn (NULL for active sites).",
  "active",          "INTEGER", "1 for active, 0 for closed.",
  "water_type",      "TEXT",    "salt, brackish, or fresh/salt.",
  "capacity",        "REAL",    "Cleared (permitted) capacity, as reported.",
  "capacity_unit",   "TEXT",    "Unit of capacity (tonnes, decare, count, ...).",
  "capacity_tonnes", "REAL",    "Capacity converted to tonnes where the unit is a mass; else NULL.",
  "fish_types",      "TEXT",    "Species farmed, translated to English (rare species keep Norwegian).",
  "placement",       "TEXT",    "sea or land.",
  "county",          "TEXT",    "County (first name kept where the source is multilingual).",
  "municipality",    "TEXT",    "Municipality (first name kept where the source is multilingual)."
) |> kable()
```

::: {.callout-note}
The refined database is deliberately incomplete by design: it stops at the reusable
building blocks. The pristine / background analyses that consume it (enrichment factors,
depth bands, pressure and pristine classification) are the pages this site will add;
their choices are never frozen into the data. `dataset`, `event`, `subsample` and
`method` are carried over unchanged, so their columns are identical to the
[merged schema](https://seafood-hazards.github.io/multised-merged/db-schema-merged.html);
`site`, `measurement` and `element` differ from it, and `normaliser` and `aquaculture`
have no counterpart there.
:::
