---
title: "DB Schema"
subtitle: "The refined marine-sediment database the analyses on this site read"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
.mermaid { max-width: 680px; margin-left: auto; margin-right: auto; }
.mermaid svg { max-width: 100% !important; width: 100%; height: auto; }
```

The **refined** database (`multised_refined.sqlite`) is a restructured, slimmed version
of the merged database, holding the reusable results of the earlier analyses so the
pristine / background-sediment analyses on this site do not recompute them. This page is
the schema reference. How the database is **built** from the merged one (the restructure,
the normaliser, the baked ratios, the aquaculture and repeat links) is documented on the
[merged site](https://seafood-hazards.github.io/multised-merged/) under **Refined DB**.

## The rule behind it

One rule decides what the refined database holds:

> **Bake in facts and already-computed reusable values; defer every classification or
> threshold.**

So it carries the **normalisation ratios** (metal / Fe, metal / Al, metal / organic
carbon), the grain-size fines, the aquaculture link and the repeat-sampling grouping, all
as ready-to-use columns. It carries **none** of the decisions the pristine work will make:
no depth bands, no offshore or pressure flags, no enrichment-factor thresholds, no pristine
label. The raw material for those (ratios, distances, raw depth) is present; the calls are
left to the analyses.

## Schema diagram

```{mermaid}
erDiagram
  dataset     ||--o{ event              : "dataset_id"
  site        ||--o{ event              : "site_id"
  aquaculture ||--o{ site               : "aqua_id"
  event       ||--o{ subsample          : "event_id"
  subsample   ||--o{ measurement        : "subsample_id"
  subsample   ||--o{ normaliser         : "subsample_id"
  element     ||--o{ measurement        : "symbol"
  element     ||--o{ method             : "symbol"
  method      ||--o{ measurement        : "method_id"

  aquaculture {
    INTEGER aqua_id PK
  }
  site {
    INTEGER site_id PK
    INTEGER aqua_id FK
    TEXT repeat_group
  }
  event {
    INTEGER event_id PK
    INTEGER dataset_id FK
    INTEGER site_id FK
  }
  subsample {
    INTEGER subsample_id PK
    INTEGER event_id FK
    REAL fines_lt63
  }
  normaliser {
    INTEGER subsample_id FK
    TEXT frac_class
    REAL fe
    REAL al
    REAL corg
  }
  measurement {
    INTEGER measurement_id PK
    INTEGER subsample_id FK
    TEXT symbol FK
    TEXT frac_class
    REAL ratio_fe
  }
  element {
    TEXT symbol PK
  }
  method {
    INTEGER method_id PK
    TEXT symbol FK
  }
```

```{r}
library(tibble)
library(knitr)
```

## measurement

One row per target-element reading, bulk and sieved told apart by `frac_class` (never
pooled). Holds only the 7 target elements (Fe, Al and organic carbon moved to
`normaliser`), with the normalisation ratios baked in. Sieve columns are NULL for bulk.

```{r}
tribble(
  ~Column,             ~Type,     ~Meaning,
  "measurement_id",    "INTEGER", "Primary key.",
  "subsample_id",      "INTEGER", "-> subsample.",
  "symbol",            "TEXT",    "Target element (-> element); one of CO CU I MN MO SE ZN.",
  "frac_class",        "TEXT",    "Sediment fraction: 'bulk' or 'sieved'.",
  "value",             "REAL",    "Value in the original unit.",
  "unit",              "TEXT",    "Original reporting unit.",
  "value_std",         "REAL",    "Standardised value, mg/kg.",
  "value_sd",          "REAL",    "SD across replicates, else NULL.",
  "n_rep",             "INTEGER", "Replicates averaged (1 = single).",
  "value_uncrt",       "REAL",    "Analytical uncertainty, mg/kg, else NULL.",
  "method_id",         "INTEGER", "-> method.",
  "ratio_fe",          "REAL",    "value_std / Fe (same subsample, fraction), else NULL.",
  "ratio_al",          "REAL",    "value_std / Al, else NULL.",
  "ratio_corg",        "REAL",    "value_std / organic carbon, else NULL.",
  "sieve_um_std",      "REAL",    "Harmonised sieve cutoff (um); NULL for bulk.",
  "sieve_class",       "TEXT",    "Sieve label, e.g. '<63um'; NULL for bulk.",
  "source",            "TEXT",    "Source database name.",
  "src_measurement_id","TEXT",    "Original key (provenance).",
  "outlier_flag",      "TEXT",    "Distributional outlier marker: 'high' / 'low', else NULL."
) |> kable()
```

## normaliser

The Fe / Al / organic-carbon values, one wide row per **subsample and fraction** (the
grain a ratio needs, since a subsample can carry both fractions).

```{r}
tribble(
  ~Column,        ~Type,     ~Meaning,
  "subsample_id", "INTEGER", "-> subsample.",
  "frac_class",   "TEXT",    "'bulk' or 'sieved'.",
  "fe",           "REAL",    "Iron, mg/kg, else NULL.",
  "al",           "REAL",    "Aluminium, mg/kg, else NULL.",
  "corg",         "REAL",    "Organic carbon, mg/kg, else NULL."
) |> kable()
```

## site (analysis columns)

One row per location. Beyond the geo-enrichment (depth, `dist_to_coast`, `sea_name`,
`dist_to_aquaculture`), it carries:

```{r}
tribble(
  ~Column,        ~Type,     ~Meaning,
  "aqua_id",      "INTEGER", "-> aquaculture: the nearest aquaculture site (Norway only, else NULL).",
  "repeat_group", "TEXT",    "Location cell (lat/lon to 2 dp, ~1.1 km) for repeat-sampling grouping.",
  "n_years",      "INTEGER", "Distinct sampling years at that cell (over all dated events)."
) |> kable()
```

## subsample

Physical properties of a depth slice, including the derived grain-size fines.

```{r}
tribble(
  ~Column,        ~Type,     ~Meaning,
  "subsample_id", "INTEGER", "Primary key.",
  "event_id",     "INTEGER", "-> event.",
  "depth_from",   "REAL",    "Top of the slice, cm (raw, not banded), else NULL.",
  "depth_to",     "REAL",    "Bottom of the slice, cm, else NULL.",
  "fines_lt63",   "REAL",    "% finer than 63 um (mud content), else NULL.",
  "fines_basis",  "TEXT",    "How fines_lt63 was derived."
) |> kable()
```

## aquaculture

The Norwegian aquaculture reference, a table inside this database, linked from
`site.aqua_id`. Norwegian marine aquaculture sites (active and closed), from the
Norwegian Directorate of Fisheries register; **Norway only**, so `aqua_id` and
`dist_to_aquaculture` are populated only for Norwegian sediment sites. How the table is
built is on the [clean-analyses site](https://seafood-hazards.github.io/multised-clean/).

```{r}
tribble(
  ~Column,           ~Type,     ~Meaning,
  "aqua_id",         "INTEGER", "Primary key (referenced by site.aqua_id).",
  "loknr",           "INTEGER", "Site number in the Norwegian register.",
  "name",            "TEXT",    "Site name (kept in Norwegian).",
  "latitude",        "REAL",    "Latitude (WGS84), from the national grid (ETRS89 / UTM 33N).",
  "longitude",       "REAL",    "Longitude (WGS84), from the national grid.",
  "start_year",      "INTEGER", "Year the site was cleared.",
  "end_year",        "INTEGER", "Year the site was withdrawn (NULL for active sites).",
  "active",          "INTEGER", "1 for active, 0 for closed.",
  "water_type",      "TEXT",    "salt, brackish, or fresh/salt.",
  "capacity",        "REAL",    "Cleared (permitted) capacity, as reported.",
  "capacity_unit",   "TEXT",    "Unit of capacity (tonnes, decare, count, ...).",
  "capacity_tonnes", "REAL",    "Capacity converted to tonnes where the unit is a mass; else NULL.",
  "fish_types",      "TEXT",    "Species farmed, translated to English (rare species keep Norwegian).",
  "placement",       "TEXT",    "sea or land.",
  "county",          "TEXT",    "County (first name kept where the source is multilingual).",
  "municipality",    "TEXT",    "Municipality (first name kept where the source is multilingual)."
) |> kable()
```

::: {.callout-note}
The refined database is deliberately incomplete by design: it stops at the reusable
building blocks. The pristine / background analyses that consume it (enrichment factors,
depth bands, pressure and pristine classification) are the pages this site will add;
their choices are never frozen into the data. Dimensions carried unchanged (`dataset`,
`event`, `method` with its detection limits) match the
[merged schema](https://seafood-hazards.github.io/multised-merged/db-schema-merged.html).
:::
