---
title: "Dataset Download"
subtitle: "A single flat table of the values needed for background and enrichment work"
---

For users who want to try their **own** background or enrichment methods, this is a single
**denormalised table**: one row per target measurement, with only the values those analyses
need. It is drawn from the [refined database](./download-database.qmd) but drops the
relational structure, the surrogate keys, the provenance and the uncertainty / method
detail. The original **source** of each measurement is kept, so results can be traced back
or filtered by programme.

It also carries **this site's own verdicts**: whether each measurement is pristine, and
whether it sits at or below background, under the definitions the analysis pages use.
Those columns are described under [The verdict columns](#the-verdict-columns) below, with
the reference values behind them, so any of it can be recomputed or replaced with your own
rule.

The file is a **tab-separated, gzip-compressed** text file (`.tsv.gz`), readable by any
data tool.

```{r}
#| label: download-button
#| output: asis
url <- "https://github.com/seafood-hazards/multised-refined/releases/latest/download/multised_refined_dataset.tsv.gz"
size_mb <- round(file.size("multised_refined_dataset.tsv.gz") / 1024^2, 2)
cat(sprintf(
  '<a class="btn btn-primary" href="%s" download role="button">Download multised_refined_dataset.tsv.gz (%s MB)</a>\n',
  url, size_mb))
```

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(knitr)

d <- read_tsv("multised_refined_dataset.tsv.gz", show_col_types = FALSE)
dict <- read_csv("data/refined_summary/refined_dataset_dictionary.csv", show_col_types = FALSE)
```

```{r}
#| label: size
#| output: asis
cat(sprintf("The dataset has **%s rows** and **%d columns**.\n\n",
            format(nrow(d), big.mark = ","), ncol(d)))
```

## Columns

Each row is one element measurement at one site, layer and fraction. The normalisers
(`al_mgkg`, `fe_mgkg`, `corg_mgkg`) and `fines_pct` are the covariates for grain-size and
Fe / Al / organic normalisation; `dist_to_coast_km` and `dist_to_aquaculture_km` support
offshore and pressure-based backgrounds.

```{r}
#| label: dictionary
dict |>
  mutate(unit = ifelse(is.na(unit) | unit == "", "-", unit)) |>
  rename(Column = column, Unit = unit, Description = description) |>
  kable()
```

## The verdict columns

Several columns are not measurements but conclusions, carried over from the analysis pages
so the file can be used without reproducing them.

::: {.callout-important}
## The verdict columns changed in v0.9.0 and again in v0.9.5

Verdicts computed before v0.9.0 will not reproduce from this file. Since v0.9.5 the
enrichment and pristine columns exist for **cobalt, copper and zinc in the bulk fraction
only**: no other element or fraction has a normaliser that predicts it, so no verdict is
issued. The methods, what forced each change and the size of every shift are set out on
[Method Revisions](./method-revisions.qmd). `ef` and `pristine_ef` keep their original
definitions, so an earlier download remains a subset of this one rather than contradicting
it, and `background_p90` and `background_mixture` use no aluminium and are unchanged for
every group.
:::

**Pristine**, from [Pristine Classification](./background-pristine.qmd), is grain-size
controlled and therefore needs aluminium. Where aluminium is missing the sample is left
**unclassified** (`classifiable` FALSE, both pristine columns empty) rather than guessed.
Two further restrictions now apply:

- `al_basis` records the aluminium measurement basis inferred for the subsample. In bulk,
  only samples on the basis their fraction adopted receive a verdict; the rest are
  unclassified for the same reason a sample with no aluminium is. It does not gate the
  sieved fractions, which use no aluminium.
- **Selenium and molybdenum carry no verdict at all.** Most of their measurements fell
  below the limit of quantification and were removed upstream, so what survives is an upper
  tail rather than a background. Their concentrations are still published.
- **Nor does manganese in bulk.** An enrichment factor is a grain-size control only if the
  normaliser carries the metal, and measured by regression aluminium does so for cobalt,
  copper and zinc in bulk (R-squared 0.46 to 0.58) and for nothing else. Manganese in bulk
  keeps its concentrations and loses the verdict, because aluminium was the control on offer
  and nothing replaces it. See [Regression Normalisation](./background-regression.qmd).
- **The sieved fractions carry `pristine_strict` but not `pristine_ef`.** They are
  grain-size controlled by the sieve rather than by aluminium, so no enrichment factor is
  defined for them and `ef`, `pristine_ef` and `bg_ratio_al` are empty by definition rather
  than for want of data. `pristine_strict` there is the two concentration criteria, the
  offshore P90 and the mixture threshold. Read those verdicts against the sieved fractions'
  own offshore reference, which is the southern North Sea rather than the Norwegian Sea:
  see [Pristine Classification](./background-pristine.qmd).

- `pristine_ef`: the permissive rule, EF < 1, where EF is the element/Al ratio divided by
  the offshore background ratio. This is the rule an EFSA submission would use.
- `pristine_strict`: the conservative rule. EF < 1 **and** below the mixture threshold
  **and** below the offshore P90. The two extra criteria are raw-concentration screens
  rather than peers of EF < 1, so a muddy but clean sample can fail this rule on grain size
  alone; see [Pristine Classification](./background-pristine.qmd). Where the mixture
  threshold is not usable for an element, that criterion drops out rather than being
  applied.
- `ef_p90ref` and `pristine_ef_p90ref`: the same rule read against the offshore 90th
  percentile of element/Al instead of its median. EF < 1 against a median denominator fails
  half the reference population by construction; against the P90 it means "below the
  offshore P90", which is how `background_p90` defines background. The gap between the two
  is the width of the answer.

**Background** is the weaker question of whether the concentration alone sits in the
background population. Both are raw-concentration rules, so unlike the pristine columns
they are **not** grain-size controlled, and a sandy sample can pass on grain size alone:

- `background_p90`: below the offshore P90 for its element and fraction
  ([Global Background](./background.qmd)).
- `background_mixture`: below the threshold separating the two fitted populations
  ([Distribution-Mixture Background](./background-mixture.qmd)).

`ef` is the numeric enrichment factor itself, for anyone who wants a different cutoff than
1. `bg_ratio_al`, `p90_off` and `mixture_threshold` are the three reference values applied
to that row, so every verdict in the file can be checked against the file alone.

```{r}
#| label: verdicts
d |>
  summarise(across(c(classifiable, pristine_ef, pristine_strict,
                     background_p90, background_mixture),
                   list(rows = ~sum(!is.na(.x)), true = ~sum(.x, na.rm = TRUE)))) |>
  pivot_longer(everything(), names_to = c("Column", ".value"), names_sep = "_(?=rows$|true$)") |>
  mutate(`Rows with a verdict` = format(rows, big.mark = ","),
         `TRUE` = format(true, big.mark = ","),
         `Share TRUE` = sprintf("%.1f%%", 100 * true / rows)) |>
  select(Column, `Rows with a verdict`, `TRUE`, `Share TRUE`) |>
  kable(align = "lrrr")
```

::: {.callout-note}
The verdicts are computed on the same basis as the analysis pages: fractions `bulk`,
`sieved63` and `sieved20` only, outlier-flagged rows excluded, non-positive values
excluded. Rows outside that scope carry empty verdicts rather than verdicts derived some
other way, which is why the counts above fall short of the row total.
:::

## How the fractions and coverage look

Counts of rows per element and fraction (bulk and the sieve cutoffs are never pooled;
iodine and selenium are sparse, as noted throughout the analyses):

```{r}
#| label: coverage
d |>
  count(element, fraction) |>
  pivot_wider(names_from = fraction, values_from = n, values_fill = 0) |>
  rename(Element = element) |>
  kable(format.args = list(big.mark = ","))
```

## Preview

The first rows of the dataset:

```{r}
#| label: preview
d |> head(8) |> kable()
```

## Notes for reuse

- Concentrations are **standardised dry-weight mg/kg** (`value_mgkg`); grain-size
  `fines_pct` is a percentage.
- Empty cells are genuine missing values (for example, `al_mgkg` is empty where aluminium
  was not measured, and `dist_to_aquaculture_km` is empty outside Norway).
- `outlier_flag` marks distributional outliers (mostly registration errors); the analyses
  on this site exclude flagged rows, and you probably want to as well.
- A site is identified by its `latitude` / `longitude` (rounded to 3 decimal places); the
  same location can appear across several years.
- The verdict columns are a starting point, not a ruling. `pristine_ef` is per element and
  fraction, not a site-level judgement, and the classifiable half of the data is not a
  random half: aluminium is scarcest close to the aquaculture farms, which is exactly where
  a verdict would matter most.
