---
title: "Dataset Download"
subtitle: "A single flat table of the values needed for background and enrichment work"
---

For users who want to try their **own** background or enrichment methods, this is a single
**denormalised table**: one row per target measurement, with only the values those analyses
need. It is drawn from the [refined database](./download-database.qmd) but drops the
relational structure, the surrogate keys, the provenance and the uncertainty / method
detail. The original **source** of each measurement is kept, so results can be traced back
or filtered by programme.

It also carries **this site's own verdicts**: whether each measurement is pristine, and
whether it sits at or below background, under the definitions the analysis pages use.
Those columns are described under [The verdict columns](#the-verdict-columns) below, with
the reference values behind them, so any of it can be recomputed or replaced with your own
rule.

The file is a **tab-separated, gzip-compressed** text file (`.tsv.gz`), readable by any
data tool.

```{r}
#| label: download-button
#| output: asis
url <- "https://github.com/seafood-hazards/multised-refined/releases/latest/download/multised_refined_dataset.tsv.gz"
size_mb <- round(file.size("multised_refined_dataset.tsv.gz") / 1024^2, 2)
cat(sprintf(
  '<a class="btn btn-primary" href="%s" download role="button">Download multised_refined_dataset.tsv.gz (%s MB)</a>\n',
  url, size_mb))
```

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(knitr)

d <- read_tsv("multised_refined_dataset.tsv.gz", show_col_types = FALSE)
dict <- read_csv("data/refined_summary/refined_dataset_dictionary.csv", show_col_types = FALSE)
```

```{r}
#| label: size
#| output: asis
cat(sprintf("The dataset has **%s rows** and **%d columns**.\n\n",
            format(nrow(d), big.mark = ","), ncol(d)))
```

## Columns

Each row is one element measurement at one site, layer and fraction. The normalisers
(`al_mgkg`, `fe_mgkg`, `corg_mgkg`) and `fines_pct` are the covariates for grain-size and
Fe / Al / organic normalisation; `dist_to_coast_km` and `dist_to_aquaculture_km` support
offshore and pressure-based backgrounds.

```{r}
#| label: dictionary
dict |>
  mutate(unit = ifelse(is.na(unit) | unit == "", "-", unit)) |>
  rename(Column = column, Unit = unit, Description = description) |>
  kable()
```

## The verdict columns

Nine of the columns are not measurements but conclusions, carried over from the analysis
pages so the file can be used without reproducing them.

**Pristine**, from [Pristine Classification](./background-pristine.qmd), is grain-size
controlled and therefore needs aluminium. Where aluminium is missing the sample is left
**unclassified** (`classifiable` FALSE, both pristine columns empty) rather than guessed:

- `pristine_ef`: the permissive rule, EF < 1, where EF is the element/Al ratio divided by
  the offshore background ratio. This is the rule an EFSA submission would use.
- `pristine_strict`: the conservative rule. EF < 1 **and** below the mixture threshold
  **and** below the offshore P90. The two extra criteria are raw-concentration screens
  rather than peers of EF < 1, so a muddy but clean sample can fail this rule on grain size
  alone; see [Pristine Classification](./background-pristine.qmd).

**Background** is the weaker question of whether the concentration alone sits in the
background population. Both are raw-concentration rules, so unlike the pristine columns
they are **not** grain-size controlled, and a sandy sample can pass on grain size alone:

- `background_p90`: below the offshore P90 for its element and fraction
  ([Global Background](./background.qmd)).
- `background_mixture`: below the threshold separating the two fitted populations
  ([Distribution-Mixture Background](./background-mixture.qmd)).

`ef` is the numeric enrichment factor itself, for anyone who wants a different cutoff than
1. `bg_ratio_al`, `p90_off` and `mixture_threshold` are the three reference values applied
to that row, so every verdict in the file can be checked against the file alone.

```{r}
#| label: verdicts
d |>
  summarise(across(c(classifiable, pristine_ef, pristine_strict,
                     background_p90, background_mixture),
                   list(rows = ~sum(!is.na(.x)), true = ~sum(.x, na.rm = TRUE)))) |>
  pivot_longer(everything(), names_to = c("Column", ".value"), names_sep = "_(?=rows$|true$)") |>
  mutate(`Rows with a verdict` = format(rows, big.mark = ","),
         `TRUE` = format(true, big.mark = ","),
         `Share TRUE` = sprintf("%.1f%%", 100 * true / rows)) |>
  select(Column, `Rows with a verdict`, `TRUE`, `Share TRUE`) |>
  kable(align = "lrrr")
```

::: {.callout-note}
The verdicts are computed on the same basis as the analysis pages: fractions `bulk`,
`sieved63` and `sieved20` only, outlier-flagged rows excluded, non-positive values
excluded. Rows outside that scope carry empty verdicts rather than verdicts derived some
other way, which is why the counts above fall short of the row total.
:::

## How the fractions and coverage look

Counts of rows per element and fraction (bulk and the sieve cutoffs are never pooled;
iodine and selenium are sparse, as noted throughout the analyses):

```{r}
#| label: coverage
d |>
  count(element, fraction) |>
  pivot_wider(names_from = fraction, values_from = n, values_fill = 0) |>
  rename(Element = element) |>
  kable(format.args = list(big.mark = ","))
```

## Preview

The first rows of the dataset:

```{r}
#| label: preview
d |> head(8) |> kable()
```

## Notes for reuse

- Concentrations are **standardised dry-weight mg/kg** (`value_mgkg`); grain-size
  `fines_pct` is a percentage.
- Empty cells are genuine missing values (for example, `al_mgkg` is empty where aluminium
  was not measured, and `dist_to_aquaculture_km` is empty outside Norway).
- `outlier_flag` marks distributional outliers (mostly registration errors); the analyses
  on this site exclude flagged rows, and you probably want to as well.
- A site is identified by its `latitude` / `longitude` (rounded to 3 decimal places); the
  same location can appear across several years.
- The verdict columns are a starting point, not a ruling. `pristine_ef` is per element and
  fraction, not a site-level judgement, and the classifiable half of the data is not a
  random half: aluminium is scarcest close to the aquaculture farms, which is exactly where
  a verdict would matter most.
