---
title: "Database Download"
subtitle: "The full refined database as a single SQLite file"
---

The complete **refined database** is available as one **SQLite** file. It holds everything
the analyses on this site are built from: the seven target elements with the Fe / Al /
organic ratios baked in, the normaliser table, the grain-size fines, the site
geo-enrichment, the repeat-sampling grouping, and the Norwegian aquaculture reference (as a
table inside the database). Its layout is described on the [DB Schema](./db-schema-refined.qmd)
page, and you can browse it in your browser on the [Tables](./refined-tables.qmd) page.

```{r}
#| label: download-button
#| output: asis
url <- "https://github.com/seafood-hazards/multised-refined/releases/latest/download/multised_refined.sqlite"
size_mb <- round(file.size("multised_refined.sqlite") / 1024^2, 1)
cat(sprintf(
  '<a class="btn btn-primary" href="%s" download role="button">Download multised_refined.sqlite (%s MB)</a>\n',
  url, size_mb))
```

## What is inside

```{r}
#| label: tables
library(DBI)
library(RSQLite)
library(dplyr)
library(knitr)

con <- dbConnect(SQLite(), "multised_refined.sqlite")
tabs <- dbListTables(con)
descr <- c(
  measurement = "The 7 target elements (bulk and sieved by frac_class) with ratio_fe / ratio_al / ratio_corg.",
  normaliser  = "Fe / Al / organic carbon per subsample and fraction (the ratio denominators).",
  subsample   = "Sediment layers (depth, grain-size fines).",
  event       = "Sampling events (year, tool, layering).",
  site        = "Locations with geo-enrichment, the aquaculture link and repeat-sampling grouping.",
  element     = "The element lookup (7 targets).",
  method      = "Analytical methods used by the targets.",
  dataset     = "Source campaigns / datasets.",
  aquaculture = "Norwegian marine aquaculture sites, linked from site.aqua_id."
)
counts <- sapply(tabs, function(t)
  dbGetQuery(con, sprintf("SELECT COUNT(*) n FROM `%s`", t))$n)
dbDisconnect(con)

tibble(Table = tabs,
       Rows = format(counts, big.mark = ","),
       Contents = descr[tabs]) |>
  arrange(match(Table, names(descr))) |>
  kable()
```

## Format and use

The file is a standard **SQLite 3** database, readable without this site. A few ways in:

- **R**: `DBI::dbConnect(RSQLite::SQLite(), "multised_refined.sqlite")`, then `dbGetQuery()`.
- **Python**: the built-in `sqlite3`, or `pandas.read_sql`.
- **Command line**: `sqlite3 multised_refined.sqlite`.

If you only need the flat values for background / enrichment work, the
[Dataset Download](./download-dataset.qmd) page offers a single denormalised table instead.
