---
title: "Dataset Download"
subtitle: "A single flat table of the values needed for background and enrichment work"
---

For users who want to try their **own** background or enrichment methods, this is a single
**denormalised table**: one row per target measurement, with only the values those analyses
need. It is drawn from the [refined database](./download-database.qmd) but drops the
relational structure, the surrogate keys, the provenance and the uncertainty / method
detail. The original **source** of each measurement is kept, so results can be traced back
or filtered by programme.

The file is a **tab-separated, gzip-compressed** text file (`.tsv.gz`), readable by any
data tool.

```{r}
#| label: download-button
#| output: asis
url <- "https://github.com/seafood-hazards/multised-refined/releases/download/v0.1.0/multised_refined_dataset.tsv.gz"
size_mb <- round(file.size("multised_refined_dataset.tsv.gz") / 1024^2, 2)
cat(sprintf(
  '<a class="btn btn-primary" href="%s" download role="button">Download multised_refined_dataset.tsv.gz (%s MB)</a>\n',
  url, size_mb))
```

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(knitr)

d <- read_tsv("multised_refined_dataset.tsv.gz", show_col_types = FALSE)
dict <- read_csv("data/refined_summary/refined_dataset_dictionary.csv", show_col_types = FALSE)
```

```{r}
#| label: size
#| output: asis
cat(sprintf("The dataset has **%s rows** and **%d columns**.\n\n",
            format(nrow(d), big.mark = ","), ncol(d)))
```

## Columns

Each row is one element measurement at one site, layer and fraction. The normalisers
(`al_mgkg`, `fe_mgkg`, `corg_mgkg`) and `fines_pct` are the covariates for grain-size and
Fe / Al / organic normalisation; `dist_to_coast_km` and `dist_to_aquaculture_km` support
offshore and pressure-based backgrounds.

```{r}
#| label: dictionary
dict |>
  mutate(unit = ifelse(is.na(unit) | unit == "", "-", unit)) |>
  rename(Column = column, Unit = unit, Description = description) |>
  kable()
```

## How the fractions and coverage look

Counts of rows per element and fraction (bulk and the sieve cutoffs are never pooled;
iodine and selenium are sparse, as noted throughout the analyses):

```{r}
#| label: coverage
d |>
  count(element, fraction) |>
  pivot_wider(names_from = fraction, values_from = n, values_fill = 0) |>
  rename(Element = element) |>
  kable(format.args = list(big.mark = ","))
```

## Preview

The first rows of the dataset:

```{r}
#| label: preview
d |> head(8) |> kable()
```

## Notes for reuse

- Concentrations are **standardised dry-weight mg/kg** (`value_mgkg`); grain-size
  `fines_pct` is a percentage.
- Empty cells are genuine missing values (for example, `al_mgkg` is empty where aluminium
  was not measured, and `dist_to_aquaculture_km` is empty outside Norway).
- `outlier_flag` marks distributional outliers (mostly registration errors); the analyses
  on this site exclude flagged rows, and you probably want to as well.
- A site is identified by its `latitude` / `longitude` (rounded to 3 decimal places); the
  same location can appear across several years.
