---
title: "Enrichment Summary"
subtitle: "How enriched the collection is, how much of it can be judged, and whether the verdict holds"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Verdicts exist for **cobalt, copper and zinc in bulk** and for nothing else. Selenium and
molybdenum are withheld because most of their measurements fell below the limit of
quantification; manganese and every sieved fraction are withheld because aluminium does not
predict them, so dividing by it is not a grain-size control
([Regression Normalisation](./background-regression.qmd)). Concentrations and the
non-normalised background verdicts are published for everything, as before.

The two Enrichment pages produce verdicts that use **no aquaculture information at all**:
the [enrichment factor](./background-ef.qmd) is metal per aluminium against an offshore
reference, and the [pristine classification](./background-pristine.qmd) combines it with
distributional and spatial screens. The aquaculture gradient is deliberately held out of
all of it, which is what allows the
[Pressure-Based Background](./background-pressure.qmd) to serve as an independent yardstick:
if these verdicts track a pressure they never saw, they are measuring something.

This page recomputes nothing. Every number is read from the files the analysis pages write.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
ps   <- rd("refined_pristine_summary.csv")
pm   <- rd("refined_pristine_meta.csv")
ed   <- rd("refined_ef_dist.csv")
em   <- rd("refined_ef_meta.csv")
efp  <- rd("refined_ef_pressure.csv")
prs  <- rd("refined_pressure_compare.csv")
cov  <- rd("refined_pristine_coverage.csv")
val  <- rd("refined_pristine_validation.csv")
vsrc <- rd("refined_pristine_validation_source.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
cat_levels  <- c("bulk", "sieved63", "sieved20")
bin_levels  <- c("<1km", "1-5km", "5-20km", ">20km")
BLUE <- "#4477AA"; RED <- "#EE6677"
theme_set(theme_minimal(base_size = 12))

# the held-out check: does the median EF rise toward the cages? Compare the nearest band
# with enough measurements against the far band, on the same element and fraction.
grad <- efp |>
  mutate(band = factor(aq_bin, levels = bin_levels)) |>
  filter(n >= pm$min_n) |>
  group_by(symbol, cat) |>
  summarise(near_band = as.character(band[which.min(as.integer(band))]),
            near_ef   = ef_p50[which.min(as.integer(band))],
            far_ef    = ef_p50[band == ">20km"][1], .groups = "drop") |>
  filter(near_band != ">20km", !is.na(far_ef), far_ef > 0) |>
  mutate(ef_ratio = round(near_ef / far_ef, 2),
         evaluation = if_else(ef_ratio >= 1.5, "confirmed", "no signal"))

# the cleaned near/far ratio: the raw >20 km band is over half stated-pressure
# monitoring, so it is not a background. See the Pressure-Based Background page.
near <- prs |> select(symbol, cat, enrich_near = enrich_near_clean,
                      n_near = `n_<1km`)

tbl <- ps |>
  filter(reliable | ((withheld | unnormalisable) & n >= pm$min_n)) |>
  left_join(ed |> select(symbol, cat, ef_p50, ef_p90, pct_lt1_p90ref, ef_p50_p90ref),
            by = c("symbol", "cat")) |>
  left_join(near, by = c("symbol", "cat")) |>
  left_join(grad |> select(symbol, cat, near_band, ef_ratio, evaluation),
            by = c("symbol", "cat")) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         cat = factor(cat, levels = cat_levels)) |>
  arrange(symbol, cat)

bulk <- tbl |> filter(cat == "bulk")
covq <- cov |> filter(axis == "distance to aquaculture")
blank <- function(x) ifelse(is.na(x), "", as.character(x))
cq <- function(b) covq$pct_classifiable[covq$band == b]
```

## The verdict

EF < 1 is read as `r em$ef_lt1`, and there are two defensible references for it. The
headline reference is the **offshore median** of metal / Al, the EFSA-style local reference
the pages cite. The second is the **offshore P90** of the same ratio, which makes EF < 1
mean "at or below the offshore 90th percentile", exactly what the background pages call
background. Half the reference population fails the first by construction. Both are
reported here, and **the spread between them is the answer**, not either column alone.

```{r}
#| label: verdict-table
tbl |>
  transmute(Element = symbol, Fraction = cat, samples = n,
            `% classifiable` = pct_classifiable,
            `% at background (median ref)` = blank(pct_ef),
            `% at background (P90 ref)` = blank(pct_lt1_p90ref),
            `% pristine (strict)` = blank(pct_strict),
            verdict = case_when(withheld ~ "withheld (below LOQ)",
                                unnormalisable ~ "withheld (Al does not predict it)",
                                TRUE ~ "")) |>
  kable(format.args = list(big.mark = ","), align = "llrrrrrl")
```

The spread is wide and it is systematic. In bulk, the share of classifiable samples sitting
at or below background is
`r min(bulk$pct_ef, na.rm = TRUE)` to `r max(bulk$pct_ef, na.rm = TRUE)`% against the median
reference and `r min(bulk$pct_lt1_p90ref, na.rm = TRUE)` to
`r max(bulk$pct_lt1_p90ref, na.rm = TRUE)`% against the P90 reference. The reviewer's
objection that produced this column was right: **"about half is adequate" was largely
arithmetic**, because the median of a distribution has half its own population below it. The
honest statement is that between roughly two fifths and nine tenths of the classifiable
sediment is at or below the local offshore background, depending on which point of the
offshore distribution you call background.

```{r}
#| label: ef-table
bulk |>
  filter(!withheld, !unnormalisable) |>
  transmute(Element = symbol,
            `median EF (median ref)` = ef_p50,
            `median EF (P90 ref)` = ef_p50_p90ref,
            `P90 EF` = ef_p90,
            `near/far P90 ratio (cleaned bg)` = enrich_near,
            `nearest band` = near_band,
            `EF near/far` = ef_ratio,
            evaluation = evaluation) |>
  kable(align = "lrrrrlrl")
```

The median EF against the median reference sits within a few percent of 1 for every element,
which is what a well-behaved local reference looks like. The last three columns are the
held-out check: the near-band median EF over the far-band one, with `confirmed` where the
ratio reaches 1.5. Read `confirmed` as "confirmed against the open shelf", because the far
band is the national pool beyond 20 km of a farm rather than the same fjord without a farm.
Copper (`r bulk$ef_ratio[bulk$symbol == "CU"]` times) and zinc
(`r bulk$ef_ratio[bulk$symbol == "ZN"]` times) rise toward the cages after grain-size control;
cobalt does not. Three rows is the whole table, because three groups is the whole set that
carries a verdict.

On raw concentrations the same gradient does **not** survive a geographic control: matched
within a Norwegian municipality, near-cage sediment carries 1.04 times the copper and 0.85
times the zinc of sediment 5 km away, against 3.2 and 1.8 times when the comparison is against
the open shelf. The aluminium-normalised gradient in the table above is a better-controlled
comparison and it still rises toward the cages, but it rests on the roughly 1% of near-cage
samples that carry aluminium, far too few to match within a municipality. The honest position
is that the strongest evidence for a near-cage effect and the least representative subset of
the data are currently the same rows. [Pressure Controls](./pressure-controls.qmd) has the
full accounting.

**Molybdenum and selenium used to be reported here as the strongest confirmations, and that
was wrong three times over.** They were quoted at about 13x and about 5x, the largest ratios
anywhere on this site. First, the enrichment was said to **survive grain-size
normalisation**. It did, but only because the normalisation did nothing: aluminium is
uncorrelated with molybdenum (Spearman's rho 0.00) and nearly so with selenium (0.15), and a
signal cannot survive a test that was never applied. Second,
[Pressure Controls](./pressure-controls.qmd) found that only 19% of the near-cage molybdenum
measurements were taken while the farm was operating. Third, and fatally, the near band was
measured to the nearest aquaculture site of any kind rather than to a fish farm. On the
corrected axis there are
`r near$n_near[near$symbol == "MO" & near$cat == "bulk"]` bulk molybdenum and
`r near$n_near[near$symbol == "SE" & near$cat == "bulk"]` bulk selenium measurements within
1 km of a sea cage, against a threshold of 30. **The two figures are withdrawn rather than
restated**, and the [Pressure-Based Background](./background-pressure.qmd) page now reports
both elements as having too little near-cage data to read.

## How much can be judged at all

```{r}
#| label: coverage-fig
#| fig-cap: "Percentage of samples that carry aluminium, and can therefore receive any enrichment verdict, by distance to the nearest farm. This is the single most decision-relevant number in the Enrichment section."
#| fig-width: 7.5
#| fig-height: 2.8
covq |>
  mutate(band = factor(band, levels = bin_levels)) |>
  ggplot(aes(band, pct_classifiable)) +
  geom_col(fill = "#AA3377", width = 0.65) +
  geom_text(aes(label = paste0(pct_classifiable, "%")), vjust = -0.4, size = 3.2) +
  scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.1))) +
  labs(x = "distance to nearest aquaculture farm", y = "% classifiable")
```

**`r cq("<1km")`% of the samples within 1 km of a farm can be judged, against
`r cq(">20km")`% beyond 20 km.** The sediment most in need of a verdict is the sediment
least able to receive one, and no amount of method improvement closes that gap, because the
measurement it needs was never made.

The cause is not near-cage sampling practice. Splitting the same measurements by data source
shows aluminium coverage that does not depend on distance at all: within Vannmiljø it is 1
to 4% at **every** band including beyond 20 km, and within Mareano it is essentially
complete at every band including under 1 km. What changes across the bands is which
programme sampled there. So the finding is sharper than "near-cage sediment cannot be
assessed": **Vannmiljø does not report aluminium, and Vannmiljø is the programme that samples
around the farms.** The recommendation follows directly, and it is cheap: report aluminium
(or another grain-size normaliser) alongside the trace elements in Norwegian coastal
monitoring, and most of this gap closes on its own.

### What the classification can and cannot say

**Open-shelf data: trustworthy, for three groups.** Beyond 20 km of a farm, `r cq(">20km")`%
of the eligible samples are classifiable, the verdicts are grain-size-controlled by a
normaliser that demonstrably works, and they are validated against a pressure gradient they
never saw. The validation is also sharper than it was before the restriction: 2, 9, 22 and 49%
pristine across the four distance bands, against 15, 15, 28 and 49% when groups with a
noise-division EF were included. The validation also holds **within a single source**, so it
is not a difference between monitoring programmes wearing a distance label:

```{r}
#| label: validation-table
vsrc |>
  mutate(Band = factor(band, levels = bin_levels)) |>
  transmute(Source = source, Band, samples = n,
            `% pristine (EF rule)` = pct_ef, `% pristine (strict)` = pct_strict) |>
  arrange(Source, Band) |>
  kable(format.args = list(big.mark = ","))
```

**Near-cage data: largely unassessable.** With `r cq("<1km")`% coverage inside 1 km, any
statement about the sediment under Norwegian fish farms rests on a handful of samples from
the rare programmes that measured aluminium there. The pristine shares that *are* computed
near the cages (15% inside 1 km against 49% beyond 20 km) point the right way and are
consistent within each source, but they are a thin base and should be read as a direction,
not a rate.

## Which flag to use

The two flags differ far less than they once did. The **EF rule** is `r pm$rule_ef`. The
**strict rule** is `r pm$rule_strict`, except that the mixture criterion is now dropped for
any group whose distribution does not support two populations rather than applied anyway.
That change removed almost all of the gap between them: in the table above the two shares are
identical for most groups and differ by at most a few points.

- Use the **EF rule** for EFSA-style reporting. It is a single grain-size-controlled
  criterion, it is the one the dataset download carries as `pristine_ef`, and it is the only
  one of the two whose logic is uniform across elements.
- Use the **strict rule** where a conservative subset is wanted, and read it for what it is:
  EF plus two raw-concentration screens, not three peer criteria. Both extra criteria are
  statistics of the concentration distribution, so a muddy but clean sample can fail them for
  reasons of texture that the EF criterion exists to control for. Where the two rules agree,
  which is nearly everywhere, that distinction does not bite.
- Do not treat the two extra criteria as independent confirmation. The mixture threshold and
  the offshore P90 are computed from largely the same measurements, so partial agreement
  between them is built in. They are corroboration, which is worth having, and not a second
  opinion.

::: {.callout-note}
**Aluminium does not normalise every element equally well.** Regressed on the offshore
reference, aluminium explains about half the variation of cobalt, copper and zinc in bulk, and
under a tenth of manganese in bulk or of anything in a sieved fraction
([Regression Normalisation](./background-regression.qmd)). Where it explains nothing, metal /
Al is close to metal divided by an unrelated number, so no verdict is issued there, on the same
principle that already withholds one where the aluminium basis is wrong. That is why the table
above has three rows with numbers in it.

**EF is relative, and relative to a choice.** The reference is the local offshore
distribution of metal / Al, not a crustal abundance, which is deliberate (EFSA guidance
prefers a data-driven local reference) and means these EF values and the 1-2, 2-5, >5 classes
are **not comparable with literature EF values** referenced to Turekian or Wedepohl. The two
reference columns above are the width of the answer within this dataset alone.

**The classifiable subset is not a random sample.** Aluminium is reported for a minority of
bulk samples, and which minority is a property of the reporting programme rather than of the
seabed, as the coverage section sets out. Every percentage on this page is conditional on
that subset.

**Bulk EF also spans two aluminium chemistries.** Sources report either near-total or
acid-extractable aluminium, differing by two to three times, so the reference is restricted
to one basis per fraction and samples on the other basis are left unclassified rather than
compared against the wrong denominator. This is why the classifiable shares for copper and
zinc in bulk are as low as `r min(bulk$pct_classifiable[bulk$symbol %in% c("CU","ZN")])`%.
[Method Revisions](./method-revisions.qmd) has the before and after.

**The yardstick is itself confounded, and now measurably so.** Distance to the nearest farm
is not the pressure a sediment experienced: for copper and zinc about one bulk measurement in
six within 1 km of a farm was taken outside that farm's operating window, and for molybdenum
four in five were. Nor is the far band the same
place with the farms removed: it sits 83 km offshore against 400 m for the near band.
Correcting both takes copper from 3.2 to about 1.1 and zinc from 1.8 to about 0.9. What
remains unseparated is a fjord-wide effect, because sediment 5 km from a farm in a Norwegian
farming municipality is not unfarmed sediment. See
[Pressure Controls](./pressure-controls.qmd). Those four figures are on the **old
aquaculture axis**, which that page still uses; the ratios in the table above are on the
corrected fish-farm axis and against the cleaned background, so the two sets are not
directly comparable until the controls are recomputed.

**Two further limits carried from the analysis pages.** The
[Enrichment Map](./enrichment-map.qmd) applies these thresholds to per-site averages, which
makes it a different classifier from the per-measurement one tabulated here. And ratio
normalisation assumes metal and aluminium are proportional through the origin; a
regression-based normalisation is the stronger standard and is not yet implemented here.

Sources: [Enrichment Factor](./background-ef.qmd),
[Pristine Classification](./background-pristine.qmd),
[Enrichment Map](./enrichment-map.qmd), and
[Pressure-Based Background](./background-pressure.qmd) as the held-out evaluation. The
concentration side is summarised on the [Background Summary](./background-summary.qmd).
:::
