---
title: "Method Revisions"
subtitle: "What changed in the background and pristine methods, and what it did to the numbers"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The methods behind the background and pristine pages have been revised. Numbers taken from
this site before the revision **will not reproduce**, and the dataset download changed
shape, so this page is the reconciliation: what each method was, what it is now, what
forced the change, and which figures moved.

Everything below is generated from the pipeline. The "before" column is the background
module's output exactly as published in **v0.8.0**, kept frozen for the purpose; the
"after" column is the current output. Neither side is typed into this page.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
ch    <- rd("refined_method_changes.csv")
basis <- rd("refined_ef_basis.csv")
cens  <- rd("refined_censoring.csv")
vsrc  <- rd("refined_pristine_validation_source.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
```

## 1. Aluminium is measured on two different bases

**Before.** The enrichment factor divided each sample's metal / Al by one reference: the
offshore median of metal / Al, taken over every data source pooled together.

**Now.** Each fraction's reference is computed within a **single aluminium measurement
basis**, and samples on the other basis are left unclassified rather than judged against a
reference they do not belong to.

**What forced it.** The sources do not all measure aluminium the same way. Some report
near-total aluminium, some report what an acid extraction leaches, which is two to three
times less. In the same sea area, Mareano's bulk aluminium averages about 15,600 mg/kg
against MUDAB's 56,000, while their iron differs by less than a factor of two: it is the
normaliser that differs, not the sediment. Since the bulk reference was 54 to 99% Mareano,
every non-Mareano sample was divided by a reference far too high, and came out too
pristine. The same seabed scored 97% adequate for cobalt if ICES-DOME measured it and 41%
if Mareano did.

No source records its digestion, so the basis is inferred per sample from **Fe/Al**. Both
elements are lithogenic and track grain size, so their ratio is nearly grain-size free,
and an acid extraction depresses aluminium far more than iron. Crustal Fe/Al is about 0.5;
the cut is at 1.0. It splits ICES-DOME, a single source, into two groups whose metal / Al
differs by 1.7 to 2.5 times, which is what says it is finding a protocol rather than
geology.

```{r}
#| label: basis-table
basis |>
  mutate(Fraction = factor(cat, levels = c("bulk", "sieved63", "sieved20"))) |>
  transmute(Fraction, Basis = al_basis,
            `Used for EF` = ifelse(adopted, "yes", "no"),
            n, `% of fraction` = pct_of_frac,
            `offshore rows` = n_offshore, `rows < 1 km from a farm` = n_aq_lt1km) |>
  arrange(Fraction, desc(`Used for EF`), desc(n)) |>
  kable(format.args = list(big.mark = ","))
```

Bulk had to take the extraction basis: the total-basis stratum has **no** near-cage samples
at all. The consequence is that bulk aluminium is acid-leachable, so **bulk EF is
internally comparable but not comparable with literature EF values**. `unplaced` rows are
those with no iron reported, which cannot be placed either way.

The restriction costs roughly half the classifiable rows and buys a reference that means
one thing. After it, Mareano's own reference sits within 2% of the adopted one, where
before it ran up to 21% away.

## 2. Selenium and molybdenum carry no verdict

**Before.** Both elements had backgrounds, thresholds, enrichment factors and pristine
verdicts, presented like the other five.

**Now.** Their concentrations are still published. Every **verdict** is withheld.

**What forced it.** Measurements below the limit of quantification are removed during
cleaning. That is the right rule for contamination screening, where a non-detect carries no
usable number, and it is the wrong rule for estimating a background, because a background
**is** the low end of the distribution and non-detects are evidence about exactly that end.

For five of the seven target elements this removes under 2% of measurements and does not
matter. For selenium and molybdenum it removes the majority:

```{r}
#| label: censoring-table
cens |>
  filter(source == "ALL") |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Element = symbol, `measurements` = n_slim,
            `below LOQ` = n_censored, `% below LOQ` = pct_censored,
            `verdicts` = ifelse(withheld, "withheld", "published")) |>
  arrange(desc(`% below LOQ`)) |>
  kable(format.args = list(big.mark = ","))
```

For Mareano, which supplies 94 to 99% of the bulk molybdenum and selenium reference, 86% of
molybdenum and 70% of selenium were removed. What survived is the detected upper tail, so
the figure published as a "90th-percentile background" sat at roughly the **98th
percentile** of the real distribution. Rather than print that with a warning, no verdict is
issued, which is the same rule applied to the aluminium basis above.

This does **not** affect the near-cage molybdenum and selenium enrichment reported on the
[Enrichment Factor](./background-ef.qmd) page. Removing low offshore values raises the
offshore reference, which suppresses near-cage EF, so that signal is conservative and
stands.

Doing it properly means keeping below-LOQ rows as censored observations and using methods
that handle them without inventing a value. That is a separate piece of work across the
whole pipeline.

## 3. The mixture's number of components is chosen, not assumed

**Before.** The distribution mixture was always fitted with two components, so it always
produced a threshold, whether or not the data held two populations.

**Now.** One, two and three components are each fitted and compared by BIC. Where the
result does not support a separated pair, the threshold is marked **not usable**: it is
still reported, because seeing how far off it was is informative, but it is dropped from
the strict pristine rule instead of being applied as a number that marks nothing.

**What forced it.** Cobalt is the clear case. Its threshold, 6.45 mg/kg, sat almost exactly
on the background geometric mean of 6.22, with only 31% of samples assigned to
"background". A background component that is a minority of the data inverts the method's own
premise. One component never wins on BIC, so these distributions are genuinely not single
normals, but three components win for copper, manganese, selenium and zinc, and combined
with the separation test this marks cobalt, copper and zinc thresholds unusable.

The thresholds themselves did not move: the two-component fit kept its own search, so this
changed which thresholds are **used**, not what they are.

## 4. A second enrichment reference

**Before.** One reference, the offshore median of metal / Al, with EF < 1 read as adequate.

**Now.** The median stays as the headline and the offshore **90th percentile** is reported
beside it, as `ef_p90ref` in the download.

**What forced it.** EF < 1 against a median denominator fails half the reference population
by construction. That is where "about half of the samples are adequate" came from: it was
arithmetic before it was ecology. Against the P90, EF < 1 means "below the offshore 90th
percentile", which is how the [Global Background](./background.qmd) page and the
`background_p90` column already define background. The pair also exposes an inconsistency
the site was carrying, with two different definitions of background on different pages.

## 5. Repeat sampling is measured, not assumed away

**Before.** The background percentiles were per measurement, with no statement of what that
weighting implied.

**Now.** They are still per measurement, and the [Global Background](./background.qmd) page
carries a site-weighted sensitivity beside them, plus the test of whether the weighting
actually biases anything.

**What forced it.** A site visited in ten years counted ten times. That matters only if
revisited sites differ systematically from the rest, which was plausible and untested.
Tested, it holds: a site sampled in four or more years sits a median 0.75 km from the
nearest farm against 1.96 km for a site sampled once. Reweighting one-value-per-site moves
the offshore bulk P90 down by 4% for cobalt and zinc, 15% for copper and **32% for
manganese**, always downward.

The headline stays per measurement, because site-weighting discards real variation across
years and depths and "site" is only a rounded coordinate. But for manganese the two differ
enough that the range is the answer rather than either figure.

## The numbers that moved

Bulk, the verdict measures. `withdrawn` marks a figure no longer published.

```{r}
#| label: changes-table
ch |>
  filter(cat == "bulk", grepl("pristine|classifiable|adequate", measure)) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  transmute(Measure = measure, Element = symbol, Before = before, After = after,
            Change = change) |>
  arrange(Measure, Element) |>
  kable(format.args = list(big.mark = ","))
```

Two things are worth reading carefully. The **classifiable** share falls everywhere,
steeply for copper and zinc, because the aluminium-basis restriction withdraws samples
rather than misjudging them. And cobalt's **strict** share rises sharply, from 20 to 47%,
because its mixture threshold is now recognised as unusable and dropped from the rule; the
old 20% was cobalt failing a threshold that marked nothing.

The full comparison, including the references and thresholds themselves, is in
`refined_method_changes.csv` in the [dataset download](./download-dataset.qmd).

## What the revision did not change

The two findings the site rests on both survive, and one is now better supported.

**Near-cage enrichment survives grain-size control.** Copper, zinc, selenium and molybdenum
are still clearly enriched approaching the cages after normalising to aluminium.

**The distance gradient is not an artefact of which programme sampled where.** This was an
open question: the near bands are largely Vannmiljo and the far bands largely Mareano, so a
gradient pooled across sources could have been a difference between programmes. Restricting
to one aluminium basis made the bands homogeneous enough to test it within a single source,
and it holds in both:

```{r}
#| label: within-source
vsrc |>
  mutate(Band = factor(band, levels = c("<1km", "1-5km", "5-20km", ">20km"))) |>
  transmute(Source = source, Band, `samples` = n,
            `% pristine (EF rule)` = pct_ef, `% pristine (strict)` = pct_strict) |>
  arrange(Source, Band) |>
  kable(format.args = list(big.mark = ","))
```

::: {.callout-note}
This page covers the background and pristine methods only. The upstream stages (parsing,
harmonisation, quality control, cross-source de-duplication) are unchanged, so the
concentrations themselves have not moved: `value_mgkg` for a given sample is the same
before and after. What changed is which samples receive a verdict, and what the verdict is
measured against.
:::
