---
title: "Pressure Controls"
subtitle: "What survives when the near-cage comparison is matched and time-aligned"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The [Pressure-Based Background](./background-pressure.qmd) page reads a near/far P90 ratio as
near-cage enrichment: copper 2.62 times, zinc 1.09 times. That reading assumes the two bands
differ only in pressure. They differ in two other ways that this database can measure, so this
page measures them.

- **Time.** Every sample is keyed to its *nearest* farm, whether or not that farm existed
  when the sediment was collected. The aquaculture reference carries the clearance year and
  the withdrawal year, so each near-farm measurement can be placed before the site was
  cleared, during its operation, or after it closed.
- **Geography.** The far band is the national pool more than 20 km from any farm, which is
  open shelf. The near band is fjord. Comparing them compares two settings as much as two
  pressures. The fix is to compare **within a municipality**: near-farm sediment against
  sediment 5 km or more away in the same municipality.

Nothing here feeds a verdict. This page qualifies the pressure gradient the
[Enrichment](./background-ef.qmd) pages use as their held-out yardstick, and the result is
that the yardstick is softer than it looked.


```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

bdir <- "data/refined_summary"
rd <- function(f) read.csv(file.path(bdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
tmp  <- rd("refined_temporal_alignment.csv")
ctl  <- rd("refined_pressure_controls.csv")
mun  <- rd("refined_matched_municipality.csv")
cov  <- rd("refined_control_covariates.csv")
meta <- rd("refined_pressure_controls_meta.csv")
prs  <- rd("refined_pressure_compare.csv")

elem_levels  <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
per_levels   <- c("pre-farm", "operating", "post-closure")
ctl_levels   <- c("published", "operating only", "municipality-matched",
                  "matched + operating")
BLUE <- "#4477AA"; RED <- "#EE6677"
theme_set(theme_minimal(base_size = 12))

near <- tmp |> filter(cat == "bulk", band == "<1km")
share <- near |>
  group_by(symbol) |>
  mutate(total = sum(n), pct = round(100 * n / total)) |>
  ungroup()
op_pct <- function(sym) share$pct[share$symbol == sym & share$period == "operating"]
blank <- function(x) ifelse(is.na(x), "", as.character(x))
```

## Was the sediment there while the farm was?

Bulk measurements within 1 km of a farm, placed in that farm's lifetime. The rule is
`r meta$period_rule`.

```{r}
#| label: period-fig
#| fig-cap: "Where the near-cage bulk measurements fall in the lifetime of the fish farm they are keyed to. Only the middle band is evidence about farm pressure. Copper, zinc and cobalt are around 87% contemporary with an operating farm; molybdenum has 18 measurements in total and is not."
#| fig-width: 8
#| fig-height: 3
share |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         period = factor(period, levels = per_levels)) |>
  ggplot(aes(pct, symbol, fill = period)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = ifelse(pct >= 8, paste0(pct, "%"), "")),
            position = position_stack(vjust = 0.5), size = 3, colour = "white") +
  scale_fill_manual(values = c(`pre-farm` = "#AA3377", operating = BLUE,
                               `post-closure` = "#CCBB44"), name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(x = "% of measurements within 1 km of a farm", y = NULL)
```

```{r}
#| label: period-table
near |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         period = factor(period, levels = per_levels)) |>
  arrange(symbol, period) |>
  transmute(Element = symbol, Period = period, n,
            `median year` = sprintf("%.0f", median_year), `P90 (mg/kg)` = p90,
            reliable = if_else(reliable, "", "n < 30")) |>
  kable(format.args = list(big.mark = ","))
```

**The alignment is good, and it was not before.** `r op_pct("CU")`% of near-cage copper
measurements and `r op_pct("ZN")`% of near-cage zinc were taken while the farm was operating,
and cobalt is at `r op_pct("CO")`%. **Molybdenum** has
`r sum(near$n[near$symbol == "MO"])` bulk measurements within 1 km of a sea cage that can be
placed in a farm's lifetime at all, of which
`r near$n[near$symbol == "MO" & near$period == "operating"]` are contemporary with an
operating one. It cannot be read.

::: {.callout-important
title="The pre-farm result reversed when the farm link was corrected"}
This page used to report that near-cage sediment sampled **before** the farm existed was
*more* contaminated than sediment sampled while it operated, and read that as evidence that
distance to a farm is partly a marker of industrialised coastline rather than of farm input.

That was an artefact. Until August 2026 the licence dates came from `aqua_id`, the nearest
aquaculture site of any kind, so a sample could be dated against a mussel raft or a
land-based smolt plant while its chemistry was attributed to a cage that was not there. The
refined `site` table now carries `fish_farm_aqua_id`, the nearest **fish farm**, and the
period each sample sits in is tested against that farm's own licence.

On the corrected link the ordering is the one the pressure hypothesis predicts. For copper the
pre-farm P90 is `r near$p90[near$symbol == "CU" & near$period == "pre-farm"]` mg/kg against
`r near$p90[near$symbol == "CU" & near$period == "operating"]` while operating and
`r near$p90[near$symbol == "CU" & near$period == "post-closure"]` after closure; for zinc
`r near$p90[near$symbol == "ZN" & near$period == "pre-farm"]`,
`r near$p90[near$symbol == "ZN" & near$period == "operating"]` and
`r near$p90[near$symbol == "ZN" & near$period == "post-closure"]`. Sediment sampled before
the cages arrived is the cleanest of the three periods, and sediment sampled after they left
is the dirtiest, which is what a persistent deposit looks like.

The pre-farm samples are not simply older: their median year is
`r near$median_year[near$symbol == "CU" & near$period == "pre-farm"]` against
`r near$median_year[near$symbol == "CU" & near$period == "operating"]` for the operating ones,
overlapping, because clearance years run to 2026 and a 2019 sample can sit before a farm
cleared in 2022.
:::

## Is the comparison like for like?

The far band is not a version of the near band with the farms taken out. It is different
sediment in a different place:

```{r}
#| label: covariate-table
cov |>
  transmute(Subset = subset, samples = n_samples,
            `median year` = sprintf("%.0f", median_year),
            `distance to coast, km` = sprintf("%.3g", median_dist_to_coast_km),
            `mud, %` = median_fines_pct,
            `% with a mud measurement` = pct_with_fines) |>
  kable(format.args = list(big.mark = ","))
```

The national background band sits a median 72 km offshore and is 77% mud. The near-cage band
sits about 400 m from shore and is 34% mud. Any comparison between them is a comparison of the
open shelf against the coast, and grain size runs the opposite way to the finding, since the
muddier band is the one with the *lower* metal.

Matching within a municipality removes the shelf-versus-coast contrast. The local far band
(5 to 20 km from a farm, same municipality) sits about as close to shore as the near band and
is twice as muddy (70% against 34%), so if anything the matched comparison now understates a
near-cage effect rather than inventing one.

Both far bands here are the **cleaned** ones, with Vannmiljø stated-pressure monitoring
removed, for the reason the [Pressure-Based Background](./background-pressure.qmd) page sets
out: a band that is half contaminated-seabed monitoring is not a background.

## What survives

```{r}
#| label: controls-fig
#| fig-cap: "The near/far P90 ratio for each element under successive controls. The dashed line is no difference. Copper keeps a clear excess through every control; zinc has almost nothing to lose and ends at no difference."
#| fig-width: 8
#| fig-height: 3.6
ctl |>
  filter(cat == "bulk", reliable) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         control = factor(control, levels = ctl_levels)) |>
  ggplot(aes(ratio_p90, symbol, colour = control)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey60") +
  geom_line(aes(group = symbol), colour = "grey85", linewidth = 0.7) +
  geom_point(size = 3, alpha = 0.9) +
  scale_x_log10() +
  scale_colour_manual(values = c(published = "#AA3377", `operating only` = "#CCBB44",
                                 `municipality-matched` = BLUE,
                                 `matched + operating` = "#228833"), name = NULL) +
  labs(x = "near / far P90 ratio (log scale)", y = NULL)
```

```{r}
#| label: controls-table
ctl |>
  filter(cat == "bulk") |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         control = factor(control, levels = ctl_levels)) |>
  arrange(symbol, control) |>
  transmute(Element = symbol, Control = control, `n near` = n_near, `n far` = n_far,
            `near/far P90` = ratio_p90,
            municipalities = blank(n_mun),
            `% of them > 1` = blank(pct_mun_gt1),
            reliable = if_else(reliable, "", "n < 30")) |>
  kable(format.args = list(big.mark = ","))
```

**Copper survives every control.** It starts at
`r ctl$ratio_p90[ctl$symbol == "CU" & ctl$control == "published"]`, is unmoved by restricting
to time-aligned samples (`r ctl$ratio_p90[ctl$symbol == "CU" & ctl$control == "operating only"]`),
and settles at `r ctl$ratio_p90[ctl$symbol == "CU" & ctl$control == "municipality-matched"]`
matched within municipality and
`r ctl$ratio_p90[ctl$symbol == "CU" & ctl$control == "matched + operating"]` matched and
time-aligned together. The per-municipality detail says the same without relying on a pooled
statistic: near sediment exceeds local far sediment in
`r ctl$pct_mun_gt1[ctl$symbol == "CU" & ctl$control == "municipality-matched"]`% of the
`r ctl$n_mun[ctl$symbol == "CU" & ctl$control == "municipality-matched"]` municipalities that
can be tested. That is not a coin flip, and it is the strongest result on this site for a
near-cage effect.

**Zinc has almost nothing to lose and loses it.** From
`r ctl$ratio_p90[ctl$symbol == "ZN" & ctl$control == "published"]` published to
`r ctl$ratio_p90[ctl$symbol == "ZN" & ctl$control == "matched + operating"]` fully controlled,
with near sediment exceeding local far in
`r ctl$pct_mun_gt1[ctl$symbol == "ZN" & ctl$control == "municipality-matched"]`% of the
`r ctl$n_mun[ctl$symbol == "ZN" & ctl$control == "municipality-matched"]` municipalities. *That*
is a coin flip. The median-based ratio is below 1
(`r ctl$ratio_p50[ctl$symbol == "ZN" & ctl$control == "municipality-matched"]`), so the small
excess that remains in the P90 sits in the tail rather than in the typical sample.

**Nothing else can be read.** Cobalt, manganese, molybdenum and selenium all fall below 30
time-aligned near-cage measurements, and none can be matched within a municipality: no
municipality has ten near-cage and ten local-far measurements of any of them. This is a
sharper limit than it was on the old axis, and it is the correct one: those counts were being
padded by samples that sat next to land-based facilities.

## What this does and does not mean

**It does not mean the farms have no effect.** The matched design asks a narrower question
than the pressure page: not "is sediment near a farm different from open-shelf sediment", but
"is sediment near a farm different from sediment a few kilometres away in the same fjord
system". In Norwegian farming municipalities the local far band is itself a farmed landscape,
often with another farm just beyond 5 km, so a null result is consistent with a fjord-wide
effect as well as with no effect. Matching cannot separate those two, and no distance-based
design on this data can.

**It does mean the published ratio is an upper bound.** Between the national
`r ctl$ratio_p90[ctl$symbol == "CU" & ctl$control == "published"]` times for copper and the
matched `r ctl$ratio_p90[ctl$symbol == "CU" & ctl$control == "municipality-matched"]` times,
the honest reading is that the answer lies in that range and the data cannot narrow it
further. For zinc the range is
`r ctl$ratio_p90[ctl$symbol == "ZN" & ctl$control == "published"]` to
`r ctl$ratio_p90[ctl$symbol == "ZN" & ctl$control == "matched + operating"]`, which is to say
no effect worth quoting.

**It supports, rather than weakens, the enrichment finding for copper.** The
[Enrichment Factor](./background-ef.qmd) page shows the same gradient after aluminium
normalisation, which is a different control and an important one, and that gradient is not
tested here because aluminium is reported for about 1% of near-cage samples, far too few to
match within a municipality. So the surviving evidence for a near-cage effect is the
grain-size-controlled one on a small classifiable subset, not the raw distance gradient on the
full one.

::: {.callout-note}
Three limits on this page itself.

**The farm link is to the nearest fish farm, not to the farm that was there.** A sample
classified `pre-farm` may sit near an older farm that has since been withdrawn and re-cleared
under a different registration, or near a second farm slightly further away. The
classification is correct about the specific farm it is keyed to and approximate about the
location's history. It is keyed to a **fish** farm since August 2026; before that it was keyed
to the nearest aquaculture site of any kind, which is what produced the reversed pre-farm
result described above.

**Municipality is a coarse matching unit.** Norwegian coastal municipalities span several
fjord arms with different depths, sills and redox regimes. It controls for region and for
programme, not for basin.

**The matched comparison is on raw concentrations**, so it inherits the grain-size caveat of
the [Pressure-Based Background](./background-pressure.qmd) page. Here the bias runs against the
finding rather than for it, because the local far band is muddier than the near band, but it is
not removed. A matched comparison on aluminium-normalised values is the analysis this page
would like to run and cannot, for want of aluminium.

Read alongside the [Pressure-Based Background](./background-pressure.qmd), which this page
qualifies, and the [Enrichment Summary](./enrichment-summary.qmd), which carries the
consequence into the verdicts.
:::
